---
title: "3. Synthetic Difference-in-Differences"
author: "Mike Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: synthdid.bib
vignette: >
  %\VignetteIndexEntry{3. Synthetic Difference-in-Differences}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
library(causalverse)
knitr::opts_chunk$set(cache = TRUE)
```

# Synthetic DID

## Block Adoption (Assignment)

This section contains the code from the `synthdid` package [@arkhangelsky2021synthetic], which can handle block adoption (i.e., when all units are treated at the same time).

```{r}
library(synthdid)

# Estimate the effect of California Proposition 99 on cigarette consumption
data('california_prop99')

setup = synthdid::panel.matrices(synthdid::california_prop99)

tau.hat = synthdid::synthdid_estimate(setup$Y, setup$N0, setup$T0)

print(summary(tau.hat))

# for only one treated unit, we can only use the placebo method
# se = sqrt(vcov(tau.hat, method='placebo'))
# sqrt(vcov(tau.hat, method='bootstrap'))
# sqrt(vcov(tau.hat, method='jackknife'))

# sprintf('point estimate: %1.2f', tau.hat)
# sprintf('95%% CI (%1.2f, %1.2f)', tau.hat - 1.96 * se, tau.hat + 1.96 * se)
plot(tau.hat) + ama_theme()
```

Using `fixest` dataset

```{r}
library(fixest)
library(tidyverse)
```

If you want to use the `base_stagg` dataset, you have to choose just 1 period, because function `synthdid_estimate` cannot handle staggered adoption.

```{r, eval = FALSE}
df <- base_stagg |>
  
  filter(year_treated %in% c(10000, 8)) |>
  transmute(
    id = as.factor(id),
    year = as.integer(year),
    y = as.double(y),
    treated = as.integer(if_else(time_to_treatment >= 0, 1, 0))
  )
head(df)

setup = df |>
  # specify column names
  panel.matrices(
    unit = "id",
    time = "year",
    outcome = "y",
    treatment = "treated"
  )
```

```{r}
head(fixest::base_did)
df = fixest::base_did |> 
  mutate(
    id = as.factor(id),
    period = as.integer(period),
    y = as.double(y),
    post = as.integer(post)
  ) |> 
  # correct treatment 
  mutate(treatment = as.integer(if_else(treat == 0, 0, post))) 

setup <- df |>
  synthdid::panel.matrices(
    unit = "id",
    time = "period",
    outcome = "y",
    treatment = "treatment"
  )


tau.hat = synthdid::synthdid_estimate(setup$Y, setup$N0, setup$T0)
tau.hat
# sqrt(vcov(tau.hat, method='bootstrap'))
sqrt(vcov(tau.hat, method='jackknife'))

# to use this SE method, must have more controls than treated units
# can try the base_stagg dataset from above 
# sqrt(vcov(tau.hat, method='placebo'))

# Comparing counterfactual estimate to real-world observation
synthdid::synthdid_plot(tau.hat
                        # , se.method = "placebo"
                        ) + ama_theme()

# Control unit contribution plot (black is the ATE, grey lines are CIs)
synthdid::synthdid_units_plot(tau.hat) + ama_theme(base_size = 8)

# pre-treatment parallel trends
# synthdid::synthdid_plot(tau.hat, overlay = 1)
# synthdid::synthdid_plot(tau.hat, overlay = .8)
# synthdid::synthdid_placebo_plot(tau.hat)
# synthdid::synthdid_rmse_plot(tau.hat)
```

To incorporate time-varying elements into the unit and time weights of the `synthdid`, consider using residuals obtained from regressing observed outcomes against time-varying variables. This can be expressed as:

$$
Y_{it}^{res} = Y_{it} - X_{it} \hat{\beta}
$$

where $\hat{\beta}$ is derived from the regression equation $Y = \beta X_{it}$.

```{r}
setup2 <- df |>
  dplyr::mutate(y_res = residuals(lm(y ~ x1))) |>
  synthdid::panel.matrices(
    unit = "id",
    time = "period",
    outcome = "y_res",
    treatment = "treatment"
  )


tau.hat2 = synthdid::synthdid_estimate(setup2$Y, setup2$N0, setup2$T0)
tau.hat2
```

```{r}
tau.sc   = synthdid::sc_estimate(setup$Y, setup$N0, setup$T0)
tau.did  = synthdid::did_estimate(setup$Y, setup$N0, setup$T0)
estimates = list(tau.did, tau.sc, tau.hat)
names(estimates) = c('Diff-in-Diff', 'Synthetic Control', 'Synthetic Diff-in-Diff')

print(unlist(estimates))

synthdid::synthdid_plot(estimates) +
  ama_theme()
```

Spaghetti plots

```{r}
top.controls = synthdid::synthdid_controls(tau.hat)[1:10, , drop=FALSE]
plot(tau.hat, spaghetti.units=rownames(top.controls)) + ama_theme()
```

```{r}
synthdid::synthdid_units_plot(estimates) +
  ama_theme()
```

```{r}
synthdid::synthdid_units_plot(tau.hat, units = rownames(top.controls)) + ama_theme()
```

Compare to other estimators

```{r}
setup = synthdid::panel.matrices(synthdid::california_prop99)

# Run for specific estimators
results_selected = causalverse::panel_estimate(setup, selected_estimators = c("synthdid", "did", "sc"))

results_selected
# to access more details in the estimate object
summary(results_selected$did$estimate)

process_panel_estimate(results_selected)
```

## Staggered Adoptions

This vignette adapts the block assignment method from @arkhangelsky2021synthetic for staggered adoption scenarios. Following the approach suggested by @ben2022synthetic, separate assignment matrices are created for each treatment period to fit the block assignment context. The `synthdid` estimator is then applied to each sub-sample. Finally, the Average Treatment Effect on the Treated (ATT) is calculated as a weighted average of all ATTs for all treated periods, with weights based on the proportion of treated units in each time period for each sub-sample.

```{r}
library(tidyverse)
library(fixest)
df <- fixest::base_stagg |>
   dplyr::mutate(treatvar = if_else(time_to_treatment >= 0, 1, 0)) |>
   dplyr::mutate(treatvar = as.integer(if_else(year_treated > (5 + 2), 0, treatvar))) |> 
  mutate(subgroup = sample(c(0,1), nrow(base_stagg), replace = T))


est <- causalverse::synthdid_est_ate(
  data               = df,
  adoption_cohorts   = 5:7,
  lags               = 2,
  leads              = 2,
  time_var           = "year",
  unit_id_var        = "id",
  treated_period_var = "year_treated",
  treat_stat_var     = "treatvar",
  outcome_var        = "y", 
)

data.frame(
    Period = names(est$TE_mean_w),
    ATE = est$TE_mean_w,
    SE = est$SE_mean_w
) |>
    causalverse::nice_tab()

causalverse::synthdid_plot_ate(est)
```

Compare to different estimators (e.g., DID and SC)

```{r, message=FALSE, eval = FALSE}
methods <- c("synthdid", "did", "sc", "sc_ridge", "difp", "difp_ridge")

estimates <- lapply(methods, function(method) {
  synthdid_est_ate(
    data               = df,
    adoption_cohorts   = 5:7,
    lags               = 2,
    leads              = 2,
    time_var           = "year",
    unit_id_var        = "id",
    treated_period_var = "year_treated",
    treat_stat_var     = "treatvar",
    outcome_var        = "y",
    method = method
  )
})

plots <- lapply(seq_along(estimates), function(i) {
  causalverse::synthdid_plot_ate(estimates[[i]],
                                 title = methods[i],
                                 theme = causalverse::ama_theme(base_size = 6))
})

gridExtra::grid.arrange(grobs = plots, ncol = 2)
```

Subgroup Analysis (i.e., exploring treatment effects heterogeneity)

```{r}
library(tidyverse)

est_sub <- causalverse::synthdid_est_ate(
  data               = df,
  adoption_cohorts   = 5:7,
  lags               = 2,
  leads              = 2,
  time_var           = "year",
  unit_id_var        = "id",
  treated_period_var = "year_treated",
  treat_stat_var     = "treatvar",
  outcome_var        = "y",
  # a vector of subgroup id (from unit id)
  subgroup           =  c(
    # some are treated
    "11", "30", "49" ,
    # some are control within this period
    "20", "25", "21")
)

data.frame(
    Period = names(est_sub$TE_mean_w),
    ATE = est_sub$TE_mean_w,
    SE = est_sub$SE_mean_w
) |>
    causalverse::nice_tab()

causalverse::synthdid_plot_ate(est)
```

To incorporate time-varying variables, use the residuals obtained from regressing the outcome on these variables.

```{r}
df_x <- fixest::base_stagg |>
   dplyr::mutate(treatvar = if_else(time_to_treatment >= 0, 1, 0)) |>
   dplyr::mutate(treatvar = as.integer(if_else(year_treated > (5 + 2), 0, treatvar))) |> 
  
  mutate(y_res = residuals(lm(y ~ x1)))


est_x <- causalverse::synthdid_est_ate(
  data               = df_x,
  adoption_cohorts   = 5:7,
  lags               = 2,
  leads              = 2,
  time_var           = "year",
  unit_id_var        = "id",
  treated_period_var = "year_treated",
  treat_stat_var     = "treatvar",
  outcome_var        = "y_res"
)

data.frame(
    Period = names(est_x$TE_mean_w),
    ATE    = est_x$TE_mean_w,
    SE     = est_x$SE_mean_w
) |>
    causalverse::nice_tab()

causalverse::synthdid_plot_ate(est_x)
```
