---
title: "2. Regression Discontinuity"
author: "Mike Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
bibliography: rd.bib
vignette: >
  %\VignetteIndexEntry{2. Regression Discontinuity}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>",
  cache = TRUE
)
```

```{r setup}
library(causalverse)
library(dplyr)
library(ggplot2)
library(fixest)
```

# 1. Introduction to Regression Discontinuity Designs

Regression Discontinuity (RD) designs are among the most credible quasi-experimental methods for estimating causal effects. The idea is simple: when treatment assignment is determined by whether a continuous **running variable** (also called a forcing variable or score) crosses a known **cutoff**, units just above and just below the cutoff are nearly identical in all observable and unobservable characteristics. Any discrete jump in the outcome at the cutoff can therefore be attributed to the treatment.

## 1.1 Potential Outcomes Framework for RD

Let $X_i$ denote the running variable and $c$ the cutoff. Define a treatment indicator:

$$
D_i = \mathbf{1}(X_i \geq c)
$$

Each unit $i$ has two potential outcomes: $Y_i(1)$ (outcome if treated) and $Y_i(0)$ (outcome if untreated). The observed outcome is:

$$
Y_i = D_i \cdot Y_i(1) + (1 - D_i) \cdot Y_i(0)
$$

The **Sharp RD** treatment effect at the cutoff is defined as:

$$
\tau_{SRD} = \lim_{x \downarrow c} E[Y_i | X_i = x] - \lim_{x \uparrow c} E[Y_i | X_i = x]
$$

Under the continuity assumption (see below), this equals:

$$
\tau_{SRD} = E[Y_i(1) - Y_i(0) | X_i = c]
$$

This is the **average treatment effect at the cutoff**, a local estimand.

## 1.2 Key Identifying Assumptions

The validity of an RD design rests on several assumptions:

1. **Continuity of conditional expectations**: The conditional expectation functions $E[Y_i(0) | X_i = x]$ and $E[Y_i(1) | X_i = x]$ are continuous at $x = c$. This is the core assumption: the potential outcomes vary smoothly through the cutoff, so the only thing that changes discontinuously is treatment status. Formally:

$$
\lim_{x \downarrow c} E[Y_i(d) | X_i = x] = \lim_{x \uparrow c} E[Y_i(d) | X_i = x], \quad d \in \{0, 1\}
$$

2. **No precise manipulation**: Units cannot precisely control their value of the running variable to sort themselves above or below the cutoff. If individuals can manipulate their score to receive (or avoid) treatment, the continuity assumption fails because units just above and below the cutoff are no longer comparable. This assumption is testable via density tests (see Section 5 on `rddensity`).

3. **SUTVA (Stable Unit Treatment Value Assumption)**: The treatment status of unit $i$ does not affect the outcome of unit $j$, and there is a single version of treatment. In the RD context, this means that the cutoff-based rule creates a well-defined treatment with no spillovers across units near the cutoff.

4. **Running variable is not affected by treatment**: The running variable itself should be determined before (or independently of) treatment. The running variable is a pre-treatment characteristic.

## 1.3 Sharp vs. Fuzzy RD

**Sharp RD**: Treatment is a deterministic function of the running variable. Every unit above the cutoff is treated; every unit below is not:

$$
D_i = \mathbf{1}(X_i \geq c)
$$

The compliance is perfect. The treatment effect is identified by the jump in $E[Y_i | X_i = x]$ at $c$.

**Fuzzy RD**: The probability of treatment changes discontinuously at the cutoff but not from 0 to 1. Some units above the cutoff are untreated (never-takers or defiers), and some below are treated (always-takers):

$$
\lim_{x \downarrow c} E[D_i | X_i = x] - \lim_{x \uparrow c} E[D_i | X_i = x] = \kappa, \quad 0 < \kappa < 1
$$

The Fuzzy RD estimand is a ratio, analogous to a Wald/IV estimand:

$$
\tau_{FRD} = \frac{\lim_{x \downarrow c} E[Y_i | X_i = x] - \lim_{x \uparrow c} E[Y_i | X_i = x]}{\lim_{x \downarrow c} E[D_i | X_i = x] - \lim_{x \uparrow c} E[D_i | X_i = x]}
$$

Under a monotonicity (no defiers) assumption, $\tau_{FRD}$ identifies the **local average treatment effect (LATE)** for compliers at the cutoff.

## 1.4 Continuity-Based vs. Local Randomization Approaches

There are two main frameworks for inference in RD designs:

**Continuity-based approach**: This is the standard framework. It relies on the continuity of conditional expectation functions and uses local polynomial regression to estimate the treatment effect. Inference is based on large-sample approximations. The `rdrobust` package implements this approach with optimal bandwidth selection and bias-corrected confidence intervals.

**Local randomization approach**: Near the cutoff, assignment to treatment is "as good as random." Within a narrow window around the cutoff, the running variable is essentially noise, and treatment assignment is effectively randomized. This framework uses randomization inference (permutation tests) rather than large-sample asymptotics. The `rdlocrand` package implements this approach.

The choice between frameworks depends on the application. The continuity-based approach is more general, while the local randomization approach can provide more robust inference when a sufficiently narrow window of "as-if random" assignment exists.

---

# 2. Simulated Data

We create a rich simulated dataset that we will use throughout this vignette. The data-generating process features a running variable, treatment assignment at a cutoff of zero, a nonlinear outcome function, and several pre-determined covariates.

```{r simulate-data}
set.seed(42)
n <- 2000

rd_data <- tibble(
  # Running variable: centered at 0

  x = runif(n, -1, 1),

  # Treatment: sharp assignment at cutoff = 0
  treated = as.integer(x >= 0),

  # Pre-determined covariates (should NOT jump at the cutoff)
  age        = 35 + 4 * x + rnorm(n, sd = 3),
  female     = rbinom(n, 1, prob = plogis(0.1 + 0.2 * x)),
  education  = 12 + 2 * x + 0.5 * x^2 + rnorm(n, sd = 1.5),
  income     = 50000 + 8000 * x - 1500 * x^2 + rnorm(n, sd = 5000),

  # Potential outcomes
  y0 = 2 + 1.5 * x + 0.8 * x^2 - 0.3 * x^3 + 0.5 * age / 35 + rnorm(n, sd = 0.5),
  y1 = y0 + 3 + 0.5 * x,  # Treatment effect = 3 at cutoff, varies with x

 # Observed outcome
  y = ifelse(treated == 1, y1, y0)
) |>
  # Interaction of running variable with treatment (for separate slopes)
  mutate(x_treated = x * treated)

# True treatment effect at the cutoff (x = 0)
# tau = 3 + 0.5 * 0 = 3
cat("True treatment effect at cutoff:", 3, "\n")
cat("Sample size:", n, "\n")
cat("Treated:", sum(rd_data$treated), "| Control:", sum(1 - rd_data$treated), "\n")
```

We also create a separate fuzzy RD dataset where compliance is imperfect:

```{r simulate-fuzzy-data}
set.seed(123)
n_fuzzy <- 3000

fuzzy_data <- tibble(
  x     = runif(n_fuzzy, -1, 1),
  above = as.integer(x >= 0),

  # Compliance is imperfect:
  # Below cutoff: 15% treated (always-takers)
 # Above cutoff: 70% treated (compliers + always-takers)
  treated = rbinom(n_fuzzy, 1, prob = 0.15 + 0.55 * above),

  # Pre-determined covariates
  age    = 40 + 3 * x + rnorm(n_fuzzy, sd = 4),
  female = rbinom(n_fuzzy, 1, prob = 0.5),

  # Outcome: true LATE for compliers = 4
  y = 1 + 4 * treated + 1.2 * x + 0.6 * x^2 + rnorm(n_fuzzy, sd = 1)
) |>
  mutate(x_above = x * above)

cat("Fuzzy RD: First stage jump ~",
    round(mean(fuzzy_data$treated[fuzzy_data$above == 1]) -
          mean(fuzzy_data$treated[fuzzy_data$above == 0]), 3), "\n")
```

---

# 3. Visual RD Analysis

Visualization is the first and most important step in any RD analysis. Before estimating anything, the researcher should inspect the data graphically.

## 3.1 Raw Scatter Plot

A scatter plot of the outcome against the running variable reveals the discontinuity (or lack thereof) at the cutoff.

```{r plot-scatter, fig.width=7, fig.height=5}
ggplot(rd_data, aes(x = x, y = y, color = factor(treated))) +
  geom_point(alpha = 0.2, size = 0.8) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("Control", "Treated")
  ) +
  labs(
    x     = "Running Variable (X)",
    y     = "Outcome (Y)",
    color = "Group",
    title = "Raw Scatter Plot: Outcome vs. Running Variable"
  ) +
  causalverse::ama_theme()
```

## 3.2 Binned Scatter Plot

With many observations, a binned scatter plot is more informative. We divide the running variable into equal-width bins and plot the mean outcome within each bin.

```{r plot-binned, fig.width=7, fig.height=5}
n_bins <- 30

binned_data <- rd_data |>
  mutate(
    bin = cut(x, breaks = n_bins, labels = FALSE),
    side = ifelse(x >= 0, "Treated", "Control")
  ) |>
  group_by(bin, side) |>
  summarise(
    x_mean = mean(x),
    y_mean = mean(y),
    y_se   = sd(y) / sqrt(n()),
    .groups = "drop"
  )

ggplot(binned_data, aes(x = x_mean, y = y_mean, color = side)) +
  geom_point(size = 2.5) +
  geom_errorbar(
    aes(ymin = y_mean - 1.96 * y_se, ymax = y_mean + 1.96 * y_se),
    width = 0.02, alpha = 0.5
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(values = c("Control" = "steelblue", "Treated" = "firebrick")) +
  labs(
    x     = "Running Variable (Bin Means)",
    y     = "Outcome (Bin Means)",
    color = "Group",
    title = "Binned Scatter Plot with 95% CIs"
  ) +
  causalverse::ama_theme()
```

## 3.3 Local Polynomial Fits

We overlay local polynomial (loess) fits on each side of the cutoff. The gap between the two fitted curves at the cutoff is a visual estimate of the treatment effect.

```{r plot-local-poly, fig.width=7, fig.height=5}
ggplot(rd_data, aes(x = x, y = y, color = factor(treated))) +
  geom_point(alpha = 0.15, size = 0.6) +
  geom_smooth(
    method = "loess", span = 0.5, se = TRUE, linewidth = 1.2
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("Control", "Treated")
  ) +
  labs(
    x     = "Running Variable (X)",
    y     = "Outcome (Y)",
    color = "Group",
    title = "Local Polynomial Fits on Each Side of the Cutoff"
  ) +
  causalverse::ama_theme()
```

## 3.4 Linear Fits with Varying Bandwidths

It is useful to see how the visual discontinuity changes with the bandwidth around the cutoff.

```{r plot-bandwidth-visual, fig.width=8, fig.height=6}
bw_list <- c(0.25, 0.5, 0.75, 1.0)

bw_plot_data <- lapply(bw_list, function(h) {
  rd_data |>
    filter(abs(x) <= h) |>
    mutate(bandwidth = paste0("h = ", h))
}) |>
  bind_rows()

ggplot(bw_plot_data, aes(x = x, y = y, color = factor(treated))) +
  geom_point(alpha = 0.2, size = 0.6) +
  geom_smooth(method = "lm", se = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~bandwidth, scales = "free") +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("Control", "Treated")
  ) +
  labs(
    x     = "Running Variable",
    y     = "Outcome",
    color = "Group",
    title = "Linear Fits at Different Bandwidths"
  ) +
  causalverse::ama_theme()
```

---

# 4. The `rdrobust` Package (Calonico, Cattaneo, Titiunik 2014)

The `rdrobust` package is the **gold standard** for RD estimation in applied research. It implements:

- MSE-optimal bandwidth selection (Imbens and Kalyanaraman, 2012; Calonico, Cattaneo, and Titiunik, 2014)
- Bias-corrected local polynomial estimators
- Robust confidence intervals that account for bandwidth selection
- RD-specific plotting routines

The key insight is that conventional confidence intervals based on the MSE-optimal bandwidth are invalid because they do not account for the smoothing bias inherent in nonparametric estimation. The `rdrobust` package corrects for this bias and constructs valid confidence intervals.

## 4.1 Basic Estimation with `rdrobust()`

The main function `rdrobust()` estimates the RD treatment effect using local polynomial regression with optimal bandwidth selection.

```{r rdrobust-basic, eval=requireNamespace("rdrobust", quietly = TRUE)}
library(rdrobust)

# Basic RD estimation with default settings
# - Local linear regression (p = 1)
# - Triangular kernel
# - MSE-optimal bandwidth
rd_est <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0)
summary(rd_est)
```

The output reports:

- **Conventional**: the point estimate and confidence interval using the MSE-optimal bandwidth
- **Bias-corrected**: the point estimate after explicit bias correction
- **Robust**: the recommended confidence interval that accounts for both bias correction and bandwidth selection

**Always report the robust confidence interval.** The conventional CI is known to have poor coverage properties.

## 4.2 RD Plots with `rdplot()`

The `rdplot()` function creates publication-quality RD plots with evenly-spaced or quantile-spaced bins and fitted polynomials on each side of the cutoff.

```{r rdrobust-plot, eval=requireNamespace("rdrobust", quietly = TRUE)}
# Default RD plot: evenly-spaced bins with 4th-order polynomial
rdplot(
  y       = rd_data$y,
  x       = rd_data$x,
  c       = 0,
  title   = "RD Plot (rdplot)",
  x.label = "Running Variable",
  y.label = "Outcome"
)

# Quantile-spaced bins (equal number of observations per bin)
rdplot(
  y         = rd_data$y,
  x         = rd_data$x,
  c         = 0,
  binselect = "qs",
  title     = "RD Plot: Quantile-Spaced Bins",
  x.label   = "Running Variable",
  y.label   = "Outcome"
)

# Evenly-spaced bins with mimicking variance (IMSE-optimal)
rdplot(
  y         = rd_data$y,
  x         = rd_data$x,
  c         = 0,
  binselect = "esmv",
  title     = "RD Plot: IMSE-Optimal Bins",
  x.label   = "Running Variable",
  y.label   = "Outcome"
)
```

## 4.3 Bandwidth Sensitivity Analysis

A crucial robustness check is to show how the RD estimate changes with the bandwidth. The estimate should be relatively stable across a range of bandwidths around the optimal one.

```{r rdrobust-bandwidth, eval=requireNamespace("rdrobust", quietly = TRUE)}
# Get the MSE-optimal bandwidth first
rd_opt <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0)
h_opt  <- rd_opt$bws[1, 1]  # MSE-optimal bandwidth

cat("MSE-optimal bandwidth:", round(h_opt, 3), "\n")

# Estimate at a grid of bandwidths
bw_grid <- seq(0.5 * h_opt, 2 * h_opt, length.out = 10)

bw_sensitivity <- lapply(bw_grid, function(h) {
  est <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0, h = h)
  tibble(
    bandwidth = h,
    estimate  = est$coef["Conventional", ],
    ci_lower  = est$ci["Robust", 1],
    ci_upper  = est$ci["Robust", 2],
    n_eff_l   = est$N_h[1],
    n_eff_r   = est$N_h[2]
  )
}) |>
  bind_rows()

# Plot bandwidth sensitivity
ggplot(bw_sensitivity, aes(x = bandwidth, y = estimate)) +
  geom_point(size = 2) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.01) +
  geom_vline(xintercept = h_opt, linetype = "dashed", color = "blue",
             linewidth = 0.8) +
  geom_hline(yintercept = 3, linetype = "dotted", color = "red") +
  annotate("text", x = h_opt, y = max(bw_sensitivity$ci_upper) + 0.2,
           label = "MSE-optimal h", color = "blue", hjust = -0.1) +
  labs(
    x     = "Bandwidth",
    y     = "RD Estimate (Robust CIs)",
    title = "Bandwidth Sensitivity of the RD Estimate"
  ) +
  causalverse::ama_theme()
```

## 4.4 Different Kernel Choices

The kernel function weights observations by their distance from the cutoff. `rdrobust` supports three kernels:

- **Triangular** (default): $K(u) = (1 - |u|) \cdot \mathbf{1}(|u| \leq 1)$. Places most weight on observations closest to the cutoff.
- **Epanechnikov**: $K(u) = \frac{3}{4}(1 - u^2) \cdot \mathbf{1}(|u| \leq 1)$. Similar to triangular but smoother.
- **Uniform**: $K(u) = \frac{1}{2} \cdot \mathbf{1}(|u| \leq 1)$. Equal weight to all observations within the bandwidth.

```{r rdrobust-kernels, eval=requireNamespace("rdrobust", quietly = TRUE)}
kernels <- c("triangular", "epanechnikov", "uniform")

kernel_results <- lapply(kernels, function(k) {
  est <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0, kernel = k)
  tibble(
    kernel    = k,
    estimate  = est$coef["Conventional", ],
    robust_ci_lower = est$ci["Robust", 1],
    robust_ci_upper = est$ci["Robust", 2],
    bandwidth = est$bws[1, 1]
  )
}) |>
  bind_rows()

kernel_results

# Plot kernel comparison
ggplot(kernel_results, aes(x = kernel, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = robust_ci_lower, ymax = robust_ci_upper),
    width = 0.15
  ) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red") +
  labs(
    x     = "Kernel Function",
    y     = "RD Estimate",
    title = "RD Estimates by Kernel Choice"
  ) +
  causalverse::ama_theme()
```

## 4.5 Different Polynomial Orders

The polynomial order $p$ controls the flexibility of the local polynomial fit. The most common choices are:

- $p = 0$: Local constant (Nadaraya-Watson)
- $p = 1$: Local linear (default in `rdrobust`, recommended in most cases)
- $p = 2$: Local quadratic

Higher-order polynomials reduce bias but increase variance. Calonico, Cattaneo, and Titiunik (2014) recommend $p = 1$ as the default.

```{r rdrobust-polyorder, eval=requireNamespace("rdrobust", quietly = TRUE)}
poly_orders <- 0:3

poly_results <- lapply(poly_orders, function(p) {
  est <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0, p = p)
  tibble(
    poly_order = p,
    estimate   = est$coef["Conventional", ],
    robust_ci_lower = est$ci["Robust", 1],
    robust_ci_upper = est$ci["Robust", 2],
    bandwidth  = est$bws[1, 1]
  )
}) |>
  bind_rows()

poly_results

ggplot(poly_results, aes(x = factor(poly_order), y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = robust_ci_lower, ymax = robust_ci_upper),
    width = 0.15
  ) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red") +
  labs(
    x     = "Polynomial Order (p)",
    y     = "RD Estimate",
    title = "RD Estimates by Polynomial Order"
  ) +
  causalverse::ama_theme()
```

## 4.6 Including Covariates in `rdrobust`

Pre-determined covariates can be included in the RD estimation to improve precision. Covariates that predict the outcome reduce residual variance and narrow confidence intervals. Importantly, the point estimate should not change substantially when covariates are included (if it does, this suggests a problem with the design).

```{r rdrobust-covariates, eval=requireNamespace("rdrobust", quietly = TRUE)}
# Prepare covariate matrix
covs <- as.matrix(rd_data[, c("age", "female", "education", "income")])

# Estimation without covariates
rd_no_covs <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0)

# Estimation with covariates
rd_with_covs <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0, covs = covs)

cat("=== Without Covariates ===\n")
cat("Estimate:", round(rd_no_covs$coef["Conventional", ], 3), "\n")
cat("Robust CI: [", round(rd_no_covs$ci["Robust", 1], 3), ",",
    round(rd_no_covs$ci["Robust", 2], 3), "]\n\n")

cat("=== With Covariates ===\n")
cat("Estimate:", round(rd_with_covs$coef["Conventional", ], 3), "\n")
cat("Robust CI: [", round(rd_with_covs$ci["Robust", 1], 3), ",",
    round(rd_with_covs$ci["Robust", 2], 3), "]\n")
```

## 4.7 Clustered Standard Errors

When the running variable takes on discrete values (e.g., test scores with integer values) or when there are groups of related observations (e.g., students within schools), clustered standard errors may be needed. The `cluster` argument in `rdrobust` implements this.

```{r rdrobust-cluster, eval=requireNamespace("rdrobust", quietly = TRUE)}
# Add a cluster variable (e.g., school ID)
rd_data_clustered <- rd_data |>
  mutate(school_id = sample(1:50, n(), replace = TRUE))

# Estimation with clustered standard errors
rd_clustered <- rdrobust(
  y       = rd_data_clustered$y,
  x       = rd_data_clustered$x,
  c       = 0,
  cluster = rd_data_clustered$school_id
)
summary(rd_clustered)
```

## 4.8 Additional `rdrobust` Options

```{r rdrobust-advanced, eval=requireNamespace("rdrobust", quietly = TRUE)}
# Using CER-optimal bandwidth (coverage error rate optimal)
# Narrower than MSE-optimal; better coverage properties
rd_cer <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0, bwselect = "cerrd")
summary(rd_cer)

# Specifying different bandwidths on each side
rd_asym <- rdrobust(
  y  = rd_data$y,
  x  = rd_data$x,
  c  = 0,
  h  = c(0.4, 0.6)  # left bandwidth = 0.4, right bandwidth = 0.6
)
summary(rd_asym)

# Using a different bias bandwidth (b) relative to main bandwidth (h)
rd_rho <- rdrobust(y = rd_data$y, x = rd_data$x, c = 0, rho = 0.8)
summary(rd_rho)

# All available bandwidth selectors
bw_selectors <- c("mserd", "msetwo", "msesum", "msecomb1", "msecomb2",
                   "cerrd", "certwo", "cersum", "cercomb1", "cercomb2")

bw_comparison <- lapply(bw_selectors, function(bw) {
  est <- tryCatch(
    rdrobust(y = rd_data$y, x = rd_data$x, c = 0, bwselect = bw),
    error = function(e) NULL
  )
  if (!is.null(est)) {
    tibble(
      selector  = bw,
      h_left    = est$bws[1, 1],
      h_right   = est$bws[1, 2],
      estimate  = est$coef["Conventional", ],
      robust_ci = paste0("[", round(est$ci["Robust", 1], 3), ", ",
                         round(est$ci["Robust", 2], 3), "]")
    )
  }
}) |>
  bind_rows()

bw_comparison
```

---

# 5. The `rddensity` Package (Cattaneo, Jansson, Ma 2020)

The `rddensity` package implements formal tests for manipulation of the running variable. The test examines whether the density of the running variable is continuous at the cutoff. A discontinuity in the density is evidence that units are sorting (bunching) around the cutoff, which would violate the no-manipulation assumption.

The test is based on local polynomial density estimation. It is a refinement of the earlier McCrary (2008) test, using bias-corrected inference and robust confidence intervals analogous to `rdrobust`.

## 5.1 Histogram Check

Before running a formal test, a histogram provides a visual check:

```{r density-histogram, fig.width=7, fig.height=4}
ggplot(rd_data, aes(x = x)) +
  geom_histogram(
    bins = 50, fill = "steelblue", color = "white", alpha = 0.8
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "red", linewidth = 1) +
  labs(
    x     = "Running Variable",
    y     = "Count",
    title = "Density of the Running Variable Around the Cutoff"
  ) +
  causalverse::ama_theme()
```

Since our running variable is drawn from a uniform distribution, we observe no bunching. In real applications, the formal test is essential.

## 5.2 Formal Density Test with `rddensity()`

```{r rddensity-test, eval=requireNamespace("rddensity", quietly = TRUE)}
library(rddensity)

# Basic manipulation test
density_test <- rddensity(X = rd_data$x, c = 0)
summary(density_test)
```

The output reports:

- The estimated density on the left and right of the cutoff
- The T-statistic and p-value for the null hypothesis of continuity
- Both conventional and robust p-values

A large p-value (e.g., > 0.05) is consistent with no manipulation. A small p-value indicates evidence of manipulation.

## 5.3 Density Plots with `rdplotdensity()`

```{r rddensity-plot, eval=requireNamespace("rddensity", quietly = TRUE)}
# Density plot with confidence intervals
# density_test is created in the rddensity-test chunk above
rdplotdensity(
  rdd       = density_test,
  X         = rd_data$x,
  title     = "Density Continuity Test at the Cutoff",
  xlabel    = "Running Variable",
  ylabel    = "Density"
)
```

The plot shows the estimated density on each side of the cutoff with pointwise confidence intervals. Overlapping confidence intervals at the cutoff are consistent with no manipulation.

## 5.4 Advanced Options for `rddensity`

```{r rddensity-advanced, eval=requireNamespace("rddensity", quietly = TRUE)}
# Different polynomial orders for the density estimator
density_p2 <- rddensity(X = rd_data$x, c = 0, p = 2)
summary(density_p2)

# Specifying bandwidths manually
density_manual_bw <- rddensity(X = rd_data$x, c = 0, h = c(0.3, 0.3))
summary(density_manual_bw)

# Using different kernel
density_uni <- rddensity(X = rd_data$x, c = 0, kernel = "uniform")
summary(density_uni)

# Binomial test (useful with discrete running variables)
# When the running variable is discrete, the density test may not apply directly.
# A binomial test checks whether the share of observations just above the cutoff
# is consistent with random assignment.
n_just_below <- sum(rd_data$x >= -0.05 & rd_data$x < 0)
n_just_above <- sum(rd_data$x >= 0 & rd_data$x < 0.05)
binom.test(n_just_above, n_just_below + n_just_above, p = 0.5)
```

---

# 6. The `rdpower` Package (Cattaneo, Titiunik, Vazquez-Bare 2019)

The `rdpower` package provides tools for power calculations and sample size determination in RD designs. Power analysis is critical for planning new studies and for assessing whether existing studies had adequate power to detect effects of a given size.

## 6.1 Power Calculations with `rdpower()`

Given data (or summary statistics) and a hypothesized effect size, `rdpower()` computes the power of the RD test.

```{r rdpower-basic, eval=requireNamespace("rdpower", quietly = TRUE)}
library(rdpower)

# Power calculation using the data directly
# tau = 3 is the hypothesized treatment effect
power_result <- rdpower(
  data   = data.frame(y = rd_data$y, x = rd_data$x),
  cutoff = 0,
  tau    = 3
)
print(power_result)
```

## 6.2 Power for Different Effect Sizes

```{r rdpower-effect-sizes, eval=requireNamespace("rdpower", quietly = TRUE)}
# How does power change with the effect size?
effect_sizes <- seq(0.5, 5, by = 0.5)

power_by_tau <- lapply(effect_sizes, function(tau) {
  pw <- rdpower(
    data   = data.frame(y = rd_data$y, x = rd_data$x),
    cutoff = 0,
    tau    = tau
  )
  tibble(
    tau   = tau,
    power = pw$power
  )
}) |>
  bind_rows()

ggplot(power_by_tau, aes(x = tau, y = power)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  geom_hline(yintercept = 0.8, linetype = "dashed", color = "red") +
  annotate("text", x = max(effect_sizes), y = 0.82, label = "80% Power",
           color = "red", hjust = 1) +
  labs(
    x     = "Hypothesized Treatment Effect (tau)",
    y     = "Statistical Power",
    title = "Power Curve for RD Design"
  ) +
  causalverse::ama_theme()
```

## 6.3 Sample Size Calculations with `rdsampsi()`

Given a desired power level and hypothesized effect size, `rdsampsi()` determines the required sample size.

```{r rdpower-sampsi, eval=requireNamespace("rdpower", quietly = TRUE)}
# Required sample size for 80% power to detect tau = 3
sampsi_result <- rdsampsi(
  data   = data.frame(y = rd_data$y, x = rd_data$x),
  cutoff = 0,
  tau    = 3
)
print(sampsi_result)

# Required sample size for 80% power to detect tau = 1.5
# (smaller effect requires more data)
sampsi_small <- rdsampsi(
  data   = data.frame(y = rd_data$y, x = rd_data$x),
  cutoff = 0,
  tau    = 1.5
)
print(sampsi_small)

# Using data-based approach with different tau
sampsi_summary <- rdsampsi(
  data   = data.frame(y = rd_data$y, x = rd_data$x),
  cutoff = 0,
  tau    = 2
)
print(sampsi_summary)
```

## 6.4 Power with Covariates

```{r rdpower-covariates, eval=requireNamespace("rdpower", quietly = TRUE)}
# Covariates that predict the outcome can increase power
covs <- as.matrix(rd_data[, c("age", "education")])

power_with_covs <- rdpower(
  data   = data.frame(y = rd_data$y, x = rd_data$x),
  cutoff = 0,
  tau    = 3,
  covs   = covs
)
print(power_with_covs)
```

---

# 7. The `rdlocrand` Package (Cattaneo, Frandsen, Titiunik 2015)

The `rdlocrand` package implements the **local randomization** approach to RD inference. Instead of relying on continuity and local polynomial estimation, this approach treats the RD design as a local experiment within a narrow window around the cutoff. Within this window, treatment assignment is assumed to be as good as random.

The key advantages of the local randomization approach are:

- Valid inference even with very few observations near the cutoff
- No need for bandwidth selection for estimation (the window is selected based on balance, not MSE)
- Finite-sample exact inference via permutation tests

## 7.1 Window Selection with `rdwinselect()`

The first step is to select the window within which the local randomization assumption holds. `rdwinselect()` tests whether pre-determined covariates are balanced across the cutoff within progressively wider windows.

```{r rdlocrand-winselect, eval=requireNamespace("rdlocrand", quietly = TRUE)}
library(rdlocrand)

# Prepare covariates for balance testing
# In rdlocrand v1.1+: R = running variable (first arg), X = covariate matrix (second arg)
covariates <- as.matrix(rd_data[, c("age", "female", "education", "income")])

window_result <- rdwinselect(
  R        = rd_data$x,   # running variable
  X        = covariates,  # pre-treatment covariates
  cutoff   = 0,
  wmin     = 0.10,
  wstep    = 0.05,
  nwindows = 8,
  seed     = 42
)
summary(window_result)
```

The function returns a sequence of windows and p-values from balance tests. The recommended window is the largest window for which covariates remain balanced.

## 7.2 Randomization Inference with `rdrandinf()`

Once the window is selected, `rdrandinf()` performs randomization-based inference for the treatment effect.

```{r rdlocrand-randinf, eval=requireNamespace("rdlocrand", quietly = TRUE)}
# Randomization inference with the selected window
# Suppose the selected window is [-0.1, 0.1]
ri_result <- rdrandinf(
  Y      = rd_data$y,
  R      = rd_data$x,
  cutoff = 0,
  wl     = -0.1,  # left endpoint of window
  wr     = 0.1,   # right endpoint of window
  seed   = 42
)
summary(ri_result)
```

## 7.3 Bernoulli vs. Fixed Margins Randomization

`rdrandinf` supports different randomization mechanisms:

```{r rdlocrand-mechanisms, eval=requireNamespace("rdlocrand", quietly = TRUE)}
# Bernoulli randomization (default)
# Each unit is independently assigned to treatment with probability p
ri_bernoulli <- rdrandinf(
  Y      = rd_data$y,
  R      = rd_data$x,
  cutoff = 0,
  wl     = -0.1,
  wr     = 0.1,
  bernoulli = TRUE,
  seed   = 42
)

# Fixed margins randomization
# The number treated and control is fixed at the observed values
ri_fixed <- rdrandinf(
  Y      = rd_data$y,
  R      = rd_data$x,
  cutoff = 0,
  wl     = -0.1,
  wr     = 0.1,
  bernoulli = FALSE,
  seed   = 42
)

cat("Bernoulli p-value:", ri_bernoulli$p.value, "\n")
cat("Fixed margins p-value:", ri_fixed$p.value, "\n")
```

## 7.4 Sensitivity to Window Choice

```{r rdlocrand-sensitivity, eval=requireNamespace("rdlocrand", quietly = TRUE)}
# Estimate across a range of windows
windows <- seq(0.05, 0.30, by = 0.05)

ri_sensitivity <- lapply(windows, function(w) {
  est <- rdrandinf(
    Y      = rd_data$y,
    R      = rd_data$x,
    cutoff = 0,
    wl     = -w,
    wr     = w,
    seed   = 42
  )
  tibble(
    window_half = w,
    estimate    = est$obs.stat,
    p_value     = est$p.value,
    n_window    = sum(abs(rd_data$x) <= w)
  )
}) |>
  bind_rows()

ri_sensitivity
```

## 7.5 Randomization Inference with Covariates

```{r rdlocrand-covariates, eval=requireNamespace("rdlocrand", quietly = TRUE)}
# Including covariates for precision
ri_covs <- rdrandinf(
  Y       = rd_data$y,
  R       = rd_data$x,
  cutoff  = 0,
  wl      = -0.1,
  wr      = 0.1,
  covariates = as.matrix(rd_data[, c("age", "education")]),
  seed    = 42
)
summary(ri_covs)
```

---

# 8. The `rdmulti` Package (Cattaneo, Keele, Titiunik, Vazquez-Bare 2021)

The `rdmulti` package extends the RD framework to settings with **multiple cutoffs** or **multiple running variables**. These settings are common in practice:

- Multiple cutoffs: Different states set different eligibility thresholds for the same program
- Multiple scores: Eligibility depends on multiple test scores simultaneously (e.g., math and reading)

## 8.1 Simulated Multi-Cutoff Data

```{r rdmulti-sim, eval=requireNamespace("rdmulti", quietly = TRUE)}
library(rdmulti)

# Simulate data with multiple cutoffs
set.seed(2024)
n_multi <- 3000

multi_data <- tibble(
  # Running variable
  x = runif(n_multi, -2, 2),

  # Three different cutoffs (e.g., different states)
  cutoff_group = sample(c("A", "B", "C"), n_multi, replace = TRUE),
  cutoff_value = case_when(
    cutoff_group == "A" ~ -0.5,
    cutoff_group == "B" ~ 0.0,
    cutoff_group == "C" ~ 0.5
  ),

  # Treatment: above group-specific cutoff
  treated = as.integer(x >= cutoff_value),

  # Outcome: different effects at different cutoffs
  effect = case_when(
    cutoff_group == "A" ~ 2.0,
    cutoff_group == "B" ~ 3.0,
    cutoff_group == "C" ~ 1.5
  ),
  y = 1 + effect * treated + 0.8 * x + rnorm(n_multi, sd = 0.8)
)
```

## 8.2 Multi-Cutoff RD with `rdmc()`

The `rdmc()` function estimates treatment effects at each cutoff separately and then combines them into an overall estimate.

```{r rdmulti-rdmc, eval=requireNamespace("rdmulti", quietly = TRUE)}
# Multi-cutoff RD estimation
mc_result <- rdmc(
  Y = multi_data$y,
  X = multi_data$x,
  C = multi_data$cutoff_value
)
summary(mc_result)
```

The function reports:

- Cutoff-specific estimates with confidence intervals
- A pooled estimate (weighted average across cutoffs)
- Effective sample sizes at each cutoff

## 8.3 Multi-Score RD with `rdms()`

When treatment depends on multiple running variables simultaneously, `rdms()` handles the multi-dimensional boundary.

```{r rdmulti-rdms, eval=requireNamespace("rdmulti", quietly = TRUE) && requireNamespace("rdrobust", quietly = TRUE)}
# rdms: Multiple cutoffs along a single running variable
# Classic example: multiple school district thresholds
set.seed(42)
n_ms <- 3000

# Running variable: test score (e.g., 0-100)
x_score <- runif(n_ms, 40, 100)

# Three cutoffs: 60, 70, 80 determine three different program levels
# Each unit is associated with its nearest cutoff
C_vec <- c(60, 70, 80)
cutoff_assign <- cut(x_score, breaks = c(-Inf, 65, 75, Inf),
                     labels = c(1, 2, 3)) |> as.integer()
C_assigned  <- C_vec[cutoff_assign]

# Treatment: above own cutoff
treated_ms <- as.integer(x_score >= C_assigned)

y_ms <- 5 + 3 * treated_ms + 0.05 * x_score + rnorm(n_ms, sd = 1.5)

# Estimate RD at each cutoff separately using rdrobust
cat("=== RD Estimates at Three Cutoffs ===\n")
for (cv in C_vec) {
  idx <- abs(x_score - cv) <= 15  # local window
  fit <- rdrobust::rdrobust(y = y_ms[idx], x = x_score[idx], c = cv)
  cat(sprintf("Cutoff = %d: tau = %.3f (SE = %.3f, p = %.3f, N = %d)\n",
              cv, fit$coef[3], fit$se[3], fit$pv[3], sum(idx)))
}
```

## 8.4 Visualization for Multi-Cutoff Designs

```{r rdmulti-plot, eval=requireNamespace("rdmulti", quietly = TRUE) && requireNamespace("rdrobust", quietly = TRUE)}
# Plot RD at each cutoff separately
cutoff_labels <- c("A" = -0.5, "B" = 0.0, "C" = 0.5)

for (grp in names(cutoff_labels)) {
  sub_data <- multi_data |> filter(cutoff_group == grp)
  c_val <- cutoff_labels[grp]

  rdplot(
    y       = sub_data$y,
    x       = sub_data$x,
    c       = c_val,
    title   = paste0("RD Plot: Cutoff Group ", grp, " (c = ", c_val, ")"),
    x.label = "Running Variable",
    y.label = "Outcome"
  )
}
```

---

# 9. RD Bounds with the `rdbounds` Package and `causalverse`

When manipulation of the running variable is present or cannot be ruled out, standard RD estimates may be biased. Gerard, Rokkanen, and Rothe (2020) develop a framework for **bounding** the treatment effect under varying assumptions about the share of "always-assigned" units -- those who manipulate their running variable to receive treatment.

The `rdbounds` package computes these bounds, and `causalverse::plot_rd_aa_share()` provides a clean visualization.

## 9.1 Conceptual Framework

The key idea is:

- Let $\tau$ denote the share of always-assigned (manipulating) units
- For each assumed value of $\tau$, the treatment effect is partially identified (bounded)
- As $\tau$ increases (more manipulation assumed), the bounds widen
- At $\tau = 0$ (no manipulation), we recover the standard RD point estimate

The data itself provides an estimate $\hat{\tau}$ of the manipulation share (from the density discontinuity).

## 9.2 Full `rdbounds` Workflow for Sharp RD

```{r rdbounds-srd, eval=requireNamespace("rdbounds", quietly = TRUE)}
library(rdbounds)

# Step 1: Generate sample data from rdbounds
set.seed(1)
rd_bound_data <- rdbounds::rdbounds_sampledata(10000, covs = FALSE)

# Step 2: Estimate bounds
rdbounds_est <- rdbounds::rdbounds(
  y              = rd_bound_data$y,
  x              = rd_bound_data$x,
  treatment      = rd_bound_data$treatment,
  c              = 0,
  discrete_x     = FALSE,
  discrete_y     = FALSE,
  bwsx           = c(.2, .5),
  bwy            = 1,
  kernel         = "epanechnikov",
  orders         = 1,
  evaluation_ys  = seq(from = 0, to = 15, by = 1),
  refinement_A   = TRUE,
  refinement_B   = TRUE,
  right_effects  = TRUE,
  potential_taus  = c(.025, .05, .1, .2),
  yextremes      = c(0, 15),
  num_bootstraps = 5
)

# Step 3: Visualize bounds for SRD
causalverse::plot_rd_aa_share(
  data       = rdbounds_est,
  rd_type    = "SRD",
  x_label    = "Share of Always-assigned Units",
  y_label    = "ATE",
  plot_title = "RD Bounds (Sharp RD)",
  tau        = TRUE,
  bounds_CI  = TRUE,
  ref_line   = 0,
  show_naive = TRUE
)
```

The resulting plot displays:

- **Solid lines**: upper and lower bounds on the treatment effect as a function of the assumed share of always-assigned units
- **Dotted lines**: confidence intervals for those bounds
- **Vertical solid line**: the estimated share of always-assigned units ($\hat{\tau}$)
- **Horizontal dashed line**: reference line (typically zero, to assess whether the treatment effect can be signed)
- **Point at zero**: the naive (standard) RD estimate

## 9.3 Full `rdbounds` Workflow for Fuzzy RD

```{r rdbounds-frd, eval=requireNamespace("rdbounds", quietly = TRUE)}
# Visualize bounds for FRD
causalverse::plot_rd_aa_share(
  data       = rdbounds_est,
  rd_type    = "FRD",
  x_label    = "Share of Always-assigned Units",
  y_label    = "LATE",
  plot_title = "RD Bounds (Fuzzy RD)",
  tau        = TRUE,
  tau_CI     = TRUE,   # show CIs for tau-hat
  bounds_CI  = TRUE,
  ref_line   = 0,
  show_naive = TRUE
)
```

## 9.4 Customizing the Bounds Plot

The `plot_rd_aa_share()` function accepts several customization arguments:

```{r rdbounds-custom, eval=requireNamespace("rdbounds", quietly = TRUE)}
# Minimal plot: just the bounds, no extras
causalverse::plot_rd_aa_share(
  data       = rdbounds_est,
  rd_type    = "SRD",
  tau        = FALSE,
  tau_CI     = FALSE,
  bounds_CI  = FALSE,
  ref_line   = NULL,
  show_naive = FALSE,
  plot_title = "Manipulation Bounds (Minimal)"
)

# Full plot with all options
causalverse::plot_rd_aa_share(
  data       = rdbounds_est,
  rd_type    = "SRD",
  tau        = TRUE,
  tau_CI     = TRUE,
  bounds_CI  = TRUE,
  ref_line   = 0,
  show_naive = TRUE,
  plot_title = "Manipulation Bounds (Full)",
  subtitle   = "Gerard, Rokkanen, and Rothe (2020)"
)
```

---

# 10. Fuzzy RD in Detail

In a Fuzzy RD design, the cutoff does not perfectly determine treatment. Instead, the probability of treatment jumps at the cutoff but does not go from 0 to 1. This setup requires an instrumental variables (IV) approach.

## 10.1 First Stage: Cutoff Predicts Treatment

The first stage verifies that crossing the cutoff significantly predicts treatment receipt.

```{r fuzzy-first-stage}
# First stage regression
first_stage <- feols(treated ~ above + x + x_above, data = fuzzy_data)
summary(first_stage)

# Visual first stage
ggplot(fuzzy_data, aes(x = x, y = treated)) +
  stat_summary_bin(
    fun = mean, bins = 30, geom = "point",
    size = 2, color = "steelblue"
  ) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  geom_smooth(
    data = fuzzy_data |> filter(x < 0),
    method = "lm", color = "steelblue", se = TRUE
  ) +
  geom_smooth(
    data = fuzzy_data |> filter(x >= 0),
    method = "lm", color = "firebrick", se = TRUE
  ) +
  labs(
    x     = "Running Variable",
    y     = "P(Treatment)",
    title = "First Stage: Probability of Treatment"
  ) +
  causalverse::ama_theme()
```

The first-stage F-statistic should be well above 10 (the Staiger-Stock rule of thumb) to avoid weak instrument problems.

## 10.2 Reduced Form: Cutoff Predicts Outcome

```{r fuzzy-reduced-form}
# Reduced form: effect of crossing cutoff on outcome
reduced_form <- feols(y ~ above + x + x_above, data = fuzzy_data)
summary(reduced_form)
```

The reduced form shows the intention-to-treat (ITT) effect -- the effect of being assigned above the cutoff on the outcome.

## 10.3 IV Estimation with `fixest::feols()`

The Fuzzy RD estimate is the ratio of the reduced form to the first stage. We can obtain this directly using the IV syntax in `fixest`:

```{r fuzzy-iv-estimate}
# Fuzzy RD via IV
# treated is endogenous; above is the instrument
fuzzy_iv <- feols(y ~ x + x_above | 0 | treated ~ above, data = fuzzy_data)
summary(fuzzy_iv)
```

The IV coefficient on `treated` estimates the LATE for compliers at the cutoff. In our simulation, the true effect is 4.

## 10.4 Bandwidth Sensitivity for Fuzzy RD

```{r fuzzy-bandwidth, fig.width=7, fig.height=4}
bandwidths_fuzzy <- c(0.2, 0.3, 0.5, 0.75, 1.0)

fuzzy_bw_results <- lapply(bandwidths_fuzzy, function(h) {
  sub <- fuzzy_data |> filter(abs(x) <= h)
  est <- tryCatch({
    feols(y ~ x + x_above | 0 | treated ~ above, data = sub)
  }, error = function(e) NULL)

  if (!is.null(est)) {
    tibble(
      bandwidth = h,
      n_obs     = nrow(sub),
      estimate  = coef(est)["fit_treated"],
      se        = se(est)["fit_treated"]
    )
  }
}) |>
  bind_rows()

ggplot(fuzzy_bw_results, aes(x = factor(bandwidth), y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
                width = 0.1) +
  geom_hline(yintercept = 4, linetype = "dashed", color = "red") +
  labs(
    x     = "Bandwidth",
    y     = "Fuzzy RD Estimate (LATE)",
    title = "Fuzzy RD: IV Estimates by Bandwidth"
  ) +
  causalverse::ama_theme()
```

## 10.5 Fuzzy RD with `rdrobust`

The `rdrobust` package directly supports Fuzzy RD via the `fuzzy` argument:

```{r fuzzy-rdrobust, eval=requireNamespace("rdrobust", quietly = TRUE)}
library(rdrobust)

# Fuzzy RD estimation with rdrobust
fuzzy_robust <- rdrobust(
  y     = fuzzy_data$y,
  x     = fuzzy_data$x,
  c     = 0,
  fuzzy = fuzzy_data$treated  # pass the treatment variable
)
summary(fuzzy_robust)

# Compare with the manual IV
cat("\nManual IV estimate:", round(coef(fuzzy_iv)["fit_treated"], 3), "\n")
cat("rdrobust fuzzy estimate:", round(fuzzy_robust$coef["Conventional", ], 3), "\n")
```

---

# 11. Covariate Balance Tests

If the RD design is valid, pre-determined covariates should vary smoothly through the cutoff. A discontinuity in a covariate at the cutoff suggests that units on either side differ in ways beyond just treatment status, which would undermine the identification strategy.

## 11.1 Covariate-as-Outcome Tests

The standard approach uses each covariate as the "outcome" in an RD regression. The estimated "effect" at the cutoff should be close to zero and statistically insignificant.

```{r balance-tests}
# Test each covariate for a discontinuity at the cutoff
covariates_to_test <- c("age", "female", "education", "income")

balance_results <- lapply(covariates_to_test, function(cov) {
  form <- as.formula(paste(cov, "~ treated + x + x_treated"))
  est  <- feols(form, data = rd_data)
  tibble(
    covariate = cov,
    estimate  = coef(est)["treated"],
    se        = se(est)["treated"],
    p_value   = pvalue(est)["treated"]
  )
}) |>
  bind_rows()

balance_results
```

## 11.2 Visual Balance Check

```{r balance-plot, fig.width=7, fig.height=5}
ggplot(balance_results, aes(x = covariate, y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
    width = 0.15
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "red") +
  labs(
    x     = "Pre-determined Covariate",
    y     = "Estimated Discontinuity at Cutoff",
    title = "Covariate Balance Check at the Cutoff"
  ) +
  causalverse::ama_theme()
```

All estimates are close to zero with confidence intervals that include zero, consistent with a valid RD design.

## 11.3 Covariate Balance with `rdrobust`

For a more rigorous test using the optimal bandwidth selection of `rdrobust`:

```{r balance-rdrobust, eval=requireNamespace("rdrobust", quietly = TRUE)}
library(rdrobust)

# Test each covariate with rdrobust
balance_robust <- lapply(covariates_to_test, function(cov) {
  est <- rdrobust(y = rd_data[[cov]], x = rd_data$x, c = 0)
  tibble(
    covariate     = cov,
    estimate      = est$coef["Conventional", ],
    robust_ci_low = est$ci["Robust", 1],
    robust_ci_up  = est$ci["Robust", 2],
    p_value_robust = est$pv["Robust", ]
  )
}) |>
  bind_rows()

balance_robust
```

## 11.4 Visual Covariate Plots

```{r balance-visual, fig.width=8, fig.height=8}
# Plot each covariate against the running variable
cov_plot_data <- rd_data |>
  select(x, treated, age, female, education, income) |>
  tidyr::pivot_longer(
    cols = c(age, female, education, income),
    names_to = "covariate",
    values_to = "value"
  )

ggplot(cov_plot_data, aes(x = x, y = value, color = factor(treated))) +
  geom_point(alpha = 0.1, size = 0.5) +
  geom_smooth(method = "loess", span = 0.5, se = TRUE) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  facet_wrap(~covariate, scales = "free_y") +
  scale_color_manual(
    values = c("0" = "steelblue", "1" = "firebrick"),
    labels = c("Control", "Treated")
  ) +
  labs(
    x     = "Running Variable",
    y     = "Covariate Value",
    color = "Group",
    title = "Covariate Smoothness Through the Cutoff"
  ) +
  causalverse::ama_theme()
```

---

# 12. Lee Bounds and the Connection to RD

Lee (2009) bounds are primarily used in settings with sample selection (attrition), but the bounding approach has connections to RD designs with manipulation. When some units manipulate their running variable to receive treatment, this is analogous to selection, and bounding methods can help.

The `causalverse::lee_bounds()` function computes Lee bounds for the average direct effect for always-takers, using the framework from Kwon and Roth (2023). While originally developed for mediation analysis, this can be applied to RD-adjacent settings where treatment effects need to be bounded.

## 12.1 Using `lee_bounds()` from `causalverse`

```{r lee-bounds-example, eval=requireNamespace("MedBounds", quietly = TRUE)}
# Example: Lee bounds for always-takers
# Requires MedBounds package (now TestMechs)
# d = treatment, m = mediator, y = outcome

# In an RD context, one can think of:
# d = above the cutoff (instrument)
# m = actual treatment receipt (endogenous)
# y = outcome

lee_result <- causalverse::lee_bounds(
  df       = fuzzy_data,
  d        = "above",
  m        = "treated",
  y        = "y",
  numdraws = 500
)

lee_result

# With a known ratio of complier-to-always-taker effects
# (point-identifies the effect)
lee_point <- causalverse::lee_bounds(
  df         = fuzzy_data,
  d          = "above",
  m          = "treated",
  y          = "y",
  c_at_ratio = 1,
  numdraws   = 500
)

lee_point
```

---

# 13. Sharp RD Estimation with `fixest`

While `rdrobust` is the gold standard for nonparametric RD estimation, a parametric approach using `fixest::feols()` is useful for understanding the mechanics and for quick analyses.

## 13.1 Local Linear Regression

```{r sharp-rd-fixest}
# Full sample: local linear with interaction
est_full <- feols(y ~ treated + x + x_treated, data = rd_data)
summary(est_full)
```

## 13.2 Bandwidth Sensitivity

```{r sharp-bandwidth, fig.width=7, fig.height=4}
bandwidths <- c(0.15, 0.25, 0.35, 0.5, 0.75, 1.0)

bw_results <- lapply(bandwidths, function(h) {
  sub <- rd_data |> filter(abs(x) <= h)
  est <- feols(y ~ treated + x + x_treated, data = sub)
  tibble(
    bandwidth = h,
    n_obs     = nrow(sub),
    estimate  = coef(est)["treated"],
    se        = se(est)["treated"]
  )
}) |>
  bind_rows()

bw_results

ggplot(bw_results, aes(x = factor(bandwidth), y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
    width = 0.1
  ) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red") +
  labs(
    x     = "Bandwidth",
    y     = "Estimated Treatment Effect",
    title = "Sharp RD Estimates by Bandwidth (fixest)"
  ) +
  causalverse::ama_theme()
```

The dashed red line marks the true effect of 3.

## 13.3 Higher-Order Polynomial Specifications

```{r sharp-polynomial}
# Local quadratic
rd_data <- rd_data |>
  mutate(
    x2 = x^2,
    x2_treated = x^2 * treated,
    x3 = x^3,
    x3_treated = x^3 * treated
  )

# Linear
est_p1 <- feols(y ~ treated + x + x_treated, data = rd_data)

# Quadratic
est_p2 <- feols(y ~ treated + x + x_treated + x2 + x2_treated, data = rd_data)

# Cubic
est_p3 <- feols(y ~ treated + x + x_treated + x2 + x2_treated + x3 + x3_treated,
                data = rd_data)

poly_comparison <- tibble(
  polynomial = c("Linear (p=1)", "Quadratic (p=2)", "Cubic (p=3)"),
  estimate   = c(coef(est_p1)["treated"], coef(est_p2)["treated"],
                 coef(est_p3)["treated"]),
  se         = c(se(est_p1)["treated"], se(est_p2)["treated"],
                 se(est_p3)["treated"])
)

poly_comparison
```

## 13.4 Triangular Kernel Weights

To mimic the triangular kernel weighting used in `rdrobust`, we can apply weights in `feols()`:

```{r sharp-kernel-weights}
# Triangular kernel: weight = 1 - |x|/h for |x| <= h
h <- 0.5
rd_data_weighted <- rd_data |>
  filter(abs(x) <= h) |>
  mutate(kernel_weight = 1 - abs(x) / h)

est_weighted <- feols(
  y ~ treated + x + x_treated,
  data    = rd_data_weighted,
  weights = ~kernel_weight
)
summary(est_weighted)
```

---

# 14. Heterogeneous Treatment Effects in RD with `rdhte`

The `rdhte` package [@calonico2025rdhte] is a recent addition (2025) to the RD ecosystem by Calonico, Cattaneo, and co-authors. It provides tools for estimating **covariate-moderated** treatment effects at the cutoff --- that is, how the RD treatment effect varies as a function of observed covariates.

Key functions include:

- `rdhte()`: Estimate heterogeneous (conditional) treatment effects at the RD cutoff for different covariate subgroups.
- `rdbwhte()`: Data-driven bandwidth selection optimized for heterogeneous treatment effect estimation.
- `rdhte_lincom()`: Compute linear combinations of heterogeneous effects (e.g., weighted averages, contrasts between subgroups).

## 14.1 Simulated Example

```{r rdhte-example, eval = requireNamespace("rdhte", quietly = TRUE)}
library(rdhte)

set.seed(42)
n <- 2000

# Running variable centered at 0
x <- runif(n, -1, 1)

# Binary covariate: group membership
group <- rbinom(n, 1, 0.5)

# Treatment indicator (sharp RD at cutoff = 0)
treated <- as.integer(x >= 0)

# Heterogeneous treatment effect: group 1 has larger effect
tau <- ifelse(group == 1, 3, 1)
y <- 0.5 * x + treated * tau + 0.3 * group + rnorm(n, sd = 0.5)

dat_hte <- data.frame(y = y, x = x, group = group)

# Estimate heterogeneous treatment effects by group
hte_fit <- rdhte(y = dat_hte$y, x = dat_hte$x, covs_hte = dat_hte$group)
summary(hte_fit)
```

## 14.2 Bandwidth Selection for HTE

```{r rdhte-bandwidth, eval = requireNamespace("rdhte", quietly = TRUE)}
# Optimal bandwidth for heterogeneous effects
bw_hte <- rdbwhte(y = dat_hte$y, x = dat_hte$x, covs_hte = dat_hte$group)
summary(bw_hte)
```

## 14.3 Linear Combinations of Effects

```{r rdhte-lincom, eval = requireNamespace("rdhte", quietly = TRUE)}
# Test whether treatment effect differs across groups
lincom_fit <- rdhte_lincom(hte_fit)
summary(lincom_fit)
```

The `rdhte` framework is especially useful when researchers want to move beyond the average treatment effect at the cutoff and explore whether the policy effect is concentrated among particular subpopulations.

---

# 15. Boundary/Geographic RD with `rd2d`

The `rd2d` package [@cattaneo2025rd2d] by Cattaneo, Titiunik, and Yu (2025) extends RD methods to settings with **bivariate running variables** --- most commonly geographic or spatial RD designs where treatment assignment depends on a unit's location relative to a boundary in two-dimensional space.

Key features:

- `rd2d()`: Estimate RD treatment effects with two running variables (e.g., latitude and longitude).
- Accommodates arbitrary boundary shapes in geographic discontinuity designs.
- Provides data-driven bandwidth selection in two dimensions.

## 15.1 Simulated Example

```{r rd2d-example, eval = requireNamespace("rd2d", quietly = TRUE)}
library(rd2d)

set.seed(42)
n <- 2000

# Two running variables: coordinates on a plane
x1 <- runif(n, -1, 1)
x2 <- runif(n, -1, 1)

# Treatment: assigned when above a linear boundary x2 > 0.3 * x1
boundary_val <- 0.3 * x1
treated_2d <- as.integer(x2 >= boundary_val)

# Outcome with treatment effect of 2
y_2d <- 0.5 * x1 - 0.3 * x2 + 2 * treated_2d + rnorm(n, sd = 0.8)

# Distance to boundary (signed)
dist_to_boundary <- x2 - boundary_val

dat_2d <- data.frame(
  y  = y_2d,
  x1 = x1,
  x2 = x2,
  d  = treated_2d
)

# Estimate geographic RD
rd2d_fit <- rd2d(
  Y  = dat_2d$y,
  X1 = dat_2d$x1,
  X2 = dat_2d$x2,
  zvar = dat_2d$d
)
summary(rd2d_fit)
```

## 15.2 Rich Geographic Visualizations

Geographic RD designs are best understood visually. The four panels below mirror how applied researchers present spatial discontinuities in top journals.

### Panel 1: Treatment Boundary Map

```{r rd2d-map1, fig.width = 8, fig.height = 6}
# ── Simulate realistic geographic coordinates ──────────────────────────────
set.seed(42)
n_geo <- 3000

# Simulate "longitude" (-100 to -80) and "latitude" (35 to 50)
# Boundary: a north-south state line at longitude = -90 (e.g., Missouri/Kansas)
lon <- runif(n_geo, -100, -80)
lat <- runif(n_geo, 35, 50)

# Treatment: units east of the boundary (lon > -90) are treated
boundary_lon <- -90
treated_geo  <- as.integer(lon > boundary_lon)

# Outcome: income ~ treatment + location + noise
income <- 30000 + 5000 * treated_geo +
          800  * (lat - 42) +
          200  * (lon + 90) +
          rnorm(n_geo, 0, 3000)

geo_df <- data.frame(lon = lon, lat = lat,
                     treated = factor(treated_geo,
                                      labels = c("Control (West)", "Treated (East)")),
                     income = income)

# Treatment boundary map
ggplot2::ggplot(geo_df, ggplot2::aes(x = lon, y = lat, colour = treated)) +
  ggplot2::geom_point(alpha = 0.25, size = 0.9) +
  ggplot2::geom_vline(xintercept = boundary_lon,
                      linewidth = 1.4, colour = "black", linetype = "solid") +
  ggplot2::annotate("label", x = boundary_lon, y = 49,
                    label = "Policy boundary\n(lon = -90)", size = 3.5,
                    colour = "black", fill = "white", label.size = 0.3) +
  ggplot2::annotate("text", x = -96, y = 36.5,
                    label = "Control region\n(No policy)", size = 3.5,
                    colour = "#4575b4", fontface = "italic") +
  ggplot2::annotate("text", x = -84, y = 36.5,
                    label = "Treated region\n(Policy active)", size = 3.5,
                    colour = "#d73027", fontface = "italic") +
  ggplot2::scale_colour_manual(
    values = c("Control (West)" = "#4575b4", "Treated (East)" = "#d73027"),
    name   = NULL) +
  ggplot2::labs(
    title    = "Geographic Regression Discontinuity Design",
    subtitle = "Treatment is assigned to units east of a state-line boundary",
    x        = "Longitude", y = "Latitude"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position  = "bottom",
    plot.title       = ggplot2::element_text(face = "bold", size = 14),
    panel.grid.minor = ggplot2::element_blank()
  )
```

### Panel 2: Outcome Heat Map (Choropleth)

```{r rd2d-map2, fig.width = 8, fig.height = 6}
# Bin geographic space into grid cells and show average income
geo_df$lon_bin <- cut(geo_df$lon, breaks = 20)
geo_df$lat_bin <- cut(geo_df$lat, breaks = 15)

grid_df <- geo_df |>
  dplyr::group_by(lon_bin, lat_bin) |>
  dplyr::summarise(
    avg_income = mean(income),
    lon_mid    = mean(lon),
    lat_mid    = mean(lat),
    .groups    = "drop"
  )

ggplot2::ggplot(grid_df, ggplot2::aes(x = lon_mid, y = lat_mid,
                                       fill = avg_income)) +
  ggplot2::geom_tile(width = 1, height = 0.95, alpha = 0.85) +
  ggplot2::geom_vline(xintercept = boundary_lon,
                      linewidth = 1.8, colour = "white", linetype = "dashed") +
  ggplot2::scale_fill_gradientn(
    colours = c("#313695","#4575b4","#74add1","#abd9e9",
                "#fee090","#fdae61","#f46d43","#d73027","#a50026"),
    name    = "Avg. Income ($)",
    labels  = scales::comma
  ) +
  ggplot2::annotate("label", x = boundary_lon, y = 49.2,
                    label = "Boundary", size = 3.5,
                    colour = "white", fill = "grey30", label.size = 0) +
  ggplot2::labs(
    title    = "Outcome Surface: Average Income by Geographic Cell",
    subtitle = "Jump in income at the policy boundary is clearly visible",
    x        = "Longitude", y = "Latitude"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    plot.title       = ggplot2::element_text(face = "bold", size = 14),
    legend.position  = "right",
    panel.grid       = ggplot2::element_blank()
  )
```

### Panel 3: Density of Units Near the Boundary

```{r rd2d-map3, fig.width = 8, fig.height = 5}
# Units within 5 degrees of the boundary — the "bandwidth window"
bw <- 5
near_df <- geo_df[abs(geo_df$lon - boundary_lon) <= bw, ]

ggplot2::ggplot(near_df, ggplot2::aes(x = lon, y = lat, colour = treated)) +
  ggplot2::geom_point(alpha = 0.35, size = 1.2) +
  ggplot2::geom_vline(xintercept = boundary_lon,
                      linewidth = 1.6, colour = "black") +
  ggplot2::geom_vline(xintercept = c(boundary_lon - bw, boundary_lon + bw),
                      linewidth = 0.7, colour = "grey50", linetype = "dotted") +
  ggplot2::scale_colour_manual(
    values = c("Control (West)" = "#4575b4", "Treated (East)" = "#d73027"),
    name   = NULL) +
  ggplot2::annotate("rect",
    xmin = boundary_lon - bw, xmax = boundary_lon + bw,
    ymin = -Inf, ymax = Inf, alpha = 0.05, fill = "gold") +
  ggplot2::annotate("text", x = boundary_lon - bw/2, y = 49.5,
                    label = "Bandwidth\nwindow", size = 3, colour = "grey40") +
  ggplot2::annotate("text", x = boundary_lon + bw/2, y = 49.5,
                    label = "Bandwidth\nwindow", size = 3, colour = "grey40") +
  ggplot2::labs(
    title    = "Zoom-In: Units Within Bandwidth of the Boundary",
    subtitle = paste0("Local comparison window: ±", bw, "° longitude"),
    x        = "Longitude", y = "Latitude"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position  = "bottom",
    plot.title       = ggplot2::element_text(face = "bold", size = 14)
  )
```

### Panel 4: The Jump at the Boundary (RD Plot)

```{r rd2d-map4, fig.width = 8, fig.height = 5}
# Bin by distance to boundary and plot mean income — the canonical RD plot
geo_df$dist_to_bound <- geo_df$lon - boundary_lon

geo_df$dist_bin <- cut(geo_df$dist_to_bound,
                       breaks = seq(-10, 10, by = 0.5), include.lowest = TRUE)

rd_plot_df <- geo_df |>
  dplyr::group_by(dist_bin) |>
  dplyr::summarise(
    avg_income = mean(income),
    dist_mid   = mean(dist_to_bound),
    n          = dplyr::n(),
    se         = sd(income) / sqrt(dplyr::n()),
    .groups    = "drop"
  ) |>
  dplyr::filter(!is.na(dist_mid)) |>
  dplyr::mutate(side = ifelse(dist_mid < 0, "Control (West)", "Treated (East)"))

ggplot2::ggplot(rd_plot_df,
                ggplot2::aes(x = dist_mid, y = avg_income,
                             colour = side, fill = side)) +
  ggplot2::geom_ribbon(ggplot2::aes(
    ymin = avg_income - 1.96 * se,
    ymax = avg_income + 1.96 * se), alpha = 0.15, colour = NA) +
  ggplot2::geom_point(size = 2.5) +
  ggplot2::geom_smooth(data = rd_plot_df[rd_plot_df$dist_mid < 0, ],
                       method = "lm", se = FALSE, linewidth = 1.2) +
  ggplot2::geom_smooth(data = rd_plot_df[rd_plot_df$dist_mid > 0, ],
                       method = "lm", se = FALSE, linewidth = 1.2) +
  ggplot2::geom_vline(xintercept = 0, linewidth = 1.2,
                      colour = "black", linetype = "dashed") +
  ggplot2::scale_colour_manual(
    values = c("Control (West)" = "#4575b4", "Treated (East)" = "#d73027"),
    name = NULL) +
  ggplot2::scale_fill_manual(
    values = c("Control (West)" = "#4575b4", "Treated (East)" = "#d73027"),
    name = NULL) +
  ggplot2::scale_y_continuous(labels = scales::comma) +
  ggplot2::annotate("segment",
    x = 0.3, xend = 0.3,
    y  = mean(rd_plot_df$avg_income[rd_plot_df$side == "Control (West)"]),
    yend = mean(rd_plot_df$avg_income[rd_plot_df$side == "Treated (East)"]),
    arrow = ggplot2::arrow(ends = "both", length = ggplot2::unit(0.08, "inches")),
    colour = "black", linewidth = 0.8) +
  ggplot2::annotate("text", x = 1.5, y = mean(rd_plot_df$avg_income),
                    label = paste0("Jump ≈ $", scales::comma(5000)),
                    size = 3.5, colour = "black", fontface = "bold") +
  ggplot2::labs(
    title    = "RD Plot: Income Jump at the Geographic Boundary",
    subtitle = "Each dot = mean income within a 0.5° longitude bin; ribbons = 95% CI",
    x        = "Distance from Policy Boundary (degrees longitude)",
    y        = "Average Annual Income ($)"
  ) +
  ggplot2::theme_minimal(base_size = 12) +
  ggplot2::theme(
    legend.position  = "bottom",
    plot.title       = ggplot2::element_text(face = "bold", size = 14)
  )
```

Geographic RD designs arise naturally in studies of redistricting, cross-border policy differences, environmental regulations, and school attendance zones. The four visualizations above — boundary map, outcome choropleth, bandwidth zoom, and RD jump plot — are the standard diagnostic figures for a geographic RD paper.

---

# 16. Regression Kink Design

The **Regression Kink Design** (RKD), introduced by @card2015inference, identifies causal effects from a **change in the slope** (kink) of a policy function at a known threshold, rather than a change in the level (discontinuity). This is useful when treatment intensity changes at the cutoff but treatment itself does not switch on or off.

## 16.1 Theory

In a standard RD, we look for a jump in $E[Y | X = x]$ at $x = c$. In an RKD, the policy function $B(x)$ is continuous at $c$ but has a kink --- its first derivative changes:

$$
\tau_{RKD} = \frac{\lim_{x \downarrow c} \frac{dE[Y|X=x]}{dx} - \lim_{x \uparrow c} \frac{dE[Y|X=x]}{dx}}{\lim_{x \downarrow c} \frac{dB(x)}{dx} - \lim_{x \uparrow c} \frac{dB(x)}{dx}}
$$

The numerator is the change in the slope of the outcome and the denominator is the change in the slope of the treatment-determining function.

## 16.2 Estimation with `rdrobust`

The `rdrobust` package supports kink estimation by setting `deriv = 1`, which estimates the change in the **first derivative** at the cutoff rather than the change in level:

```{r rkd-example, eval = requireNamespace("rdrobust", quietly = TRUE)}
library(rdrobust)

set.seed(42)
n <- 4000

# Running variable
x_rkd <- runif(n, -2, 2)

# Policy function with a kink at 0: slope changes from 0.5 to 1.5
B_rkd <- ifelse(x_rkd < 0, 0.5 * x_rkd, 1.5 * x_rkd)

# Outcome: responds to the policy function with coefficient 2
y_rkd <- 2 * B_rkd + 0.3 * x_rkd^2 + rnorm(n, sd = 0.8)

# Estimate the kink: deriv = 1 targets the slope change
rkd_fit <- rdrobust(y = y_rkd, x = x_rkd, deriv = 1)
summary(rkd_fit)
```

## 16.3 Visualizing the Kink

```{r rkd-plot, eval = requireNamespace("rdrobust", quietly = TRUE), fig.width = 7, fig.height = 5}
dat_rkd <- data.frame(x = x_rkd, y = y_rkd, B = B_rkd)

ggplot(dat_rkd, aes(x = x, y = y)) +
  geom_point(alpha = 0.15, size = 0.6, color = "grey50") +
  geom_smooth(data = subset(dat_rkd, x < 0),
              method = "lm", formula = y ~ poly(x, 2),
              se = FALSE, color = "steelblue", linewidth = 1) +
  geom_smooth(data = subset(dat_rkd, x >= 0),
              method = "lm", formula = y ~ poly(x, 2),
              se = FALSE, color = "firebrick", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    title = "Regression Kink Design: Slope Change at the Cutoff",
    x     = "Running Variable",
    y     = "Outcome"
  ) +
  causalverse::ama_theme()
```

The key identifying assumption in an RKD is that all other determinants of the outcome have a smooth conditional density at the kink point, analogous to the continuity assumption in a standard RD.

---

# 17. RD with Discrete Running Variable

When the running variable takes on only a finite number of values (e.g., integer test scores, age in years, vote counts), standard RD methods face challenges because the continuity framework breaks down. @lee2008randomized discuss how discreteness introduces specification uncertainty and propose methods to account for it.

## 17.1 Key Challenges

1. **Mass points**: Multiple observations share the same value of the running variable, creating atoms in the distribution.
2. **Specification sensitivity**: With few unique values near the cutoff, polynomial choice and bandwidth selection become fragile.
3. **Asymptotic invalidity**: Standard local polynomial estimators rely on a continuous running variable; with discrete scores, bias may not vanish asymptotically.

## 17.2 Binomial Test Approach

One approach is to treat observations at each value of the running variable as a group and use a **binomial test** to assess whether treatment assignment at the cutoff is consistent with local random assignment:

```{r rd-discrete, fig.width = 7, fig.height = 5}
set.seed(42)
n <- 3000

# Discrete running variable: integer scores from -10 to 10
x_disc <- sample(-10:10, n, replace = TRUE)

# Sharp treatment at cutoff = 0
treated_disc <- as.integer(x_disc >= 0)

# Treatment effect = 2, with some noise
y_disc <- 0.4 * x_disc + 2 * treated_disc + rnorm(n, sd = 1.5)

dat_disc <- data.frame(x = x_disc, y = y_disc, treated = treated_disc)

# Compute cell means at each score value
cell_means <- dat_disc |>
  dplyr::group_by(x) |>
  dplyr::summarise(
    y_mean = mean(y),
    y_se   = sd(y) / sqrt(dplyr::n()),
    n_obs  = dplyr::n(),
    .groups = "drop"
  )

ggplot(cell_means, aes(x = x, y = y_mean, color = factor(as.integer(x >= 0)))) +
  geom_point(size = 3) +
  geom_errorbar(aes(ymin = y_mean - 1.96 * y_se,
                    ymax = y_mean + 1.96 * y_se),
                width = 0.3) +
  geom_vline(xintercept = -0.5, linetype = "dashed") +
  scale_color_manual(values = c("0" = "steelblue", "1" = "firebrick"),
                     labels = c("Control", "Treated")) +
  labs(
    title = "RD with Discrete Running Variable: Cell Means and 95% CIs",
    x     = "Running Variable (Integer Score)",
    y     = "Mean Outcome",
    color = "Group"
  ) +
  causalverse::ama_theme()
```

## 17.3 Local Linear Estimation with Clustering

When using `rdrobust` with a discrete running variable, clustering standard errors by the running variable accounts for the mass-point structure:

```{r rd-discrete-rdrobust, eval = requireNamespace("rdrobust", quietly = TRUE)}
# With discrete running variable, use masspoints = "adjust" to handle mass points
rd_disc_fit <- rdrobust::rdrobust(
  y           = dat_disc$y,
  x           = dat_disc$x,
  masspoints  = "adjust"
)
summary(rd_disc_fit)

# Also estimate with jittered running variable as robustness check
set.seed(99)
x_jitter <- dat_disc$x + runif(nrow(dat_disc), -0.49, 0.49)
rd_jitter <- rdrobust::rdrobust(y = dat_disc$y, x = x_jitter)
cat("\nJittered running variable estimate:", round(rd_jitter$coef[3], 3),
    "(SE:", round(rd_jitter$se[3], 3), ")\n")
```

Researchers working with discrete running variables should report results across multiple bandwidths and polynomial orders to demonstrate robustness, and should cluster standard errors at the level of the running variable.

---

# 18. Practical Guidelines for Applied Researchers

This section provides a comprehensive checklist for conducting and reporting RD analyses.

## 14.1 Design Phase

1. **Document the assignment rule clearly.** Explain what the running variable is, where the cutoff comes from, and why the cutoff is relevant. Is it a policy rule, an administrative threshold, or a test score?

2. **Determine Sharp vs. Fuzzy.** Inspect compliance: does everyone above the cutoff receive treatment? If not, the design is fuzzy and requires IV estimation.

3. **Conduct a power analysis.** Use `rdpower` to determine whether your sample size is sufficient to detect a meaningful effect. RD designs require large samples because only observations near the cutoff contribute to estimation.

## 14.2 Analysis Checklist

**Step 1: Visualize**

- Plot the raw data (scatter plot of outcome vs. running variable)
- Create a binned scatter plot with confidence intervals
- Overlay local polynomial fits on each side of the cutoff

**Step 2: Test Assumptions**

- Run the density test (`rddensity`) to check for manipulation
- Test covariate balance at the cutoff (each covariate as outcome)
- Plot the density of the running variable (histogram)

**Step 3: Estimate**

- Use `rdrobust` with default settings (local linear, triangular kernel, MSE-optimal bandwidth)
- Report the robust confidence interval (not the conventional one)
- Include covariates for precision (but verify the point estimate does not change)

**Step 4: Robustness**

- Vary the bandwidth (0.5x to 2x of the MSE-optimal)
- Try different polynomial orders (p = 0, 1, 2)
- Try different kernels (triangular, epanechnikov, uniform)
- Conduct placebo cutoff tests (estimate the RD at fake cutoffs where no effect should exist)
- If manipulation is a concern, compute bounds using `rdbounds`

**Step 5: Report**

- Report the RD estimate, robust CI, bandwidth, effective sample sizes on each side
- Show the main RD plot
- Report all robustness checks in an appendix or supplementary table
- Discuss threats to validity (manipulation, sorting, attrition)

## 14.3 Common Pitfalls

1. **Using global polynomial regressions.** Do not fit a single high-order polynomial to the entire sample. This is known to produce misleading results (Gelman and Imbens, 2019). Always use local methods.

2. **Choosing bandwidths by eyeballing.** Use data-driven bandwidth selection (`rdrobust` does this automatically). Manual bandwidth choice invites specification searching.

3. **Ignoring the bias-variance tradeoff.** The MSE-optimal bandwidth balances bias and variance for point estimation, but the resulting conventional CI has poor coverage. Always use the robust CI from `rdrobust`.

4. **Reporting only one specification.** Show that results are robust to bandwidth, polynomial order, kernel choice, and covariate inclusion.

5. **Forgetting the first stage in Fuzzy RD.** Always verify and report the first-stage relationship. A weak first stage leads to unreliable IV estimates.

## 14.4 Placebo Cutoff Tests

A useful falsification exercise is to estimate the RD at "fake" cutoffs where no treatment effect should exist. Significant effects at placebo cutoffs suggest that the observed discontinuity at the true cutoff may be spurious.

```{r placebo-cutoffs, fig.width=7, fig.height=4}
# Test at placebo cutoffs (no true effect at these points)
placebo_cutoffs <- c(-0.5, -0.25, 0.25, 0.5)

placebo_results <- lapply(placebo_cutoffs, function(c_val) {
  sub <- rd_data |>
    filter(
      # Only use observations on the correct side of the real cutoff
      # to avoid contamination
      (c_val < 0 & x < 0) | (c_val > 0 & x >= 0)
    ) |>
    mutate(
      placebo_treat = as.integer(x >= c_val),
      x_centered    = x - c_val,
      x_pt          = x_centered * placebo_treat
    )

  est <- feols(y ~ placebo_treat + x_centered + x_pt, data = sub)

  tibble(
    cutoff   = c_val,
    estimate = coef(est)["placebo_treat"],
    se       = se(est)["placebo_treat"],
    p_value  = pvalue(est)["placebo_treat"]
  )
}) |>
  bind_rows()

# Add the true cutoff result
true_est <- feols(y ~ treated + x + x_treated, data = rd_data)
true_result <- tibble(
  cutoff   = 0,
  estimate = coef(true_est)["treated"],
  se       = se(true_est)["treated"],
  p_value  = pvalue(true_est)["treated"]
)

all_cutoffs <- bind_rows(placebo_results, true_result) |>
  mutate(type = ifelse(cutoff == 0, "True Cutoff", "Placebo"))

ggplot(all_cutoffs, aes(x = factor(cutoff), y = estimate, color = type)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
    width = 0.15
  ) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "grey60") +
  scale_color_manual(values = c("Placebo" = "steelblue", "True Cutoff" = "firebrick")) +
  labs(
    x     = "Cutoff Value",
    y     = "Estimated Effect",
    color = "Type",
    title = "Placebo Cutoff Tests"
  ) +
  causalverse::ama_theme()
```

The true cutoff (c = 0) shows a large, significant effect. The placebo cutoffs show estimates near zero, as expected.

## 14.5 Donut Hole Tests

Another robustness check is the "donut hole" test: exclude observations very close to the cutoff and re-estimate. If the result is driven by a few observations right at the cutoff (perhaps manipulators), the estimate will change substantially when they are removed.

```{r donut-hole, fig.width=7, fig.height=4}
donut_sizes <- c(0, 0.02, 0.05, 0.10, 0.15)

donut_results <- lapply(donut_sizes, function(d) {
  sub <- rd_data |> filter(abs(x) > d)
  est <- feols(y ~ treated + x + x_treated, data = sub)
  tibble(
    donut     = d,
    n_obs     = nrow(sub),
    estimate  = coef(est)["treated"],
    se        = se(est)["treated"]
  )
}) |>
  bind_rows()

ggplot(donut_results, aes(x = factor(donut), y = estimate)) +
  geom_point(size = 3) +
  geom_errorbar(
    aes(ymin = estimate - 1.96 * se, ymax = estimate + 1.96 * se),
    width = 0.1
  ) +
  geom_hline(yintercept = 3, linetype = "dashed", color = "red") +
  labs(
    x     = "Donut Hole Radius (excluded |x| < d)",
    y     = "Estimated Treatment Effect",
    title = "Donut Hole Robustness Check"
  ) +
  causalverse::ama_theme()
```

---

# 19. Comprehensive Visualization Suite for RD

## 19.1 Publication-Ready RD Plot with Binned Scatter

A binned scatter plot is the standard visualization in top journals. It groups observations into equal-sized bins on either side of the cutoff and plots the mean of the outcome within each bin, overlaid with a smooth polynomial fit.

```{r rd-binscatter, message=FALSE, fig.width=8, fig.height=5}
library(rdrobust)

# Simulate clean RD data
set.seed(2024)
n_rd <- 1500
X_rd <- runif(n_rd, -1, 1)
treat_rd <- as.integer(X_rd >= 0)

# True effect = 1.5, nonlinear baseline
Y_rd <- 2 + 3 * X_rd - 2 * X_rd^2 + 1.5 * treat_rd + rnorm(n_rd, 0, 0.5)

# Binned scatter with rdplot
rdplot_out <- rdplot(Y_rd, X_rd, c = 0, nbins = c(20, 20),
                     title    = "RD: Binned Scatter with Polynomial Fit",
                     x.label  = "Running Variable (centered at cutoff)",
                     y.label  = "Outcome")
```

## 19.2 Full RD Assumption Test Battery

Use `causalverse::rd_assumption_tests()` to run all standard robustness checks in one call:

```{r rd-assumption-battery, message=FALSE, eval=TRUE}
df_rd_test <- data.frame(Y = Y_rd, X = X_rd)
rd_tests   <- causalverse::rd_assumption_tests(
    data        = df_rd_test,
    outcome     = "Y",
    running_var = "X",
    cutoff      = 0,
    bw_seq      = c(0.5, 0.75, 1, 1.25, 1.5),
    poly_orders = 1:3,
    verbose     = FALSE
)
cat(rd_tests$summary_report)
```

## 19.3 Bandwidth Sensitivity Visualization

```{r rd-bw-viz, message=FALSE, fig.width=7, fig.height=5}
# Run sensitivity using causalverse built-in
bw_sens <- causalverse::rd_bandwidth_sensitivity(
    data   = df_rd_test,
    y      = "Y",
    x      = "X",
    c      = 0
)
bw_sens$plot + causalverse::ama_theme()
```

## 19.4 Placebo Cutoff Visualization

```{r rd-placebo-viz, message=FALSE, fig.width=7, fig.height=5}
pc_results <- causalverse::rd_placebo_cutoffs(
    data        = df_rd_test,
    y           = "Y",
    x           = "X",
    true_cutoff = 0
)
pc_results$plot + causalverse::ama_theme()
```

## 19.5 Density Continuity (McCrary) Test Visualization

```{r rd-density-viz, message=FALSE, fig.width=7, fig.height=4}
if (requireNamespace("rddensity", quietly = TRUE)) {
    dens_test <- rddensity::rddensity(X = X_rd, c = 0)
    rddensity::rdplotdensity(
        rdd    = dens_test,
        X      = X_rd,
        title  = "McCrary Density Test: No Manipulation",
        xlabel = "Running Variable",
        ylabel = "Density"
    )
} else {
    # Histogram-based density check
    hist(X_rd, breaks = 40, col = "steelblue",
         main = "Running Variable Distribution",
         xlab = "Running Variable")
    abline(v = 0, col = "red", lwd = 2, lty = 2)
}
```

## 19.6 Covariate Balance at the Cutoff

```{r rd-cov-balance, message=FALSE}
# Simulate covariates (should NOT jump at the cutoff)
df_rd_test$age    <- 25 + 10 * X_rd + rnorm(n_rd, 0, 5)
df_rd_test$income <- 40000 + 5000 * X_rd + rnorm(n_rd, 0, 8000)

cov_bal <- causalverse::rd_covariate_balance(
    data       = df_rd_test,
    covariates = c("age", "income"),
    x          = "X",
    c          = 0
)
print(cov_bal)
```

## 19.7 Confidence Interval Comparison: Conventional vs. Robust

A key feature of `rdrobust` is the **robust bias-corrected** (RBC) confidence interval, which is preferred in modern practice. Visualization of the difference:

```{r rd-ci-comparison, message=FALSE, fig.width=7, fig.height=4}
rd_main <- rdrobust(Y_rd, X_rd, c = 0)
rd_sum  <- summary(rd_main)

# Extract both sets of CIs
ci_df <- data.frame(
    type     = c("Conventional", "Bias-Corrected", "Robust (RBC)"),
    estimate = rep(rd_main$coef[1], 3),
    ci_lower = c(rd_main$ci[1, 1], rd_main$ci[2, 1], rd_main$ci[3, 1]),
    ci_upper = c(rd_main$ci[1, 2], rd_main$ci[2, 2], rd_main$ci[3, 2])
)

ggplot2::ggplot(ci_df, ggplot2::aes(x = type, y = estimate)) +
    ggplot2::geom_hline(yintercept = 1.5, linetype = "dashed",
                        color = "green3", linewidth = 0.8) +
    ggplot2::geom_point(size = 4, color = "#2166AC") +
    ggplot2::geom_errorbar(
        ggplot2::aes(ymin = ci_lower, ymax = ci_upper),
        width = 0.25, linewidth = 0.9, color = "#2166AC"
    ) +
    ggplot2::annotate("text", x = 0.6, y = 1.5, label = "True effect = 1.5",
                      color = "green4", size = 3.5, hjust = 0) +
    ggplot2::labs(
        x     = "CI Type",
        y     = "RD Estimate and 95% CI",
        title = "Conventional vs. Bias-Corrected vs. Robust CIs",
        subtitle = "RBC CI (rightmost) is the recommended modern approach"
    ) +
    causalverse::ama_theme()
```

## 19.8 Local Randomization Framework: Balance Tests

Under the local randomization framework (Cattaneo, Frandsen, Titiunik 2015), units near the cutoff are treated as if randomly assigned. Balance tests within increasingly narrow windows around the cutoff assess this assumption.

```{r rd-local-rand, message=FALSE, fig.width=9, fig.height=5}
# Test balance for multiple narrow windows
windows_lr <- c(0.05, 0.10, 0.15, 0.20, 0.25, 0.30)

balance_results <- lapply(windows_lr, function(w) {
    idx <- abs(X_rd) <= w
    dat <- df_rd_test[idx, ]
    D   <- as.integer(dat$X >= 0)

    # t-test for age balance
    age_p <- tryCatch(
        t.test(dat$age ~ D)$p.value,
        error = function(e) NA
    )
    inc_p <- tryCatch(
        t.test(dat$income ~ D)$p.value,
        error = function(e) NA
    )
    data.frame(window = w, n = sum(idx), age_p = age_p, income_p = inc_p)
})
bal_df <- do.call(rbind, balance_results)

ggplot2::ggplot(bal_df, ggplot2::aes(x = window)) +
    ggplot2::geom_line(ggplot2::aes(y = age_p, color = "Age"),
                       linewidth = 1) +
    ggplot2::geom_line(ggplot2::aes(y = income_p, color = "Income"),
                       linewidth = 1, linetype = "dashed") +
    ggplot2::geom_point(ggplot2::aes(y = age_p, color = "Age"), size = 3) +
    ggplot2::geom_point(ggplot2::aes(y = income_p, color = "Income"), size = 3) +
    ggplot2::geom_hline(yintercept = 0.05, linetype = "dotted", color = "red") +
    ggplot2::scale_color_manual(values = c("Age" = "#2166AC",
                                           "Income" = "#D73027")) +
    ggplot2::labs(
        x     = "Window Half-Width around Cutoff",
        y     = "p-value (covariate balance t-test)",
        title = "Local Randomization: Covariate Balance by Window Width",
        subtitle = "p > 0.05 (above red line) = covariates balanced",
        color = "Covariate"
    ) +
    ggplot2::scale_y_continuous(limits = c(0, 1)) +
    causalverse::ama_theme() +
    ggplot2::theme(plot.margin = ggplot2::margin(0.5, 0.5, 0.5, 1.5, "cm"))
```

## 19.9 Fuzzy RD: First Stage and Reduced Form Visualization

For fuzzy RD designs, always plot both the first stage (treatment take-up at the cutoff) and the reduced form (ITT effect), alongside the IV estimate (LATE).

```{r rd-fuzzy-viz, message=FALSE, fig.width=9, fig.height=4}
# Simulate fuzzy RD
set.seed(77)
n_f   <- 1000
X_f   <- runif(n_f, -1, 1)
# Compliance: 85% comply when eligible (X >= 0), 5% take up when ineligible
D_f   <- rbinom(n_f, 1,
                ifelse(X_f >= 0, 0.85, 0.05))
Y_f   <- 1 + 2 * X_f + 1.8 * D_f + rnorm(n_f, 0, 0.6)

# Bin by X
bins_f <- cut(X_f, breaks = seq(-1, 1, by = 0.1),
              include.lowest = TRUE)
df_f   <- data.frame(X = X_f, D = D_f, Y = Y_f, bin = bins_f)
agg_f  <- aggregate(cbind(D, Y, X) ~ bin, df_f, mean)

p_fs <- ggplot2::ggplot(agg_f, ggplot2::aes(x = X, y = D)) +
    ggplot2::geom_point(color = "#2166AC") +
    ggplot2::geom_smooth(data = subset(agg_f, X < 0),
                         method = "lm", se = FALSE, color = "#2166AC") +
    ggplot2::geom_smooth(data = subset(agg_f, X >= 0),
                         method = "lm", se = FALSE, color = "#2166AC") +
    ggplot2::geom_vline(xintercept = 0, linetype = "dashed") +
    ggplot2::labs(title = "First Stage", x = "Running Variable", y = "P(Treated)") +
    causalverse::ama_theme()

p_rf <- ggplot2::ggplot(agg_f, ggplot2::aes(x = X, y = Y)) +
    ggplot2::geom_point(color = "#D73027") +
    ggplot2::geom_smooth(data = subset(agg_f, X < 0),
                         method = "lm", se = FALSE, color = "#D73027") +
    ggplot2::geom_smooth(data = subset(agg_f, X >= 0),
                         method = "lm", se = FALSE, color = "#D73027") +
    ggplot2::geom_vline(xintercept = 0, linetype = "dashed") +
    ggplot2::labs(title = "Reduced Form (ITT)", x = "Running Variable", y = "Outcome") +
    causalverse::ama_theme()

gridExtra::grid.arrange(p_fs, p_rf, ncol = 2)
```

## 19.10 Complete RD Publication Pipeline

The following template assembles the full set of tables and figures required for a top-journal RD paper:

```{r rd-full-pipeline, message=FALSE, eval=TRUE}
cat("============================================================\n")
cat("COMPLETE RD ANALYSIS PIPELINE\n")
cat("============================================================\n\n")

# Step 1: Main estimate
cat("--- Step 1: Main RD Estimate ---\n")
main_est <- rdrobust(Y_rd, X_rd, c = 0)
cat("RD estimate (RBC): ", round(main_est$coef[3], 4), "\n")
cat("SE (robust):        ", round(main_est$se[3],   4), "\n")
cat("95% CI:            [", round(main_est$ci[3,1], 4),
    ",", round(main_est$ci[3,2], 4), "]\n")
cat("p-value (robust):  ", round(main_est$pv[3], 4), "\n\n")

# Step 2: Bandwidth
cat("--- Step 2: Optimal Bandwidth ---\n")
cat("MSE-optimal h:     ", round(main_est$bws[1,1], 4), "\n")
cat("N within h:        ", main_est$N_h[1], "(left),",
    main_est$N_h[2], "(right)\n\n")

# Step 3: Quick robustness summary
cat("--- Step 3: Robustness Summary ---\n")
cat("Density test:       Run rddensity::rddensity() - should be p > 0.05\n")
cat("Covariate balance:  Checked above - all p > 0.05\n")
cat("Bandwidth sensitiv: Estimate stable across 0.5x-1.5x bandwidth\n")
cat("Placebo cutoffs:    No significant effects at placebo cutoffs\n\n")

cat("=== PIPELINE COMPLETE ===\n")
```

---

# 20. Comprehensive References

## Foundational References

- Hahn, J., Todd, P., and Van der Klaauw, W. (2001). "Identification and Estimation of Treatment Effects with a Regression-Discontinuity Design." *Econometrica*, 69(1), 201-209.

- Imbens, G. W. and Lemieux, T. (2008). "Regression Discontinuity Designs: A Guide to Practice." *Journal of Econometrics*, 142(2), 615-635.

- Lee, D. S. and Lemieux, T. (2010). "Regression Discontinuity Designs in Economics." *Journal of Economic Literature*, 48(2), 281-355.

## Estimation and Inference

- Calonico, S., Cattaneo, M. D., and Titiunik, R. (2014). "Robust Nonparametric Confidence Intervals for Regression-Discontinuity Designs." *Econometrica*, 82(6), 2295-2326.

- Calonico, S., Cattaneo, M. D., and Farrell, M. H. (2020). "Optimal Bandwidth Choice for Robust Bias-Corrected Inference in Regression Discontinuity Designs." *Econometrics Journal*, 23(2), 192-210.

- Imbens, G. W. and Kalyanaraman, K. (2012). "Optimal Bandwidth Choice for the Regression Discontinuity Estimator." *Review of Economic Studies*, 79(3), 933-959.

## Manipulation Testing

- McCrary, J. (2008). "Manipulation of the Running Variable in the Regression Discontinuity Design: A Density Test." *Journal of Econometrics*, 142(2), 698-714.

- Cattaneo, M. D., Jansson, M., and Ma, X. (2020). "Simple Local Polynomial Density Estimators." *Journal of the American Statistical Association*, 115(531), 1449-1455.

## Power Analysis

- Cattaneo, M. D., Titiunik, R., and Vazquez-Bare, G. (2019). "Power Calculations for Regression-Discontinuity Designs." *Stata Journal*, 19(1), 210-245.

## Local Randomization

- Cattaneo, M. D., Frandsen, B. R., and Titiunik, R. (2015). "Randomization Inference in the Regression Discontinuity Design: An Application to Party Advantages in the U.S. Senate." *Journal of Causal Inference*, 3(1), 1-24.

## Multiple Cutoffs and Scores

- Cattaneo, M. D., Keele, L., Titiunik, R., and Vazquez-Bare, G. (2016). "Interpreting Regression Discontinuity Designs with Multiple Cutoffs." *Journal of Politics*, 78(4), 1229-1248.

- Cattaneo, M. D., Keele, L., Titiunik, R., and Vazquez-Bare, G. (2021). "Extrapolating Treatment Effects in Multi-Cutoff Regression Discontinuity Designs." *Journal of the American Statistical Association*, 116(536), 1941-1952.

## Bounds Under Manipulation

- Gerard, F., Rokkanen, M., and Rothe, C. (2020). "Bounds on Treatment Effects in Regression Discontinuity Designs with a Manipulated Running Variable." *Quantitative Economics*, 11(3), 839-870.

## Lee Bounds and Selection

- Lee, D. S. (2009). "Training, Wages, and Sample Selection: Estimating Sharp Bounds on Treatment Effects." *Review of Economic Studies*, 76(3), 1071-1102.

## Polynomial Warnings

- Gelman, A. and Imbens, G. (2019). "Why High-Order Polynomials Should Not Be Used in Regression Discontinuity Designs." *Journal of Business and Economic Statistics*, 37(3), 447-456.

## Software and Implementation

- Calonico, S., Cattaneo, M. D., and Titiunik, R. (2015). "rdrobust: An R Package for Robust Nonparametric Inference in Regression-Discontinuity Designs." *R Journal*, 7(1), 38-51.

- Cattaneo, M. D., Idrobo, N., and Titiunik, R. (2020). *A Practical Introduction to Regression Discontinuity Designs: Foundations*. Cambridge Elements.

- Cattaneo, M. D., Idrobo, N., and Titiunik, R. (2024). *A Practical Introduction to Regression Discontinuity Designs: Extensions*. Cambridge Elements.
