---
title: "0. Introduction to CausalVerse"
author: "Mike Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{0. Introduction to CausalVerse}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(causalverse)
```

## **Introduction to the `causalverse` Package**

Welcome to the **`causalverse`** package - a dedicated toolkit tailored for researchers embarking on causal inference analyses. Our primary mission is to simplify and enhance the research process by offering robust tools specifically designed for various causal inference methodologies.

### **Vignettes Overview**

To enable a comprehensive understanding of each causal inference method, **`causalverse`** boasts a series of in-depth vignettes. Each vignette offers a blend of theoretical background and hands-on demonstrations, ensuring you have both the knowledge and skills to implement each method effectively.

Here's a snapshot of our method-specific vignettes:

1.  **Regression Discontinuity**: The Regression Discontinuity vignette delves into the nuances of the Regression Discontinuity approach. From its foundational principles to its practical implementation in **`causalverse`**, this guide offers a thorough exploration.

2.  **Difference in Differences**: Navigate to the [Difference-in-Differences vignette](https://mikenguyen13.github.io/causalverse/articles/did.html) for a deep dive into the Difference in Differences methodology. Grasp the core concepts, witness its application, and explore use-cases.

3.  **Synthetic Controls (SC)**: The **`sc`** vignette demystifies the Synthetic Control method, providing insights into its algorithm, advantages, limitations, and usage within our package.

4.  **Instrumental Variables (IV)**: Turn to the **`iv`** vignette for a comprehensive look at the Instrumental Variables approach, where you'll learn about its theoretical underpinnings and practical application in **`causalverse`**.

5.  **Event Studies (EV)**: The **`ev`** vignette offers a deep dive into Event Studies, ensuring you're equipped to harness its full potential for causal analysis.

6.  **Randomized Control Trials (RCT)**: Although **`causalverse`** primarily focuses on quasi-experimental methods using observational data, we recognize the importance of experimental methods in causal research. The **`rct`** vignette presents a detailed overview of Randomized Control Trials, the gold standard of experimental research.

### **A Note on Methodologies**

The majority of methods covered in **`causalverse`** pertain to quasi-experimental approaches, typically applied to observational data. These methods are instrumental in scenarios where running randomized experiments might be infeasible, unethical, or costly. However, we also touch upon experimental methods, specifically RCTs, recognizing their unparalleled significance in establishing causal relationships.

### **Getting Started**

As you embark on your journey with **`causalverse`**, we recommend starting with this introductory vignette to familiarize yourself with the package's architecture and offerings. Then, delve into the method-specific vignettes that align with your research objectives or peruse them all for a holistic understanding.

I hope this provides a clearer and professional introduction for your vignette. If you have any additional inputs or refinements, please let me know!
