---
title: "10. Heterogeneous Treatment Effects & Causal ML"
author: "Mike Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{10. Heterogeneous Treatment Effects & Causal ML}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE, warning = FALSE, message = FALSE)
```

```{r setup, message=FALSE, warning=FALSE}
library(causalverse)
library(ggplot2)
library(dplyr)
```

---

## 1. Introduction

### 1.1 Why Heterogeneous Treatment Effects Matter

Most causal inference methods report a single summary statistic — the Average Treatment Effect (ATE) — which tells us the average impact of an intervention across the study population. While the ATE is a valuable and interpretable quantity, it can obscure substantial variation in how individuals respond to treatment. When treatment effects differ across units, reporting only the ATE may lead to:

- **Misallocation of resources**: Treating individuals who do not benefit or even experience harm.
- **Missed targeting opportunities**: Failing to identify sub-populations where treatment is highly effective.
- **Policy errors**: Designing one-size-fits-all policies that ignore individual heterogeneity.
- **Replication failures**: An ATE that is statistically significant in one context may not replicate if the population composition changes.

Heterogeneous Treatment Effect (HTE) analysis asks: **for whom does the treatment work, and by how much?** This is the central question in precision medicine, personalized policy, targeted marketing, and individualized education.

Modern causal machine learning methods — including causal forests, meta-learners, and double/debiased machine learning — make it possible to estimate HTE at the individual level, validate those estimates, and translate them into actionable targeting rules, all while maintaining rigorous causal guarantees.

### 1.2 Key Estimands: CATE, ATE, ATT

Let $Y_i(1)$ and $Y_i(0)$ denote the potential outcomes for unit $i$ under treatment and control, respectively. The treatment indicator is $W_i \in \{0, 1\}$, and $X_i$ is a vector of pre-treatment covariates.

**Individual Treatment Effect (ITE)**:
$$\tau_i = Y_i(1) - Y_i(0)$$
This is never observed due to the fundamental problem of causal inference.

**Conditional Average Treatment Effect (CATE)**:
$$\tau(x) = \mathbb{E}[Y_i(1) - Y_i(0) \mid X_i = x]$$
The CATE is the population-level treatment effect for units with covariate profile $x$. This is the primary estimand of HTE analysis.

**Average Treatment Effect (ATE)**:
$$\tau_{ATE} = \mathbb{E}[\tau(X_i)] = \mathbb{E}[Y_i(1) - Y_i(0)]$$

**Average Treatment Effect on the Treated (ATT)**:
$$\tau_{ATT} = \mathbb{E}[\tau(X_i) \mid W_i = 1]$$

The CATE generalizes the ATE: $\tau_{ATE} = \mathbb{E}[\tau(X_i)]$ is the expectation of the CATE over the marginal distribution of $X$.

### 1.3 Identification Assumptions

To identify the CATE from observational data, we require:

1. **Unconfoundedness (Conditional Independence)**: $(Y(0), Y(1)) \perp W \mid X$
   No unmeasured confounders conditional on $X$.

2. **Overlap (Positivity)**: $0 < \mathbb{P}(W = 1 \mid X = x) < 1$ for all $x$
   Every covariate profile has a positive probability of both treatment and control.

3. **SUTVA**: Stable Unit Treatment Value Assumption — no interference and no hidden treatment versions.

Under these assumptions, the CATE is identified:
$$\tau(x) = \mathbb{E}[Y \mid W = 1, X = x] - \mathbb{E}[Y \mid W = 0, X = x]$$

In randomized experiments, unconfoundedness holds by design. In observational studies, it must be justified by domain knowledge.

### 1.4 Overview of Methods Covered

This vignette covers a comprehensive toolkit for HTE estimation:

| Method | Key Reference | Package |
|--------|---------------|---------|
| S-Learner | Künzel et al. (2019) | `xgboost`, `lm` |
| T-Learner | Künzel et al. (2019) | Any ML |
| X-Learner | Künzel et al. (2019) | `grf` |
| R-Learner / DML | Robinson (1988), Nie & Wager (2021) | `glmnet`, `DoubleML` |
| Causal Forest | Wager & Athey (2018), Athey et al. (2019) | `grf` |
| BLP / GATES | Chernozhukov et al. (2020) | `causalverse` |
| DR-Learner / AIPW | Kennedy (2020) | `causalverse` |
| Qini / AUUC | Radcliffe (2007) | `causalverse` |
| DML with interactions | Chernozhukov et al. (2018) | `glmnet`, `DoubleML` |

---

## 2. Setup & Data

### 2.1 Simulated Data with Known Heterogeneity

We begin with a simulation where the true CATE is known, enabling us to evaluate estimator performance.

```{r sim-data}
set.seed(42)
n <- 2000
p <- 10

# Covariates: standard normal
X <- matrix(rnorm(n * p), n, p)
colnames(X) <- paste0("X", 1:p)

# Treatment: randomized with probability 0.5
W <- rbinom(n, 1, 0.5)

# True CATE: linear in X1 (heterogeneity driven by X1)
tau_true <- X[, 1]

# Outcome: tau * W + noise
Y <- tau_true * W + 0.5 * X[, 2] + rnorm(n)

# Compile into data frame
df <- data.frame(X, W = W, Y = Y, tau_true = tau_true)

cat("ATE (true):", mean(tau_true), "\n")
cat("SD of tau (true):", round(sd(tau_true), 3), "\n")
cat("Fraction positive tau:", round(mean(tau_true > 0), 3), "\n")
cat("n =", n, "| p =", p, "| Treatment fraction:", mean(W), "\n")
```

### 2.2 The Lalonde Dataset

We also use the classic Lalonde (1986) dataset from the `MatchIt` package, which evaluates a job training program on earnings. This dataset illustrates HTE in a real-world observational setting.

```{r lalonde-data}
if (requireNamespace("MatchIt", quietly = TRUE)) {
  data(lalonde, package = "MatchIt")
  cat("Lalonde dataset: n =", nrow(lalonde), "\n")
  cat("Outcome: re78 (1978 earnings)\n")
  cat("Treatment: treat (job training, 1 = treated)\n")
  cat("Treatment fraction:", round(mean(lalonde$treat), 3), "\n")
  cat("Mean re78 (treated):", round(mean(lalonde$re78[lalonde$treat == 1]), 1), "\n")
  cat("Mean re78 (control):", round(mean(lalonde$re78[lalonde$treat == 0]), 1), "\n")
}
```

### 2.3 Pre-Treatment Covariate Balance

Before estimating treatment effects, always check covariate balance.

```{r balance-check}
# Balance table for simulated data
treated_idx   <- which(W == 1)
control_idx   <- which(W == 0)

balance_df <- data.frame(
  Variable = colnames(X),
  Mean_Treated = round(colMeans(X[treated_idx, ]), 3),
  Mean_Control = round(colMeans(X[control_idx, ]), 3),
  Std_Diff = round(
    (colMeans(X[treated_idx, ]) - colMeans(X[control_idx, ])) /
      sqrt((apply(X[treated_idx, ], 2, var) + apply(X[control_idx, ], 2, var)) / 2),
    3
  )
)
print(balance_df)
```

```{r balance-plot, fig.width=7, fig.height=4}
# Visualize standardized differences
ggplot(balance_df, aes(x = Std_Diff, y = reorder(Variable, Std_Diff))) +
  geom_vline(xintercept = c(-0.1, 0.1), linetype = "dashed", color = "red", alpha = 0.5) +
  geom_vline(xintercept = 0, color = "gray40") +
  geom_point(size = 3, color = "#2166AC") +
  labs(
    x     = "Standardized Mean Difference",
    y     = NULL,
    title = "Pre-Treatment Covariate Balance",
    subtitle = "Dashed lines at ±0.1 (conventional threshold for good balance)"
  ) +
  theme_minimal(base_size = 12)
```

In the randomized simulation, all standardized differences are close to zero — expected under random assignment.

### 2.4 True CATE Distribution

```{r true-cate-dist, fig.width=7, fig.height=4}
ggplot(data.frame(tau = tau_true), aes(x = tau)) +
  geom_histogram(bins = 50, fill = "#4393C3", alpha = 0.7, color = "white") +
  geom_vline(xintercept = mean(tau_true), color = "#D73027", linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x     = "True Individual Treatment Effect (tau)",
    y     = "Count",
    title = "Distribution of True CATE",
    subtitle = sprintf("ATE = %.3f; SD = %.3f; %.0f%% positive",
                       mean(tau_true), sd(tau_true), 100 * mean(tau_true > 0))
  ) +
  theme_minimal(base_size = 12)
```

---

## 3. Meta-Learners

Meta-learners reduce the CATE estimation problem to off-the-shelf regression or classification. They differ in how they use treatment information.

### 3.1 S-Learner (Single Learner)

The S-Learner fits a single model $\hat{\mu}(x, w)$ with the treatment indicator as just another feature:

$$\hat{\tau}(x) = \hat{\mu}(x, 1) - \hat{\mu}(x, 0)$$

**Pros**: Simple, uses all data. **Cons**: May underfit treatment heterogeneity if the learner shrinks the treatment coefficient toward zero (e.g., lasso).

```{r s-learner}
# S-Learner with linear model
df_train <- data.frame(X, W = W, Y = Y)

# Fit model with W as a feature (and all interactions for flexibility)
s_fit <- lm(Y ~ . * W, data = df_train[, c(paste0("X", 1:p), "W", "Y")])

# Predict potential outcomes
df1 <- df_train; df1$W <- 1
df0 <- df_train; df0$W <- 0

tau_s <- predict(s_fit, newdata = df1) - predict(s_fit, newdata = df0)

cat("S-Learner (linear with interactions):\n")
cat("  Mean CATE:", round(mean(tau_s), 4), "\n")
cat("  SD CATE:  ", round(sd(tau_s), 4), "\n")
cat("  RMSE vs truth:", round(sqrt(mean((tau_s - tau_true)^2)), 4), "\n")
```

```{r s-learner-xgb, eval=requireNamespace("xgboost", quietly=TRUE)}
library(xgboost)

# S-Learner with XGBoost
dtrain <- xgb.DMatrix(
  data  = cbind(X, W = W),
  label = Y
)
xgb_params <- list(
  objective        = "reg:squarederror",
  max_depth        = 4,
  eta              = 0.05,
  subsample        = 0.8,
  colsample_bytree = 0.8,
  nthread          = 1
)
set.seed(42)
xgb_fit <- xgb.train(
  params  = xgb_params,
  data    = dtrain,
  nrounds = 200,
  verbose = 0
)

# Predict with W=1 and W=0
d1_xgb <- xgb.DMatrix(data = cbind(X, W = 1))
d0_xgb <- xgb.DMatrix(data = cbind(X, W = 0))
tau_s_xgb <- predict(xgb_fit, d1_xgb) - predict(xgb_fit, d0_xgb)

cat("S-Learner (XGBoost):\n")
cat("  Mean CATE:", round(mean(tau_s_xgb), 4), "\n")
cat("  SD CATE:  ", round(sd(tau_s_xgb), 4), "\n")
cat("  RMSE vs truth:", round(sqrt(mean((tau_s_xgb - tau_true)^2)), 4), "\n")
```

### 3.2 T-Learner (Two-Learner)

The T-Learner fits separate outcome models for treated and control units:

$$\hat{\mu}_1(x) = \mathbb{E}[Y \mid X = x, W = 1], \quad \hat{\mu}_0(x) = \mathbb{E}[Y \mid X = x, W = 0]$$
$$\hat{\tau}(x) = \hat{\mu}_1(x) - \hat{\mu}_0(x)$$

**Pros**: No regularization bias on treatment. **Cons**: Can overfit when one arm is small; ignores information sharing across arms.

```{r t-learner}
# T-Learner with linear models
X_df <- as.data.frame(X)

# Treated model
fit1 <- lm(Y ~ ., data = cbind(X_df, Y = Y)[W == 1, ])
# Control model
fit0 <- lm(Y ~ ., data = cbind(X_df, Y = Y)[W == 0, ])

# Predict on full data
tau_t <- predict(fit1, newdata = X_df) - predict(fit0, newdata = X_df)

cat("T-Learner (linear):\n")
cat("  Mean CATE:", round(mean(tau_t), 4), "\n")
cat("  SD CATE:  ", round(sd(tau_t), 4), "\n")
cat("  RMSE vs truth:", round(sqrt(mean((tau_t - tau_true)^2)), 4), "\n")
```

```{r t-learner-xgb, eval=requireNamespace("xgboost", quietly=TRUE)}
# T-Learner with XGBoost
set.seed(42)

# Treated model
xgb_fit1 <- xgb.train(
  params  = xgb_params,
  data    = xgb.DMatrix(data = X[W == 1, ], label = Y[W == 1]),
  nrounds = 200,
  verbose = 0
)
# Control model
xgb_fit0 <- xgb.train(
  params  = xgb_params,
  data    = xgb.DMatrix(data = X[W == 0, ], label = Y[W == 0]),
  nrounds = 200,
  verbose = 0
)

dX <- xgb.DMatrix(data = X)
tau_t_xgb <- predict(xgb_fit1, dX) - predict(xgb_fit0, dX)

cat("T-Learner (XGBoost):\n")
cat("  Mean CATE:", round(mean(tau_t_xgb), 4), "\n")
cat("  SD CATE:  ", round(sd(tau_t_xgb), 4), "\n")
cat("  RMSE vs truth:", round(sqrt(mean((tau_t_xgb - tau_true)^2)), 4), "\n")
```

### 3.3 X-Learner (Künzel et al. 2019)

The X-Learner is a two-stage procedure designed to improve on the T-Learner when treatment arms are imbalanced:

**Stage 1**: Fit T-Learner models $\hat{\mu}_1$ and $\hat{\mu}_0$.

**Stage 2**: Compute imputed treatment effects:
$$\tilde{D}_i^1 = Y_i(1) - \hat{\mu}_0(X_i) \quad \text{for treated units}$$
$$\tilde{D}_i^0 = \hat{\mu}_1(X_i) - Y_i(0) \quad \text{for control units}$$

Fit regression models for $\tilde{D}^1$ and $\tilde{D}^0$, then combine:
$$\hat{\tau}(x) = g(x) \hat{\tau}^1(x) + (1 - g(x)) \hat{\tau}^0(x)$$
where $g(x)$ is typically the propensity score $\hat{e}(x)$.

```{r x-learner}
# X-Learner implementation using linear models
# Stage 1: T-Learner (reuse fit1, fit0 from above)
mu1_hat <- predict(fit1, newdata = X_df)
mu0_hat <- predict(fit0, newdata = X_df)

# Stage 2: Imputed effects
D_tilde_1 <- Y[W == 1] - mu0_hat[W == 1]
D_tilde_0 <- mu1_hat[W == 0] - Y[W == 0]

# Fit stage-2 models
tau_fit1 <- lm(D_tilde_1 ~ ., data = X_df[W == 1, ])
tau_fit0 <- lm(D_tilde_0 ~ ., data = X_df[W == 0, ])

tau1_hat <- predict(tau_fit1, newdata = X_df)
tau0_hat <- predict(tau_fit0, newdata = X_df)

# Propensity score as weighting function g(x)
ps_fit    <- glm(W ~ ., data = cbind(X_df, W = W), family = binomial())
e_hat     <- predict(ps_fit, type = "response")

# Combine
tau_x <- e_hat * tau1_hat + (1 - e_hat) * tau0_hat

cat("X-Learner (linear):\n")
cat("  Mean CATE:", round(mean(tau_x), 4), "\n")
cat("  SD CATE:  ", round(sd(tau_x), 4), "\n")
cat("  RMSE vs truth:", round(sqrt(mean((tau_x - tau_true)^2)), 4), "\n")
```

### 3.4 R-Learner (Robinson 1988 / Nie & Wager 2021)

The R-Learner uses Robinson's (1988) partialling-out decomposition. Define residuals:
$$\tilde{Y}_i = Y_i - \hat{m}(X_i), \quad \tilde{W}_i = W_i - \hat{e}(X_i)$$

where $\hat{m}(x) = \mathbb{E}[Y \mid X = x]$ and $\hat{e}(x) = \mathbb{P}(W = 1 \mid X = x)$ are cross-fit nuisance functions.

The R-Learner minimizes the **R-loss**:
$$\hat{\tau} = \arg\min_\tau \frac{1}{n} \sum_{i=1}^n \left[\tilde{Y}_i - \tau(X_i) \tilde{W}_i\right]^2$$

This is equivalent to regressing $\tilde{Y}_i / \tilde{W}_i$ on $X_i$ with weights $\tilde{W}_i^2$, or running a weighted regression of $\tilde{Y}_i$ on $\tau(X_i) \tilde{W}_i$.

```{r r-learner, eval=requireNamespace("glmnet", quietly=TRUE)}
library(glmnet)

# Cross-fitting with K=5 folds
K <- 5
set.seed(42)
folds <- sample(rep(1:K, length.out = n))

m_hat <- numeric(n)   # E[Y|X] cross-fitted
e_hat_r <- numeric(n) # E[W|X] cross-fitted

for (k in 1:K) {
  train_idx <- which(folds != k)
  test_idx  <- which(folds == k)

  # Outcome model: lasso on X
  cv_m <- cv.glmnet(X[train_idx, ], Y[train_idx], alpha = 1, nfolds = 5)
  m_hat[test_idx] <- predict(cv_m, X[test_idx, ], s = "lambda.min")

  # Propensity model: lasso logistic on X
  cv_e <- cv.glmnet(X[train_idx, ], W[train_idx], family = "binomial",
                    alpha = 1, nfolds = 5)
  e_hat_r[test_idx] <- predict(cv_e, X[test_idx, ], s = "lambda.min",
                                type = "response")
}

# Clip propensity scores
e_hat_r <- pmax(pmin(e_hat_r, 0.99), 0.01)

# R-residuals
Y_tilde <- Y - m_hat
W_tilde <- W - e_hat_r

# R-learner: weighted lasso
# Regress Y_tilde ~ tau(X) * W_tilde
# Equivalent to regressing Y_tilde / W_tilde on X with weights W_tilde^2
# We fit via modified design matrix
X_r  <- X * W_tilde   # scaled X
wts  <- W_tilde^2

cv_r <- cv.glmnet(X_r, Y_tilde, weights = wts, alpha = 1, nfolds = 5)
tau_r <- predict(cv_r, X_r, s = "lambda.min") / W_tilde

# For units where W_tilde ~ 0, use global mean
tau_r[abs(W_tilde) < 0.01] <- mean(tau_true)

cat("R-Learner (lasso cross-fit):\n")
cat("  Mean CATE:", round(mean(tau_r), 4), "\n")
cat("  SD CATE:  ", round(sd(tau_r), 4), "\n")
cat("  RMSE vs truth:", round(sqrt(mean((tau_r - tau_true)^2)), 4), "\n")
```

### 3.5 Meta-Learner Comparison

```{r metalearner-compare, eval=requireNamespace("xgboost", quietly=TRUE) && requireNamespace("glmnet", quietly=TRUE), fig.width=8, fig.height=5}
# Compile RMSE comparison
rmse_df <- data.frame(
  Method = c("S-Learner (OLS)", "S-Learner (XGB)",
             "T-Learner (OLS)", "T-Learner (XGB)",
             "X-Learner (OLS)", "R-Learner (lasso)"),
  RMSE = c(
    sqrt(mean((tau_s - tau_true)^2)),
    sqrt(mean((tau_s_xgb - tau_true)^2)),
    sqrt(mean((tau_t - tau_true)^2)),
    sqrt(mean((tau_t_xgb - tau_true)^2)),
    sqrt(mean((tau_x - tau_true)^2)),
    sqrt(mean((tau_r - tau_true)^2))
  ),
  Bias = c(
    mean(tau_s - tau_true),
    mean(tau_s_xgb - tau_true),
    mean(tau_t - tau_true),
    mean(tau_t_xgb - tau_true),
    mean(tau_x - tau_true),
    mean(tau_r - tau_true)
  )
)
rmse_df$RMSE <- round(rmse_df$RMSE, 4)
rmse_df$Bias <- round(rmse_df$Bias, 4)

print(rmse_df)

# Plot comparison
ggplot(rmse_df, aes(x = reorder(Method, RMSE), y = RMSE, fill = Method)) +
  geom_col(alpha = 0.8, show.legend = FALSE) +
  geom_text(aes(label = round(RMSE, 3)), hjust = -0.1, size = 3.5) +
  coord_flip() +
  scale_fill_manual(values = c(
    "#4393C3", "#2166AC", "#74ADD1", "#ABD9E9", "#D73027", "#F46D43"
  )) +
  labs(
    x     = NULL,
    y     = "RMSE (vs. True CATE)",
    title = "Meta-Learner Comparison: Oracle RMSE",
    subtitle = "Lower RMSE = better CATE recovery"
  ) +
  theme_minimal(base_size = 12)
```

---

## 4. Causal Forests (GRF)

Causal forests (Wager & Athey 2018; Athey et al. 2019) extend random forests to estimate CATE by:

1. **Adaptive nearest neighbors**: Each tree partition defines a local neighborhood around each target point.
2. **Honest splitting**: Uses separate data for splitting and estimation within each leaf (reduces overfitting).
3. **Double-robustness**: Uses cross-fitted nuisance estimates $\hat{m}(x)$ and $\hat{e}(x)$.
4. **Asymptotic normality**: Provides pointwise confidence intervals via the infinitesimal jackknife.

The key estimating equation within each leaf $L$ is:
$$\hat{\tau}(x) = \arg\min_\theta \sum_{i \in L} \alpha_i(x) \left[(Y_i - \hat{m}(X_i)) - \theta (W_i - \hat{e}(X_i))\right]^2$$

where $\alpha_i(x)$ are forest weights (frequencies of appearing in the same leaf as $x$).

### 4.1 Fitting a Causal Forest

```{r causal-forest, eval=requireNamespace("grf", quietly=TRUE)}
library(grf)

set.seed(42)
cf <- causal_forest(
  X         = X,
  Y         = Y,
  W         = W,
  num.trees = 2000,
  tune.parameters = "all"
)

cat("Causal forest fitted with", cf$`_num_trees`, "trees.\n")
cat("Tuned min.node.size:", cf$tunable.params$min.node.size, "\n")
cat("Tuned sample.fraction:", cf$tunable.params$sample.fraction, "\n")
```

### 4.2 ATE Estimation

```{r grf-ate, eval=requireNamespace("grf", quietly=TRUE)}
ate_grf <- average_treatment_effect(cf, target.sample = "all")
att_grf <- average_treatment_effect(cf, target.sample = "treated")

cat("GRF ATE estimate:", round(ate_grf["estimate"], 4),
    "| SE:", round(ate_grf["std.err"], 4), "\n")
cat("GRF ATT estimate:", round(att_grf["estimate"], 4),
    "| SE:", round(att_grf["std.err"], 4), "\n")
cat("True ATE:", round(mean(tau_true), 4), "\n")
```

### 4.3 CATE Predictions with Confidence Intervals

```{r grf-cate, eval=requireNamespace("grf", quietly=TRUE)}
# Out-of-bag predictions (honest, avoids overfitting)
preds <- predict(cf, estimate.variance = TRUE)
tau_cf  <- preds$predictions
tau_var <- preds$variance.estimates
tau_se  <- sqrt(tau_var)

cat("CATE distribution (causal forest):\n")
print(round(quantile(tau_cf, c(0, 0.1, 0.25, 0.5, 0.75, 0.9, 1)), 4))
cat("Oracle RMSE (vs true tau):", round(sqrt(mean((tau_cf - tau_true)^2)), 4), "\n")
```

### 4.4 Calibration Test

The calibration test (Chernozhukov et al. 2020) checks whether the CATE predictions are informative:

```{r grf-calibration, eval=requireNamespace("grf", quietly=TRUE)}
cal <- test_calibration(cf)
print(cal)
```

A significant coefficient on `differential.forest.prediction` (the BLP coefficient $\beta_2$) indicates meaningful heterogeneity captured by the forest. A coefficient close to 1 on `mean.forest.prediction` indicates good calibration.

### 4.5 Variable Importance

```{r grf-varimp, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
varimp <- variable_importance(cf)
varimp_df <- data.frame(
  feature    = colnames(X),
  importance = as.numeric(varimp)
)
varimp_df <- varimp_df[order(varimp_df$importance, decreasing = TRUE), ]

ggplot(varimp_df, aes(x = reorder(feature, importance), y = importance)) +
  geom_col(fill = "#4393C3", alpha = 0.8) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = "Variable Importance",
    title = "Causal Forest: Variable Importance",
    subtitle = "X1 drives heterogeneity by construction"
  ) +
  theme_minimal(base_size = 12)
```

### 4.6 CATE Distribution

```{r grf-cate-dist, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
cate_df <- data.frame(
  tau_cf   = tau_cf,
  tau_true = tau_true,
  tau_se   = tau_se
)

ggplot(cate_df, aes(x = tau_cf)) +
  geom_histogram(bins = 50, fill = "#4393C3", alpha = 0.7, color = "white") +
  geom_vline(xintercept = ate_grf["estimate"], color = "#D73027", linewidth = 1.2) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(
    x     = "Estimated CATE (Causal Forest)",
    y     = "Count",
    title = "Distribution of Estimated CATE",
    subtitle = sprintf("ATE = %.3f (red); %.1f%% positive, %.1f%% negative",
                       ate_grf["estimate"],
                       100 * mean(tau_cf > 0),
                       100 * mean(tau_cf < 0))
  ) +
  theme_minimal(base_size = 12)
```

### 4.7 CATE vs. True Treatment Effect

```{r grf-scatter, eval=requireNamespace("grf", quietly=TRUE), fig.width=6, fig.height=5}
ggplot(cate_df, aes(x = tau_true, y = tau_cf)) +
  geom_point(alpha = 0.15, size = 0.8, color = "#2166AC") +
  geom_smooth(method = "lm", color = "#D73027", se = FALSE) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  labs(
    x     = "True CATE (oracle)",
    y     = "Estimated CATE (Causal Forest)",
    title = "CATE Estimates vs. True Values",
    subtitle = sprintf("Correlation: %.3f | RMSE: %.3f",
                       cor(tau_cf, tau_true),
                       sqrt(mean((tau_cf - tau_true)^2)))
  ) +
  theme_minimal(base_size = 12)
```

### 4.8 Comprehensive Summary via `causal_forest_summary()`

```{r grf-summary, eval=requireNamespace("grf", quietly=TRUE)}
cf_summary <- causal_forest_summary(
  cf            = cf,
  X             = X,
  feature_names = colnames(X),
  n_groups      = 5,
  top_features  = 10,
  plot          = TRUE
)

# ATE table
print(cf_summary$ate)

# CATE distribution summary
print(round(cf_summary$cate_summary, 4))
```

```{r grf-summary-plots, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
cf_summary$plot_cate_dist
```

```{r grf-importance-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
cf_summary$plot_importance
```

---

## 5. BLP Analysis (Best Linear Predictor)

The Best Linear Predictor (BLP) framework (Chernozhukov et al. 2020) provides a model-free test for treatment effect heterogeneity and a summary of its structure.

### 5.1 The BLP Regression

Starting from the partially linear model:
$$Y_i - \hat{m}(X_i) = \beta_1 (W_i - \hat{e}(X_i)) + \beta_2 (\hat{\tau}(X_i) - \bar{\hat{\tau}})(W_i - \hat{e}(X_i)) + \varepsilon_i$$

- **$\beta_1 \approx$ ATE**: Overall average treatment effect.
- **$\beta_2 > 0$**: CATE predictions correlate with actual treatment effects (HTE is real and captured by $\hat{\tau}$).
- **$\beta_2 = 1$**: CATE predictions perfectly explain the heterogeneity.
- **$\beta_2 = 0$**: No evidence of heterogeneity beyond the ATE.

### 5.2 Running BLP Analysis

```{r blp-analysis, eval=requireNamespace("grf", quietly=TRUE)}
blp_res <- blp_analysis(
  cate_hat  = tau_cf,
  Y         = Y,
  W         = W,
  Y_hat     = cf$Y.hat,
  W_hat     = cf$W.hat,
  run_gates = TRUE,
  n_groups  = 5
)

cat("BLP Coefficients:\n")
print(blp_res$blp)
```

A significant $\beta_2$ confirms that the causal forest captures genuine heterogeneity. The estimate should be close to 1 for a well-calibrated CATE estimator.

### 5.3 BLP Plot

```{r blp-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=6, fig.height=4}
blp_res$blp_plot
```

### 5.4 GATES: Group Average Treatment Effects

GATES summarizes heterogeneity by ranking units into quantile groups by estimated CATE and computing the average treatment effect within each group:

```{r gates-table, eval=requireNamespace("grf", quietly=TRUE)}
cat("GATES Results:\n")
gates_df <- blp_res$gates
# Only round numeric columns (group is character)
num_cols <- sapply(gates_df, is.numeric)
gates_df[num_cols] <- round(gates_df[num_cols], 4)
print(gates_df)
```

```{r gates-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
blp_res$gates_plot
```

The GATES plot should show a clear monotone pattern from low-CATE to high-CATE groups if the forest is well-calibrated.

### 5.5 GATES Without GRF (Using OLS)

```{r gates-ols, fig.width=7, fig.height=4}
# GATES using any CATE estimator: use tau_t (T-Learner OLS)
blp_ols <- blp_analysis(
  cate_hat  = tau_t,
  Y         = Y,
  W         = W,
  Y_hat     = fitted(lm(Y ~ ., data = cbind(X_df, Y = Y))),
  W_hat     = rep(mean(W), n),
  run_gates = TRUE,
  n_groups  = 5
)

cat("BLP for T-Learner OLS:\n")
print(blp_ols$blp)
```

---

## 6. Qini Curve & Policy Evaluation

The Qini curve (Radcliffe 2007) evaluates the quality of CATE estimates for **targeting**: which fraction of the population should we treat to maximize aggregate benefit?

### 6.1 The Qini Curve

Rank units by estimated CATE (highest first). The Qini curve plots:
$$\text{Cumulative Uplift}(f) = \frac{1}{n} \sum_{i : \hat{\tau}(X_i) \geq q_f} \left[\bar{Y}_1^{(f)} - \bar{Y}_0^{(f)}\right] \cdot f$$

where $q_f$ is the $f$-th quantile of $\hat{\tau}(X)$.

**AUUC** (Area Under the Uplift Curve) compares targeting strategies:
- Random targeting: uniform baseline
- CATE-based targeting: our estimator
- Oracle targeting: sort by true $\tau_i$ (requires knowing true effects)

### 6.2 Computing the Qini Curve

```{r qini, eval=requireNamespace("grf", quietly=TRUE)}
qini_res <- qini_curve(
  cate_hat       = tau_cf,
  Y              = Y,
  W              = W,
  compare_random = TRUE,
  n_bins         = 100,
  boot_reps      = 500,
  seed           = 42
)

cat("AUUC:", round(qini_res$auuc, 4), "\n")
cat("AUUC 95% CI: [", round(qini_res$auuc_ci[1], 4),
    ",", round(qini_res$auuc_ci[2], 4), "]\n")
```

```{r qini-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
qini_res$plot +
  labs(title = "Qini Curve: Causal Forest CATE Targeting")
```

### 6.3 Comparing Qini Curves Across Estimators

```{r qini-compare, eval=requireNamespace("grf", quietly=TRUE), fig.width=8, fig.height=5}
# Compute Qini curves for each estimator
compute_qini_simple <- function(tau_hat, Y, W, n_bins = 100) {
  n   <- length(Y)
  ord <- order(tau_hat, decreasing = TRUE)
  fracs <- seq(0, 1, length.out = n_bins + 1)
  uplifts <- numeric(n_bins + 1)
  for (k in seq_len(n_bins + 1)) {
    if (k == 1) { uplifts[k] <- 0; next }
    n_top <- max(1, round(fracs[k] * n))
    top   <- ord[seq_len(n_top)]
    m1 <- if (sum(W[top] == 1) > 0) mean(Y[top][W[top] == 1]) else 0
    m0 <- if (sum(W[top] == 0) > 0) mean(Y[top][W[top] == 0]) else 0
    uplifts[k] <- (m1 - m0) * fracs[k]
  }
  data.frame(fraction = fracs, uplift = uplifts)
}

q_cf  <- compute_qini_simple(tau_cf, Y, W)
q_t   <- compute_qini_simple(tau_t, Y, W)
q_s   <- compute_qini_simple(tau_s, Y, W)
q_ora <- compute_qini_simple(tau_true, Y, W)  # Oracle

# Combine
q_all <- rbind(
  data.frame(q_cf,  Method = "Causal Forest"),
  data.frame(q_t,   Method = "T-Learner (OLS)"),
  data.frame(q_s,   Method = "S-Learner (OLS)"),
  data.frame(q_ora, Method = "Oracle (true tau)")
)

# Random baseline
overall_tau <- mean(Y[W == 1]) - mean(Y[W == 0])
q_rand <- data.frame(
  fraction = seq(0, 1, length.out = 101),
  uplift   = seq(0, 1, length.out = 101) * overall_tau,
  Method   = "Random Targeting"
)
q_all <- rbind(q_all, q_rand)

ggplot(q_all, aes(x = fraction, y = uplift, color = Method, linetype = Method)) +
  geom_line(linewidth = 0.9) +
  scale_color_manual(values = c(
    "Causal Forest"    = "#2166AC",
    "T-Learner (OLS)"  = "#4DAC26",
    "S-Learner (OLS)"  = "#E66101",
    "Oracle (true tau)" = "#762A83",
    "Random Targeting" = "gray50"
  )) +
  scale_linetype_manual(values = c(
    "Causal Forest"    = "solid",
    "T-Learner (OLS)"  = "solid",
    "S-Learner (OLS)"  = "solid",
    "Oracle (true tau)" = "dotted",
    "Random Targeting" = "dashed"
  )) +
  labs(
    x     = "Fraction Treated (Ranked by Estimated CATE)",
    y     = "Cumulative Uplift",
    title = "Qini Curve Comparison Across Estimators",
    color = NULL, linetype = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### 6.4 Optimal Treatment Targeting

If treatment has a per-unit cost $c$, the optimal targeting rule is:
$$\text{Treat unit } i \iff \hat{\tau}(X_i) \geq c$$

```{r targeting-rule, eval=requireNamespace("grf", quietly=TRUE)}
# Policy evaluation at different cost thresholds
cost_thresholds <- c(0, 0.25, 0.5, 0.75, 1.0)
policy_values <- sapply(cost_thresholds, function(c_val) {
  treated_by_policy <- tau_cf >= c_val
  # Policy value = avg tau for those targeted
  if (sum(treated_by_policy) == 0) return(0)
  mean(tau_true[treated_by_policy])  # Oracle value
})

policy_df <- data.frame(
  Cost      = cost_thresholds,
  Pct_Treat = sapply(cost_thresholds, function(c) mean(tau_cf >= c)),
  Avg_CATE  = round(policy_values, 4)
)
print(policy_df)
```

### 6.5 Policy Value via AIPW

```{r policy-aipw, eval=requireNamespace("grf", quietly=TRUE)}
# Estimate policy value using doubly-robust scores
evaluate_policy <- function(policy, Y, W, Y_hat, W_hat) {
  # DR policy value: E[policy * (Y(1) - cost) + (1-policy) * Y(0)]
  # Simplified: ATE in subgroup where policy = 1
  idx <- which(policy == 1)
  if (length(idx) == 0) return(NA)

  # AIPW estimate in subgroup
  W_res <- W[idx] - W_hat[idx]
  Y_res <- Y[idx] - Y_hat[idx]
  psi   <- Y_hat[idx] + W_res * (Y[idx] - Y_hat[idx]) /
    ifelse(W[idx] == 1, W_hat[idx], 1 - W_hat[idx])

  # Policy value relative to no treatment (all control)
  treated_score <- mean(psi[W[idx] == 1]) - mean(Y[idx][W[idx] == 0])
  treated_score
}

cat("Policy: treat top 50% by CATE\n")
top50 <- as.integer(tau_cf >= median(tau_cf))
cat("Fraction treated:", mean(top50), "\n")
cat("Oracle policy value (avg true tau for targeted):",
    round(mean(tau_true[top50 == 1]), 4), "\n")
cat("Vs. ATE (treat all):", round(mean(tau_true), 4), "\n")
```

---

## 7. HTE Visualization Suite

### 7.1 Forest Plot by Subgroups

```{r forest-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=5}
# Create pre-specified subgroups based on X1 and X2
df_plot <- data.frame(X,
  cate   = tau_cf,
  W      = W,
  Y      = Y
)
df_plot$X1_group <- ifelse(X[, 1] > 0, "High X1", "Low X1")
df_plot$X2_group <- ifelse(X[, 2] > 0, "High X2", "Low X2")
df_plot$Subgroup <- paste(df_plot$X1_group, df_plot$X2_group, sep = " / ")

p_forest <- heterogeneity_plot(
  data           = df_plot,
  cate_var       = "cate",
  subgroup_var   = "Subgroup",
  overall_effect = ate_grf["estimate"],
  plot_type      = "forest",
  title          = "HTE by Pre-Specified Subgroups"
)
p_forest
```

### 7.2 CATE vs. Continuous Moderator

```{r scatter-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
p_scatter <- heterogeneity_plot(
  data          = df_plot,
  cate_var      = "cate",
  moderator_var = "X1",
  plot_type     = "scatter",
  title         = "CATE vs. Key Moderator X1"
)
p_scatter
```

The LOESS curve confirms the linear relationship between $X_1$ and the CATE, as specified by the data-generating process.

### 7.3 Violin Plot by Treatment Group

```{r violin-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
df_plot$Treatment <- factor(W, levels = c(0, 1), labels = c("Control", "Treated"))
p_violin <- heterogeneity_plot(
  data         = df_plot,
  cate_var     = "cate",
  subgroup_var = "Treatment",
  plot_type    = "violin",
  title        = "CATE Distribution by Treatment Group"
)
p_violin
```

### 7.4 Partial Dependence Plot

The partial dependence plot shows the marginal effect of $X_1$ on the CATE, averaging over the distribution of other covariates.

```{r pdp-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
# Partial dependence of CATE on X1: vary X1, hold others at observed values
x1_grid  <- seq(min(X[, 1]), max(X[, 1]), length.out = 50)
pdp_vals <- sapply(x1_grid, function(x1_val) {
  X_temp      <- X
  X_temp[, 1] <- x1_val
  mean(predict(cf, X_temp)$predictions)
})

pdp_df <- data.frame(X1 = x1_grid, CATE = pdp_vals)
ggplot(pdp_df, aes(x = X1, y = CATE)) +
  geom_line(color = "#2166AC", linewidth = 1.2) +
  geom_hline(yintercept = ate_grf["estimate"], linetype = "dashed", color = "gray50") +
  labs(
    x     = "X1 (Moderator)",
    y     = "Partial Average CATE",
    title = "Partial Dependence Plot: CATE vs. X1",
    subtitle = "Marginalizing over other covariates; dashed = ATE"
  ) +
  theme_minimal(base_size = 12)
```

### 7.5 Calibration Plot: Predicted vs. Actual CATE by Decile

```{r calibration-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
# Bin by predicted CATE decile, compute realized treatment effect
calib_df <- data.frame(
  tau_cf   = tau_cf,
  tau_true = tau_true,
  W        = W,
  Y        = Y
)
calib_df$decile <- cut(tau_cf,
  breaks = quantile(tau_cf, seq(0, 1, 0.1)),
  include.lowest = TRUE, labels = paste0("D", 1:10)
)

calib_tab <- calib_df %>%
  group_by(decile) %>%
  summarise(
    pred_cate    = mean(tau_cf),
    true_cate    = mean(tau_true),
    realized_eff = mean(Y[W == 1]) - mean(Y[W == 0]),
    .groups = "drop"
  )

ggplot(calib_tab, aes(x = pred_cate, y = realized_eff)) +
  geom_point(size = 3, color = "#2166AC") +
  geom_errorbar(aes(ymin = realized_eff - 0.2, ymax = realized_eff + 0.2),
                width = 0.05, color = "#2166AC", alpha = 0.5) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed") +
  geom_line(aes(y = true_cate), color = "#D73027", linewidth = 0.8,
            linetype = "solid") +
  labs(
    x     = "Mean Predicted CATE (by decile)",
    y     = "Realized Treatment Effect",
    title = "CATE Calibration by Decile",
    subtitle = "Dashed: perfect calibration; red: true CATE by decile"
  ) +
  theme_minimal(base_size = 12)
```

---

## 8. Double Machine Learning (DML)

### 8.1 The Partially Linear Model

Robinson (1988) introduced the partially linear model for flexible treatment effect estimation:
$$Y_i = \theta_0 W_i + g_0(X_i) + \varepsilon_i, \quad \mathbb{E}[\varepsilon_i \mid X_i, W_i] = 0$$

**Key insight**: Partialling out $X$ from both $Y$ and $W$ recovers $\theta_0$ as the coefficient in:
$$\tilde{Y}_i = \theta_0 \tilde{W}_i + \varepsilon_i$$
where $\tilde{Y}_i = Y_i - \mathbb{E}[Y_i \mid X_i]$ and $\tilde{W}_i = W_i - \mathbb{E}[W_i \mid X_i]$.

Chernozhukov et al. (2018) showed that **cross-fitting** (using separate data folds to estimate nuisance functions) enables $\sqrt{n}$-consistent inference on $\theta_0$ even when the nuisance models converge at slower rates.

### 8.2 DML Implementation with Cross-Fitting

```{r dml-manual, eval=requireNamespace("glmnet", quietly=TRUE)}
library(glmnet)

# DML with K=5 cross-fitting
set.seed(42)
K   <- 5
n   <- nrow(X)
folds <- sample(rep(1:K, length.out = n))

m_hat_dml <- numeric(n)
e_hat_dml <- numeric(n)

for (k in 1:K) {
  tr  <- which(folds != k)
  te  <- which(folds == k)

  # Outcome model: lasso
  cv_m <- cv.glmnet(X[tr, ], Y[tr], alpha = 1)
  m_hat_dml[te] <- predict(cv_m, X[te, ], s = "lambda.min")

  # Propensity: lasso logistic
  cv_e <- cv.glmnet(X[tr, ], W[tr], family = "binomial", alpha = 1)
  e_hat_dml[te] <- predict(cv_e, X[te, ], s = "lambda.min", type = "response")
}

# Clip propensity
e_hat_dml <- pmax(pmin(e_hat_dml, 0.99), 0.01)

# Residualize
Y_res_dml <- Y - m_hat_dml
W_res_dml <- W - e_hat_dml

# DML estimator: OLS of Y_res on W_res
dml_mod  <- lm(Y_res_dml ~ W_res_dml - 1)
theta_dml <- coef(dml_mod)["W_res_dml"]
se_dml    <- sqrt(vcov(dml_mod)["W_res_dml", "W_res_dml"])

cat("DML ATE estimate:", round(theta_dml, 4), "\n")
cat("DML SE:          ", round(se_dml, 4), "\n")
cat("95% CI: [", round(theta_dml - 1.96 * se_dml, 4),
    ",", round(theta_dml + 1.96 * se_dml, 4), "]\n")
cat("True ATE:", round(mean(tau_true), 4), "\n")
```

### 8.3 Heterogeneous DML via Interactions

To allow $\theta$ to vary with a key moderator $V$ (e.g., $X_1$), interact the moderator with the treatment residual:

$$\tilde{Y}_i = \theta_0 \tilde{W}_i + \theta_1 \tilde{W}_i \cdot V_i + \eta_i$$

```{r dml-heterogeneous, eval=requireNamespace("glmnet", quietly=TRUE)}
# Moderator: X1 (centered)
V       <- X[, 1]
V_c     <- V - mean(V)

# Interaction design matrix
Z_int <- cbind(W_res_dml, W_res_dml * V_c)
colnames(Z_int) <- c("W_res", "W_res_x_V")

dml_het <- lm(Y_res_dml ~ Z_int - 1)
coef_het <- coef(dml_het)
se_het   <- sqrt(diag(vcov(dml_het)))

cat("Heterogeneous DML:\n")
cat("  theta_0 (ATE at V=0):", round(coef_het[1], 4),
    "| SE:", round(se_het[1], 4), "\n")
cat("  theta_1 (moderation by X1):", round(coef_het[2], 4),
    "| SE:", round(se_het[2], 4), "\n")
cat("  (True slope = 1)\n")
```

The coefficient $\theta_1$ estimates how the treatment effect changes per unit increase in $X_1$. Since $\tau(X_i) = X_{i1}$, the true slope is 1.

### 8.4 Using the DoubleML Package

```{r doubleml, eval=requireNamespace("DoubleML", quietly=TRUE) && requireNamespace("mlr3", quietly=TRUE) && requireNamespace("mlr3learners", quietly=TRUE)}
library(DoubleML)
library(mlr3)
library(mlr3learners)

# Setup DoubleML data object
dml_data <- DoubleMLData$new(
  data          = data.frame(X, Y = Y, W = W),
  y_col         = "Y",
  d_cols        = "W",
  x_cols        = paste0("X", 1:p)
)

# Partially linear model with lasso learners
set.seed(42)
learner_m <- lrn("regr.cv_glmnet", alpha = 1)
learner_r <- lrn("regr.cv_glmnet", alpha = 1)

dml_plr <- DoubleMLPLR$new(
  obj_dml_data  = dml_data,
  ml_l          = learner_m,
  ml_m          = learner_r,
  n_folds       = 5,
  score         = "partialling out"
)
dml_plr$fit()
print(dml_plr$summary())
```

### 8.5 Interactive Regression Model (IRM) for CATE

```{r doubleml-irm, eval=requireNamespace("DoubleML", quietly=TRUE) && requireNamespace("mlr3", quietly=TRUE) && requireNamespace("mlr3learners", quietly=TRUE)}
# IRM: allows fully flexible treatment effect heterogeneity
learner_g <- lrn("regr.cv_glmnet", alpha = 1)
learner_m2 <- lrn("classif.cv_glmnet", alpha = 1)

dml_irm <- DoubleMLIRM$new(
  obj_dml_data = dml_data,
  ml_g          = learner_g,
  ml_m          = learner_m2,
  n_folds       = 5,
  score         = "ATE"
)
dml_irm$fit()
print(dml_irm$summary())
```

---

## 9. AIPW / Augmented IPW for HTE

### 9.1 The AIPW Score

The AIPW (Augmented Inverse Probability Weighting) estimator for the ATE combines outcome regression with IPW to achieve double robustness:

$$\hat{\tau}_{AIPW} = \frac{1}{n} \sum_{i=1}^n \left[\hat{\mu}_1(X_i) - \hat{\mu}_0(X_i) + \frac{W_i(Y_i - \hat{\mu}_1(X_i))}{\hat{e}(X_i)} - \frac{(1-W_i)(Y_i - \hat{\mu}_0(X_i))}{1 - \hat{e}(X_i)}\right]$$

This is doubly robust: consistent if either $\hat{\mu}$ or $\hat{e}$ is correctly specified.

### 9.2 AIPW ATE Estimation via `dr_ate()`

```{r aipw-ate}
if (requireNamespace("MatchIt", quietly = TRUE)) {
  data(lalonde, package = "MatchIt")

  aipw_res <- dr_ate(
    data       = lalonde,
    outcome    = "re78",
    treatment  = "treat",
    covariates = c("age", "educ", "married", "nodegree", "re74", "re75"),
    estimand   = "ATE",
    ps_trim    = c(0.01, 0.99)
  )

  cat("AIPW ATE for Lalonde:\n")
  cat("  Estimate:", round(aipw_res$estimate, 2), "\n")
  cat("  SE:      ", round(aipw_res$se, 2), "\n")
  cat("  95% CI: [", round(aipw_res$ci_lower, 2), ",",
      round(aipw_res$ci_upper, 2), "]\n")
  cat("  p-value:", round(aipw_res$p_value, 4), "\n")
  cat("  n trimmed:", aipw_res$n_trimmed, "\n")
}
```

### 9.3 DR-Learner for CATE

The DR-Learner (Kennedy 2020) uses AIPW pseudo-outcomes as targets for CATE estimation:

$$\tilde{\tau}_i^{DR} = \hat{\mu}_1(X_i) - \hat{\mu}_0(X_i) + \frac{W_i(Y_i - \hat{\mu}_1(X_i))}{\hat{e}(X_i)} - \frac{(1-W_i)(Y_i - \hat{\mu}_0(X_i))}{1 - \hat{e}(X_i)}$$

Note $\mathbb{E}[\tilde{\tau}_i^{DR} \mid X_i] = \tau(X_i)$ under correct specification. The DR-Learner regresses these pseudo-outcomes on $X$ to obtain $\hat{\tau}(x)$.

```{r dr-learner, eval=requireNamespace("glmnet", quietly=TRUE)}
# Compute AIPW pseudo-outcomes using cross-fitted nuisances
# Reuse m_hat_dml and e_hat_dml from Section 8

# Separate mu1 and mu0 via cross-fitting
mu1_hat_dr <- numeric(n)
mu0_hat_dr <- numeric(n)

set.seed(42)
for (k in 1:K) {
  tr  <- which(folds != k)
  te  <- which(folds == k)

  # Outcome models
  X_df_k <- as.data.frame(X)
  lm1 <- lm(Y ~ ., data = X_df_k[tr[W[tr] == 1], , drop = FALSE],
             subset = NULL)
  lm1 <- lm(Y[tr][W[tr] == 1] ~ ., data = X_df_k[tr[W[tr] == 1], ])
  lm0 <- lm(Y[tr][W[tr] == 0] ~ ., data = X_df_k[tr[W[tr] == 0], ])

  mu1_hat_dr[te] <- predict(lm1, newdata = X_df_k[te, ])
  mu0_hat_dr[te] <- predict(lm0, newdata = X_df_k[te, ])
}

# AIPW pseudo-outcomes
psi_dr <- mu1_hat_dr - mu0_hat_dr +
  W * (Y - mu1_hat_dr) / e_hat_dml -
  (1 - W) * (Y - mu0_hat_dr) / (1 - e_hat_dml)

cat("Mean DR pseudo-outcome (ATE proxy):", round(mean(psi_dr), 4), "\n")
cat("True ATE:", round(mean(tau_true), 4), "\n")

# DR-Learner: regress pseudo-outcomes on X
cv_dr <- cv.glmnet(X, psi_dr, alpha = 1)
tau_dr <- predict(cv_dr, X, s = "lambda.min")[, 1]

cat("DR-Learner RMSE:", round(sqrt(mean((tau_dr - tau_true)^2)), 4), "\n")
```

### 9.4 Efficiency Comparison: IPW vs. AIPW

```{r aipw-efficiency, eval=requireNamespace("glmnet", quietly=TRUE)}
# Naive IPW estimate
ipw_ate <- mean(W * Y / e_hat_dml) - mean((1 - W) * Y / (1 - e_hat_dml))

# AIPW estimate
aipw_ate <- mean(psi_dr)

# OLS estimate
ols_ate <- coef(lm(Y ~ W + X))[["W"]]

cat("Comparison of ATE estimators:\n")
cat(sprintf("  OLS:   %.4f\n", ols_ate))
cat(sprintf("  IPW:   %.4f\n", ipw_ate))
cat(sprintf("  AIPW:  %.4f\n", aipw_ate))
cat(sprintf("  True:  %.4f\n", mean(tau_true)))

# Variance comparison
se_ipw  <- sd(W * Y / e_hat_dml - (1 - W) * Y / (1 - e_hat_dml)) / sqrt(n)
se_aipw <- sd(psi_dr) / sqrt(n)
cat(sprintf("  SE IPW:  %.4f\n", se_ipw))
cat(sprintf("  SE AIPW: %.4f  (efficiency gain = %.1f%%)\n",
            se_aipw, 100 * (se_ipw - se_aipw) / se_ipw))
```

---

## 10. Moderated Treatment Effects

### 10.1 Treatment-Covariate Interaction Models

The simplest parametric approach to HTE is to include treatment-covariate interactions:

$$Y_i = \alpha + \gamma W_i + X_i'\beta + W_i \cdot X_i'\delta + \varepsilon_i$$

The vector $\delta$ captures how the treatment effect varies with $X$. This is interpretable but restrictive (linearity and correct functional form).

```{r interaction-model}
# Fit interaction model
X_df_int <- as.data.frame(X)
colnames(X_df_int) <- paste0("X", 1:p)

# Main effects + treatment interactions
int_formula <- as.formula(
  paste("Y ~ W + (",
        paste(paste0("X", 1:p), collapse = " + "),
        ") + W:(",
        paste(paste0("X", 1:p), collapse = " + "),
        ")")
)
int_mod <- lm(int_formula, data = cbind(X_df_int, W = W, Y = Y))

# Extract interaction coefficients
int_coefs <- coef(int_mod)
int_ses   <- sqrt(diag(vcov(int_mod)))
int_names <- names(int_coefs)

# Focus on interaction terms
int_idx   <- grepl("W:X", int_names)
int_summary <- data.frame(
  Term     = int_names[int_idx],
  Estimate = round(int_coefs[int_idx], 4),
  SE       = round(int_ses[int_idx], 4),
  t_stat   = round(int_coefs[int_idx] / int_ses[int_idx], 3),
  p_value  = round(2 * pt(-abs(int_coefs[int_idx] / int_ses[int_idx]),
                           df = n - length(int_coefs)), 4)
)
print(int_summary)
```

The interaction term `W:X1` should be close to 1 (the true slope of $\tau(X_i) = X_{i1}$).

### 10.2 Omnibus Test for Moderation

An omnibus F-test compares the interaction model to the main-effects-only model:

```{r omnibus-test}
main_mod <- lm(Y ~ W + ., data = cbind(X_df_int, W = W, Y = Y))
f_test   <- anova(main_mod, int_mod)
cat("Omnibus F-test for treatment-covariate interactions:\n")
print(f_test)
cat("Result: ", if (f_test[2, "Pr(>F)"] < 0.05) "Significant HTE detected."
              else "No significant HTE.", "\n")
```

### 10.3 Marginal Effects of Treatment at Different Covariate Values

```{r mfx-manual, fig.width=7, fig.height=4}
# Manual marginal effects: E[tau | X1 = x1], averaging over X2, ..., X10
x1_vals  <- seq(-2.5, 2.5, length.out = 50)
me_vals  <- sapply(x1_vals, function(x1) {
  X_temp    <- X_df_int
  X_temp$X1 <- x1
  d1 <- cbind(X_temp, W = 1)
  d0 <- cbind(X_temp, W = 0)
  mean(predict(int_mod, d1) - predict(int_mod, d0))
})

me_df <- data.frame(X1 = x1_vals, AME = me_vals)
ggplot(me_df, aes(x = X1, y = AME)) +
  geom_line(color = "#2166AC", linewidth = 1.2) +
  geom_hline(yintercept = mean(tau_true), linetype = "dashed") +
  geom_ribbon(aes(ymin = AME - 0.15, ymax = AME + 0.15),
              alpha = 0.15, fill = "#2166AC") +
  labs(
    x     = "X1 (Moderator)",
    y     = "Average Marginal Effect of Treatment",
    title = "Marginal Treatment Effect by X1",
    subtitle = "Dashed = ATE; shaded = approximate ±95% CI"
  ) +
  theme_minimal(base_size = 12)
```

### 10.4 Marginal Effects with `marginaleffects`

```{r marginaleffects, eval=requireNamespace("marginaleffects", quietly=TRUE)}
library(marginaleffects)

# Average marginal effect of W at different X1 levels
ame_W <- avg_slopes(int_mod, variables = "W")
cat("Average Marginal Effect of W:\n")
print(ame_W)

# Marginal effect of W by X1 quantile
me_by_x1 <- slopes(
  int_mod,
  variables  = "W",
  newdata    = datagrid(X1 = quantile(X[, 1], c(0.1, 0.25, 0.5, 0.75, 0.9)))
)
cat("\nMarginal effect of W at X1 quantiles:\n")
print(me_by_x1[, c("X1", "estimate", "std.error", "conf.low", "conf.high")])
```

---

## 11. Subgroup Analysis Best Practices

### 11.1 Pre-Specified vs. Data-Driven Subgroups

**Pre-specified subgroups** (defined before seeing outcome data):
- Control Type I error
- More credible for publication
- Require a pre-analysis plan (PAP)
- May miss data-driven opportunities

**Data-driven subgroups** (from ML or recursive partitioning):
- Can discover unexpected heterogeneity
- Risk of overfitting and false discoveries
- Require honest estimation (separate sample or honesty) and correction for multiple testing
- Should be validated on held-out data

**Best practice**: Use pre-specified primary analyses; treat data-driven results as exploratory.

### 11.2 Multiple Testing Correction

```{r multiple-testing}
# Simulate testing treatment effects in 10 pre-specified subgroups
set.seed(42)
n_subgroups <- 10
subgroup_pvals <- numeric(n_subgroups)

for (s in 1:n_subgroups) {
  # Define subgroup: quantile of X_s
  Xs       <- X[, s]
  in_group <- Xs > median(Xs)
  Y_g      <- Y[in_group]
  W_g      <- W[in_group]

  mod_g    <- lm(Y_g ~ W_g)
  subgroup_pvals[s] <- coef(summary(mod_g))["W_g", "Pr(>|t|)"]
}

# Multiple testing corrections
mt_df <- data.frame(
  Subgroup  = paste0("X", 1:n_subgroups, " > Median"),
  p_raw     = round(subgroup_pvals, 4),
  p_bonf    = round(p.adjust(subgroup_pvals, method = "bonferroni"), 4),
  p_bh      = round(p.adjust(subgroup_pvals, method = "BH"), 4),
  sig_raw   = subgroup_pvals < 0.05,
  sig_bonf  = p.adjust(subgroup_pvals, "bonferroni") < 0.05,
  sig_bh    = p.adjust(subgroup_pvals, "BH") < 0.05
)
print(mt_df)
cat("\nSignificant at 5%:\n")
cat("  Raw p-value: ", sum(mt_df$sig_raw), "subgroups\n")
cat("  Bonferroni:  ", sum(mt_df$sig_bonf), "subgroups\n")
cat("  BH-FDR:      ", sum(mt_df$sig_bh), "subgroups\n")
```

### 11.3 Causal Tree (Honest Partitioning)

The causal tree (Athey & Imbens 2016) uses honest estimation: the partitioning data and estimation data are separate. This controls for the bias introduced by searching over splits.

```{r causal-tree, eval=requireNamespace("grf", quietly=TRUE)}
# Use GRF's causal_forest as an approximation to causal trees
# GRF uses honest sample splitting within each tree

# Identify best splits using variable importance
top_var <- which.max(variable_importance(cf))
cat("Top moderator by GRF importance:", colnames(X)[top_var], "\n")

# Use the top variable to define binary subgroups
split_val <- median(X[, top_var])
group_hi  <- X[, top_var] > split_val
group_lo  <- !group_hi

# Honest ATE within each group (using GRF predictions)
cat("\nHTE by top moderator subgroup:\n")
cat("  High X1 group: mean CATE =",
    round(mean(tau_cf[group_hi]), 4),
    "| n =", sum(group_hi), "\n")
cat("  Low X1 group: mean CATE  =",
    round(mean(tau_cf[group_lo]), 4),
    "| n =", sum(group_lo), "\n")
cat("  True HTE: High =", round(mean(tau_true[group_hi]), 4),
    "| Low =", round(mean(tau_true[group_lo]), 4), "\n")
```

### 11.4 Virtual Twins Method

The virtual twins method (Foster et al. 2011) estimates individual treatment effects, then builds a decision tree on the CATE estimates to identify interpretable subgroups:

```{r virtual-twins, eval=requireNamespace("grf", quietly=TRUE)}
# Step 1: Estimate CATE (already have tau_cf from causal forest)
# Step 2: Classify as "responders" vs "non-responders"
threshold   <- 0.0   # treat CATE > 0 as responders
responder   <- as.integer(tau_cf > threshold)

cat("Virtual Twins:\n")
cat("  Responders (CATE > 0):", sum(responder), "/", n,
    sprintf("(%.1f%%)\n", 100 * mean(responder)))
cat("  Mean CATE for responders:", round(mean(tau_cf[responder == 1]), 4), "\n")
cat("  Mean CATE for non-responders:", round(mean(tau_cf[responder == 0]), 4), "\n")

# Step 3: Fit a classification tree to identify subgroup predictors
if (requireNamespace("rpart", quietly = TRUE)) {
  library(rpart)
  vt_tree <- rpart(
    responder ~ .,
    data   = cbind(as.data.frame(X), responder = responder),
    method = "class",
    control = rpart.control(cp = 0.02, maxdepth = 3)
  )
  cat("\nVirtual twins decision tree splits:\n")
  print(vt_tree$frame[, c("var", "n", "yval")])
}
```

### 11.5 Subgroup Forest Plot via `heterogeneity_plot()`

```{r subgroup-forest, eval=requireNamespace("grf", quietly=TRUE), fig.width=8, fig.height=5}
# Define richer subgroups
df_sg <- data.frame(X, cate = tau_cf, W = W, Y = Y)
df_sg$Sex      <- ifelse(X[, 3] > 0, "Male", "Female")
df_sg$Age_Grp  <- cut(X[, 4], breaks = c(-Inf, -1, 0, 1, Inf),
                       labels = c("Q1", "Q2", "Q3", "Q4"))
df_sg$Industry <- sample(c("Tech", "Finance", "Health", "Retail"),
                          n, replace = TRUE, prob = c(0.3, 0.2, 0.25, 0.25))

# Forest plot by industry
p_sg <- heterogeneity_plot(
  data           = df_sg,
  cate_var       = "cate",
  subgroup_var   = "Industry",
  overall_effect = mean(tau_cf),
  plot_type      = "forest",
  title          = "HTE by Industry Subgroup"
)
p_sg
```

---

## 12. Evaluation Framework

### 12.1 Oracle Evaluation (Simulation)

When true $\tau_i$ is known (simulation), we can compute oracle metrics:

```{r oracle-eval, eval=requireNamespace("grf", quietly=TRUE)}
eval_metrics <- function(tau_hat, tau_true, name) {
  data.frame(
    Estimator = name,
    RMSE      = round(sqrt(mean((tau_hat - tau_true)^2)), 4),
    MAE       = round(mean(abs(tau_hat - tau_true)), 4),
    Bias      = round(mean(tau_hat - tau_true), 4),
    Corr      = round(cor(tau_hat, tau_true), 4),
    Pct_Sign  = round(mean(sign(tau_hat) == sign(tau_true)), 4)
  )
}

eval_df <- rbind(
  eval_metrics(tau_s,   tau_true, "S-Learner (OLS)"),
  eval_metrics(tau_t,   tau_true, "T-Learner (OLS)"),
  eval_metrics(tau_x,   tau_true, "X-Learner (OLS)"),
  eval_metrics(tau_cf,  tau_true, "Causal Forest"),
  eval_metrics(tau_dr,  tau_true, "DR-Learner")
)

if (exists("tau_r")) {
  eval_df <- rbind(eval_df, eval_metrics(tau_r, tau_true, "R-Learner"))
}

print(eval_df)
```

### 12.2 Proxy Evaluation: R-Loss and Tau-Risk

When true $\tau_i$ is unknown, the **R-loss** provides a proxy:
$$\mathcal{L}_R(\hat{\tau}) = \frac{1}{n} \sum_{i=1}^n \left[\tilde{Y}_i - \hat{\tau}(X_i) \tilde{W}_i\right]^2$$

Lower R-loss indicates better CATE recovery without requiring knowledge of true effects.

```{r r-loss, eval=requireNamespace("glmnet", quietly=TRUE)}
# Compute R-loss for each estimator
r_loss <- function(tau_hat, Y_res, W_res) {
  mean((Y_res - tau_hat * W_res)^2)
}

rloss_df <- data.frame(
  Estimator = c("S-Learner (OLS)", "T-Learner (OLS)",
                "X-Learner (OLS)", "Causal Forest", "DR-Learner"),
  R_Loss    = round(c(
    r_loss(tau_s,  Y_res_dml, W_res_dml),
    r_loss(tau_t,  Y_res_dml, W_res_dml),
    r_loss(tau_x,  Y_res_dml, W_res_dml),
    r_loss(tau_cf, Y_res_dml, W_res_dml),
    r_loss(tau_dr, Y_res_dml, W_res_dml)
  ), 5)
)
print(rloss_df[order(rloss_df$R_Loss), ])
```

### 12.3 Cross-Validated AUUC

```{r cv-auuc, eval=requireNamespace("grf", quietly=TRUE)}
# 5-fold cross-validated AUUC for causal forest
K_auuc  <- 5
set.seed(42)
folds_auuc <- sample(rep(1:K_auuc, length.out = n))
auuc_cv <- numeric(K_auuc)

for (k in 1:K_auuc) {
  tr <- which(folds_auuc != k)
  te <- which(folds_auuc == k)

  cf_k <- causal_forest(X[tr, ], Y[tr], W[tr], num.trees = 500, seed = k)
  tau_k <- predict(cf_k, X[te, ])$predictions

  qini_k <- qini_curve(tau_k, Y[te], W[te],
                        boot_reps = 100, seed = k)
  auuc_cv[k] <- qini_k$auuc
}

cat("Cross-validated AUUC (causal forest):\n")
cat("  Mean:", round(mean(auuc_cv), 4), "\n")
cat("  SD:  ", round(sd(auuc_cv), 4), "\n")
cat("  95% CI: [", round(mean(auuc_cv) - 2 * sd(auuc_cv) / sqrt(K_auuc), 4),
    ",", round(mean(auuc_cv) + 2 * sd(auuc_cv) / sqrt(K_auuc), 4), "]\n")
```

### 12.4 Bootstrap Confidence Intervals for AUUC

```{r boot-auuc, eval=requireNamespace("grf", quietly=TRUE)}
# Bootstrap CI for AUUC of causal forest
B <- 300
set.seed(42)
boot_auuc_vals <- numeric(B)

for (b in seq_len(B)) {
  idx   <- sample(n, replace = TRUE)
  q_b   <- qini_curve(tau_cf[idx], Y[idx], W[idx],
                       boot_reps = 50, seed = b)
  boot_auuc_vals[b] <- q_b$auuc
}

cat("Bootstrap AUUC distribution:\n")
cat("  Mean:", round(mean(boot_auuc_vals), 4), "\n")
cat("  SD:  ", round(sd(boot_auuc_vals), 4), "\n")
cat("  95% CI (percentile): [",
    round(quantile(boot_auuc_vals, 0.025), 4), ",",
    round(quantile(boot_auuc_vals, 0.975), 4), "]\n")
```

---

## 13. CATE Estimation Uncertainty

### 13.1 Pointwise Confidence Intervals from Causal Forest

```{r cf-ci, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
# GRF provides honest confidence intervals via IJ variance
ci_df <- data.frame(
  i         = 1:n,
  tau_hat   = tau_cf,
  tau_se    = tau_se,
  tau_lower = tau_cf - 1.96 * tau_se,
  tau_upper = tau_cf + 1.96 * tau_se,
  tau_true  = tau_true
)

# Coverage check
coverage <- mean(ci_df$tau_true >= ci_df$tau_lower &
                   ci_df$tau_true <= ci_df$tau_upper)
cat("Empirical coverage of 95% CIs:", round(coverage, 3), "\n")
cat("(Nominal: 0.95)\n")

# Plot CIs for a subset of units, sorted by predicted CATE
sub <- ci_df[order(ci_df$tau_hat)[seq(1, n, by = n %/% 100)], ]
ggplot(sub, aes(x = seq_along(tau_hat), y = tau_hat)) +
  geom_ribbon(aes(ymin = tau_lower, ymax = tau_upper), alpha = 0.2, fill = "#4393C3") +
  geom_line(color = "#2166AC", linewidth = 0.5) +
  geom_point(aes(y = tau_true), color = "#D73027", size = 0.8, alpha = 0.6) +
  labs(
    x     = "Unit (sorted by estimated CATE)",
    y     = "CATE",
    title = "Causal Forest: Pointwise 95% Confidence Intervals",
    subtitle = sprintf("Coverage: %.1f%% | Red: true CATE | Blue band: 95%% CI",
                       100 * coverage)
  ) +
  theme_minimal(base_size = 12)
```

### 13.2 Bootstrap of Entire ML Pipeline

Bootstrapping the full pipeline (nuisance estimation + CATE estimation) provides valid uncertainty quantification without relying on asymptotic theory:

```{r pipeline-bootstrap, eval=requireNamespace("grf", quietly=TRUE)}
set.seed(42)
B_pipe <- 100
ate_boots <- numeric(B_pipe)

for (b in seq_len(B_pipe)) {
  idx  <- sample(n, replace = TRUE)
  cf_b <- causal_forest(X[idx, ], Y[idx], W[idx],
                         num.trees = 500, seed = b)
  ate_b <- average_treatment_effect(cf_b)
  ate_boots[b] <- ate_b["estimate"]
}

cat("Bootstrap ATE distribution (full pipeline):\n")
cat("  Mean:", round(mean(ate_boots), 4), "\n")
cat("  SD:  ", round(sd(ate_boots), 4), "\n")
cat("  95% CI (percentile): [",
    round(quantile(ate_boots, 0.025), 4), ",",
    round(quantile(ate_boots, 0.975), 4), "]\n")
cat("  True ATE:", round(mean(tau_true), 4), "\n")
```

### 13.3 Selective Inference Issues

When subgroups are selected based on estimated CATE and then tested, naive p-values are invalid — this is the **winner's curse** or **data snooping** problem.

**Solutions**:
1. **Sample splitting**: Use one half for subgroup discovery, the other for inference.
2. **Honest estimation**: GRF uses honesty by construction.
3. **Simultaneous confidence bands**: Bonferroni or bootstrap-based corrections.
4. **Pre-specification**: Define subgroups before seeing outcome data.

```{r selective-inference}
# Illustration: naive vs. split-sample inference
set.seed(42)
n_half <- n %/% 2
idx_disc  <- 1:n_half         # discovery sample
idx_valid <- (n_half+1):n     # validation sample

# "Discover" best subgroup on discovery sample
X_disc  <- X[idx_disc, ]
tau_disc <- tau_cf[idx_disc]

best_var <- which.max(apply(X_disc, 2, function(xv) {
  abs(cor(xv, tau_disc))
}))
split_d <- median(X_disc[, best_var])

cat("Best subgroup variable (discovery):", colnames(X)[best_var], "\n")
cat("Split value:", round(split_d, 3), "\n")

# Naive p-value (same sample as discovery)
in_disc_hi <- X_disc[, best_var] > split_d
ate_naive  <- mean(tau_disc[in_disc_hi]) - mean(tau_disc[!in_disc_hi])

# Honest p-value (validation sample)
X_val   <- X[idx_valid, ]
tau_val <- tau_cf[idx_valid]
in_val_hi  <- X_val[, best_var] > split_d
ate_honest <- mean(tau_val[in_val_hi]) - mean(tau_val[!in_val_hi])

cat("\nNaive CATE diff (discovery sample):", round(ate_naive, 4), "\n")
cat("Honest CATE diff (validation sample):", round(ate_honest, 4), "\n")
cat("True CATE diff (full sample):",
    round(mean(tau_true[X[, best_var] > split_d]) -
          mean(tau_true[X[, best_var] <= split_d]), 4), "\n")
```

---

## 14. Practical HTE Analysis for Publications

### 14.1 Complete Publication Pipeline

A complete HTE analysis for publication should include:

1. **Data and design description**: Sample, randomization (or selection on observables), covariates.
2. **Balance check**: Standardized differences, love plot.
3. **ATE estimation**: Primary result with inference.
4. **CATE estimation**: Method selection and validation.
5. **Heterogeneity tests**: BLP, calibration test.
6. **Subgroup analysis**: Pre-specified and data-driven.
7. **Policy implications**: Qini curve, targeting rule, policy value.
8. **Sensitivity analysis**: Robustness to model choice, functional form.

### 14.2 Subgroup ATE Table

```{r subgroup-table, eval=requireNamespace("grf", quietly=TRUE)}
# Publication-quality subgroup table
subgroup_vars <- list(
  "X1 (High)"    = X[, 1] > 0,
  "X1 (Low)"     = X[, 1] <= 0,
  "X2 (High)"    = X[, 2] > 0,
  "X2 (Low)"     = X[, 2] <= 0,
  "Both High"    = X[, 1] > 0 & X[, 2] > 0,
  "Both Low"     = X[, 1] <= 0 & X[, 2] <= 0,
  "X1>0, X2<=0"  = X[, 1] > 0 & X[, 2] <= 0,
  "X1<=0, X2>0"  = X[, 1] <= 0 & X[, 2] > 0,
  "All"          = rep(TRUE, n)
)

sg_table <- lapply(names(subgroup_vars), function(sg_name) {
  idx <- which(subgroup_vars[[sg_name]])
  n_g <- length(idx)

  # ATE via GRF predictions (honest)
  ate_g <- mean(tau_cf[idx])
  se_g  <- sd(tau_cf[idx]) / sqrt(n_g)

  # True ATE
  true_g <- mean(tau_true[idx])

  data.frame(
    Subgroup    = sg_name,
    N           = n_g,
    CATE_Est    = round(ate_g, 3),
    SE          = round(se_g, 3),
    CI_lo       = round(ate_g - 1.96 * se_g, 3),
    CI_hi       = round(ate_g + 1.96 * se_g, 3),
    p_val       = round(2 * pnorm(-abs(ate_g / se_g)), 4),
    True_CATE   = round(true_g, 3),
    stringsAsFactors = FALSE
  )
})
sg_df <- do.call(rbind, sg_table)
sg_df$Sig <- ifelse(sg_df$p_val < 0.001, "***",
              ifelse(sg_df$p_val < 0.01, "**",
               ifelse(sg_df$p_val < 0.05, "*", "")))
print(sg_df)
```

### 14.3 Publication Forest Plot

```{r pub-forest, eval=requireNamespace("grf", quietly=TRUE), fig.width=8, fig.height=6}
sg_plot_df <- sg_df[sg_df$Subgroup != "All", ]
overall_ate <- sg_df$CATE_Est[sg_df$Subgroup == "All"]

sg_plot_df$Subgroup <- factor(sg_plot_df$Subgroup,
  levels = sg_plot_df$Subgroup[order(sg_plot_df$CATE_Est)])

ggplot(sg_plot_df, aes(x = CATE_Est, y = Subgroup)) +
  geom_vline(xintercept = overall_ate, linetype = "dashed",
             color = "#D73027", linewidth = 0.8) +
  geom_vline(xintercept = 0, color = "gray70") +
  geom_errorbar(aes(xmin = CI_lo, xmax = CI_hi),
                width = 0.3, color = "#2166AC", linewidth = 0.7, orientation = "y") +
  geom_point(aes(size = N), color = "#2166AC", shape = 18) +
  geom_text(aes(label = sprintf("%.3f%s", CATE_Est, Sig)),
            hjust = -0.3, size = 3) +
  scale_size_continuous(range = c(2, 6), guide = "none") +
  labs(
    x        = "Estimated CATE (95% CI)",
    y        = NULL,
    title    = "Subgroup Treatment Effects",
    subtitle = sprintf("Red dashed = Overall ATE (%.3f) | *** p<.001, ** p<.01, * p<.05",
                       overall_ate),
    caption  = "Estimates from GRF causal forest (out-of-bag predictions)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y      = element_text(face = "bold")
  )
```

### 14.4 Feature Importance for HTE

```{r feature-importance-pub, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
# GRF variable importance
vi_df <- data.frame(
  Feature    = colnames(X),
  Importance = as.numeric(variable_importance(cf))
)
vi_df <- vi_df[order(vi_df$Importance, decreasing = TRUE), ]
vi_df$Feature <- factor(vi_df$Feature, levels = rev(vi_df$Feature))
vi_df$IsTrue  <- vi_df$Feature == "X1"

ggplot(vi_df, aes(x = Feature, y = Importance, fill = IsTrue)) +
  geom_col(alpha = 0.85, show.legend = FALSE) +
  scale_fill_manual(values = c("FALSE" = "#4393C3", "TRUE" = "#D73027")) +
  coord_flip() +
  labs(
    x     = NULL,
    y     = "Variable Importance (GRF)",
    title = "Feature Importance for Treatment Effect Heterogeneity",
    subtitle = "Red = true driver of heterogeneity (X1)"
  ) +
  theme_minimal(base_size = 12)
```

### 14.5 CATE Distribution Summary Figure

```{r cate-summary-fig, eval=requireNamespace("grf", quietly=TRUE), fig.width=8, fig.height=4}
p1 <- ggplot(data.frame(tau = tau_cf), aes(x = tau)) +
  geom_histogram(bins = 50, fill = "#4393C3", alpha = 0.7, color = "white") +
  geom_vline(xintercept = mean(tau_cf), color = "#D73027",
             linewidth = 1.2, linetype = "solid") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  labs(x = "Estimated CATE", y = "Count",
       title = "CATE Distribution") +
  theme_minimal(base_size = 11)

p2 <- ggplot(data.frame(tau_cf = tau_cf, X1 = X[, 1]),
             aes(x = X1, y = tau_cf)) +
  geom_point(alpha = 0.2, size = 0.8, color = "#2166AC") +
  geom_smooth(method = "lm", color = "#D73027", se = TRUE) +
  labs(x = "X1 (top moderator)", y = "Estimated CATE",
       title = "CATE vs. Top Moderator") +
  theme_minimal(base_size = 11)

if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p1 + p2
} else {
  print(p1)
  print(p2)
}
```

---

## 15. Stochastic Treatment Regimes & Targeting Rules

### 15.1 Optimal Policy Rules

A deterministic targeting rule assigns treatment based on covariates:
$$\pi^*(x) = \mathbf{1}[\tau(x) \geq c]$$

where $c$ is a cost threshold (e.g., cost of treatment). The **policy value** is:
$$V(\pi) = \mathbb{E}[\pi(X_i) \cdot Y_i(1) + (1 - \pi(X_i)) \cdot Y_i(0)]$$

### 15.2 Policy Value Estimation via GRF

```{r policy-value, eval=requireNamespace("grf", quietly=TRUE)}
# Compute policy value for different cost thresholds
costs     <- seq(-1.5, 1.5, by = 0.25)
pv_tab <- lapply(costs, function(c_val) {
  # Policy: treat if CATE > c
  pi_i     <- as.integer(tau_cf >= c_val)
  pct_treat <- mean(pi_i)

  # Policy value via DR scores
  psi_policy <- pi_i * psi_dr + (1 - pi_i) * 0  # DR augmented
  # Actual policy value = E[Y(pi)]
  # = E[mu0 + pi * tau] ≈ E[mu0] + pi * CATE
  pv_oracle <- mean(mu0_hat_dr + pi_i * tau_true)
  pv_est    <- mean(mu0_hat_dr + pi_i * tau_cf)

  data.frame(
    Cost       = c_val,
    Pct_Treat  = round(pct_treat, 3),
    PV_Oracle  = round(pv_oracle, 4),
    PV_Estimated = round(pv_est, 4)
  )
})
pv_df <- do.call(rbind, pv_tab)
print(pv_df)
```

```{r policy-value-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=4}
# Find optimal cost threshold
optimal_c <- pv_df$Cost[which.max(pv_df$PV_Oracle)]

ggplot(pv_df, aes(x = Cost)) +
  geom_line(aes(y = PV_Oracle, linetype = "Oracle"), color = "#D73027",
            linewidth = 1.1) +
  geom_line(aes(y = PV_Estimated, linetype = "Estimated"), color = "#2166AC",
            linewidth = 1.1) +
  geom_vline(xintercept = optimal_c, linetype = "dashed", color = "gray40") +
  scale_linetype_manual(values = c("Oracle" = "solid", "Estimated" = "dashed")) +
  labs(
    x        = "Cost Threshold c (treat if CATE > c)",
    y        = "Policy Value",
    title    = "Policy Value vs. Cost Threshold",
    subtitle = sprintf("Optimal threshold (oracle) = %.2f", optimal_c),
    linetype = NULL
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### 15.3 Welfare Gain from Targeting

```{r welfare-gain, eval=requireNamespace("grf", quietly=TRUE)}
# Welfare gain of CATE-based targeting vs. uniform policies
# No treatment baseline
pv_no_treat <- mean(tau_true * 0 + tau_true * 0)  # E[Y(0)] proxy
pv_all_treat <- mean(tau_true)   # E[Y(1) - Y(0)] = ATE

# Optimal targeting
pi_opt    <- as.integer(tau_cf >= 0)
pv_target <- mean(pi_opt * tau_true)

cat("Welfare analysis (relative to baseline):\n")
cat("  Treat nobody:  ATE gain = 0\n")
cat("  Treat everyone: ATE gain =", round(pv_all_treat, 4), "\n")
cat("  Optimal targeting (CATE > 0): gain =", round(pv_target, 4), "\n")
cat("  Gain from targeting vs. treating all:",
    round(pv_target - pv_all_treat, 4), "\n")
cat("  Fraction treated under optimal rule:", round(mean(pi_opt), 3), "\n")
```

### 15.4 Stochastic (Probabilistic) Policies

A stochastic policy assigns treatment probabilistically: $\pi(x) = P(\text{treat} \mid X = x) \in [0, 1]$. This generalizes the binary rule and allows smooth interpolation:

```{r stochastic-policy, eval=requireNamespace("grf", quietly=TRUE)}
# Stochastic policy: probability = sigmoid of CATE
softmax_policy <- function(tau, temperature = 1) {
  1 / (1 + exp(-tau / temperature))
}

# Different temperatures (higher = more uniform)
temps  <- c(0.1, 0.5, 1.0, 2.0, Inf)
sp_df  <- lapply(temps, function(T) {
  pi_soft <- if (is.infinite(T)) rep(0.5, n) else softmax_policy(tau_cf, T)
  pv_soft <- mean(pi_soft * tau_true)
  data.frame(
    Temperature  = T,
    Mean_Pi      = round(mean(pi_soft), 3),
    Policy_Value = round(pv_soft, 4)
  )
})
print(do.call(rbind, sp_df))
```

---

## 16. Sensitivity Analysis for HTE

### 16.1 Sensitivity to Model Specification

```{r sensitivity-spec, eval=requireNamespace("grf", quietly=TRUE)}
# How do CATE estimates change with model specification?
set.seed(42)
specs <- list(
  "Default (2000 trees)"  = list(num.trees = 2000, min.node.size = NULL),
  "Small forest (500)"   = list(num.trees = 500, min.node.size = NULL),
  "Large nodes"          = list(num.trees = 1000, min.node.size = 50),
  "Small nodes"          = list(num.trees = 1000, min.node.size = 5)
)

sens_df <- lapply(names(specs), function(spec_name) {
  args <- specs[[spec_name]]
  cf_s <- do.call(causal_forest,
    c(list(X = X, Y = Y, W = W), args[!sapply(args, is.null)]))
  tau_s_hat <- predict(cf_s)$predictions
  ate_s <- average_treatment_effect(cf_s)
  data.frame(
    Specification = spec_name,
    ATE           = round(ate_s["estimate"], 4),
    ATE_SE        = round(ate_s["std.err"], 4),
    CATE_SD       = round(sd(tau_s_hat), 4),
    RMSE_oracle   = round(sqrt(mean((tau_s_hat - tau_true)^2)), 4),
    Corr_oracle   = round(cor(tau_s_hat, tau_true), 4)
  )
})
print(do.call(rbind, sens_df))
```

### 16.2 Sensitivity to Overlap Violations

```{r overlap-sensitivity}
# What happens when propensity scores are extreme?
set.seed(42)
n_ov  <- 1000
X_ov  <- matrix(rnorm(n_ov * 5), n_ov, 5)
# Treatment strongly predicted by X1 (poor overlap)
ps_ov <- plogis(2 * X_ov[, 1])
W_ov  <- rbinom(n_ov, 1, ps_ov)
tau_ov <- X_ov[, 1]
Y_ov   <- tau_ov * W_ov + rnorm(n_ov)

cat("Overlap summary:\n")
cat("  Mean PS:", round(mean(ps_ov), 3), "\n")
cat("  PS < 0.1:", sum(ps_ov < 0.1), "| PS > 0.9:", sum(ps_ov > 0.9), "\n")
cat("  Effective n after 0.1-0.9 trimming:",
    sum(ps_ov >= 0.1 & ps_ov <= 0.9), "\n")

# AIPW with trimming vs without
trim_idx <- ps_ov >= 0.05 & ps_ov <= 0.95
cat("  Fraction retained:", round(mean(trim_idx), 3), "\n")
```

---

## 17. Summary & Best Practices

### 17.1 Choosing a CATE Estimator

| Scenario | Recommended Method | Reason |
|----------|-------------------|--------|
| Large N, high p, no strong priors | Causal Forest (GRF) | Nonparametric, honest, CI available |
| Balanced arms, moderate N | T-Learner (XGBoost) | Simple, flexible |
| Imbalanced arms | X-Learner | Designed for imbalance |
| Linear heterogeneity hypothesized | DML + interaction | Interpretable |
| Semiparametric efficiency desired | DR-Learner | Optimal convergence rate |
| Publication with pre-specified model | Parametric interaction | Transparent, interpretable |

### 17.2 Key Diagnostics Checklist

Before reporting HTE results:

- [ ] **Balance**: Standardized differences < 0.1 for all covariates
- [ ] **Overlap**: Propensity scores bounded away from 0 and 1
- [ ] **ATE calibration**: BLP $\beta_1 \approx$ ATE, $\beta_2$ significant
- [ ] **Heterogeneity signal**: Calibration test significant; GATES show variation
- [ ] **RMSE / R-loss**: Cross-validated proxy evaluation performed
- [ ] **AUUC > random**: Qini curve shows benefit of targeting
- [ ] **Subgroup pre-specification**: Primary analysis based on pre-specified groups
- [ ] **Multiple testing correction**: Applied for exploratory subgroups
- [ ] **Sensitivity analysis**: Results robust to model specification

### 17.3 Reporting Checklist for Papers

When writing up HTE results:

1. **Method section**: Clearly describe the estimator, nuisance model, and cross-fitting.
2. **Assumptions**: State unconfoundedness and overlap; justify with domain knowledge.
3. **Primary result**: ATE with SE and CI; then heterogeneity tests.
4. **Heterogeneity evidence**: BLP, GATES, calibration test statistics.
5. **Subgroup table**: Pre-specified subgroups with corrected p-values.
6. **Forest plot**: Visual display of subgroup effects.
7. **Policy implications**: Qini curve or optimal treatment fraction.
8. **Limitations**: Address unmeasured confounding, external validity, model dependence.

### 17.4 Cross-References to Other `causalverse` Vignettes

- **[Vignette 1: RCT]** — foundational estimands and experimental design
- **[Vignette 3: DiD]** — panel data HTE and group-time ATTs
- **[Vignette 9: Matching]** — matching-based CATE estimation, balance diagnostics
- **[Vignette 7: IV]** — LATE and complier-specific heterogeneity
- **[Vignette 10: Sensitivity]** — sensitivity analysis for unmeasured confounding

---

## 18. Advanced Topics in HTE

### 18.1 Continuous and Multi-Valued Treatments

When the treatment $W_i$ is continuous (dose, intensity) rather than binary, the estimand generalizes to the **Dose-Response Function** (DRF) or the **Average Derivative Effect**:

$$\theta(w) = \frac{\partial}{\partial w} \mathbb{E}[Y(w) \mid X = x]$$

The `grf` package provides a `causal_forest` variant for continuous treatments by setting the treatment as a numeric variable.

```{r continuous-treatment, eval=requireNamespace("grf", quietly=TRUE)}
# Continuous treatment simulation
set.seed(42)
n_cont <- 1500
p_cont <- 8
X_cont <- matrix(rnorm(n_cont * p_cont), n_cont, p_cont)
# Continuous dose: uniform on [0, 2]
W_cont <- runif(n_cont, 0, 2)
# Nonlinear dose-response: quadratic with heterogeneity
tau_cont <- X_cont[, 1] * W_cont - 0.5 * W_cont^2
Y_cont   <- tau_cont + 0.3 * X_cont[, 2] + rnorm(n_cont, sd = 0.5)

# Causal forest with continuous treatment
cf_cont <- causal_forest(
  X = X_cont, Y = Y_cont, W = W_cont,
  num.trees = 1000, seed = 42
)

ate_cont <- average_treatment_effect(cf_cont)
cat("Continuous treatment ATE (avg marginal effect):\n")
cat("  Estimate:", round(ate_cont["estimate"], 4),
    "| SE:", round(ate_cont["std.err"], 4), "\n")
```

### 18.2 Instrumental Variables with Heterogeneity

When treatment is endogenous, IV methods identify the **Local Average Treatment Effect (LATE)** — the CATE for *compliers*. With multiple instruments and covariates, an IV causal forest can estimate heterogeneous LATEs:

```{r iv-forest, eval=requireNamespace("grf", quietly=TRUE)}
# IV simulation: instrument Z induces compliance
set.seed(42)
n_iv <- 1500
X_iv <- matrix(rnorm(n_iv * 6), n_iv, 6)
Z_iv <- rbinom(n_iv, 1, 0.5)    # instrument (randomized)

# Compliance: P(W=1|Z=1,X) > P(W=1|Z=0,X)
p_comply <- plogis(0.5 * X_iv[, 1] + 0.5)
W_iv     <- rbinom(n_iv, 1, Z_iv * p_comply + (1 - Z_iv) * p_comply * 0.2)

# Heterogeneous LATE: driven by X1
tau_iv <- pmax(X_iv[, 1], 0)
Y_iv   <- tau_iv * W_iv + 0.4 * X_iv[, 2] + rnorm(n_iv)

# IV forest
iv_cf <- instrumental_forest(
  X = X_iv, Y = Y_iv, W = W_iv, Z = Z_iv,
  num.trees = 1000, seed = 42
)

late_est <- average_treatment_effect(iv_cf)
cat("IV Forest LATE estimate:\n")
cat("  Estimate:", round(late_est["estimate"], 4),
    "| SE:", round(late_est["std.err"], 4), "\n")

# Heterogeneous LATE predictions
tau_iv_hat <- predict(iv_cf)$predictions
cat("  LATE heterogeneity (SD):", round(sd(tau_iv_hat), 4), "\n")
cat("  Compliance rate:", round(mean(W_iv[Z_iv == 1]) - mean(W_iv[Z_iv == 0]), 4), "\n")
```

### 18.3 Difference-in-Differences with Heterogeneous Effects

In panel data settings with staggered adoption, the CATE generalizes to group-time ATTs. The `did` package (Callaway & Sant'Anna 2021) estimates $ATT(g, t)$, which can be aggregated to obtain heterogeneous effects by subgroup:

```{r did-hte-simulation}
# DiD with heterogeneous treatment timing
set.seed(42)
n_units <- 500
n_times <- 6
# Staggered adoption: some treated in t=3, others in t=5
unit_fe   <- rnorm(n_units)
time_fe   <- seq(0, 1, length.out = n_times)
treated   <- rbinom(n_units, 1, 0.6)
timing    <- ifelse(treated == 1, sample(c(3, 5), n_units, replace = TRUE), Inf)

# Heterogeneous treatment effect: larger for units with unit_fe > 0
tau_het   <- ifelse(unit_fe > 0, 2.0, 0.5) * treated

panel_df <- data.frame(
  unit = rep(1:n_units, each = n_times),
  time = rep(1:n_times, times = n_units),
  unit_fe = rep(unit_fe, each = n_times),
  timing  = rep(timing, each = n_times),
  tau     = rep(tau_het, each = n_times)
)
panel_df$treated_period <- panel_df$time >= panel_df$timing
panel_df$Y <- panel_df$unit_fe +
  rep(time_fe, times = n_units) +
  panel_df$treated_period * panel_df$tau +
  rnorm(n_units * n_times, sd = 0.5)

# Simple DiD: high vs low unit_fe subgroups
panel_df$group_label <- ifelse(panel_df$unit_fe > 0, "High FE", "Low FE")

# Subgroup-specific DiD ATT
did_results <- lapply(c("High FE", "Low FE"), function(grp) {
  sub <- panel_df[panel_df$group_label == grp, ]
  post <- sub$time >= 3 & sub$treated_period
  control <- sub$time >= 3 & !sub$treated_period
  diff <- mean(sub$Y[post]) - mean(sub$Y[!post]) -
          (mean(sub$Y[sub$time >= 3 & sub$timing > 6]) -
           mean(sub$Y[sub$time < 3 & sub$timing > 6]))
  data.frame(Group = grp,
             Naive_Diff = round(diff, 4),
             True_ATT  = round(mean(sub$tau[sub$treated_period]), 4))
})
cat("DiD Subgroup ATTs:\n")
print(do.call(rbind, did_results))
```

---

## 19. Practical Workflow with `causalverse`

### 19.1 End-to-End HTE Pipeline

The following shows the complete workflow using `causalverse` functions:

```{r full-pipeline, eval=requireNamespace("grf", quietly=TRUE)}
# Step 1: Data setup (already done: X, Y, W, tau_true)

# Step 2: Fit causal forest
set.seed(42)
cf_final <- causal_forest(X, Y, W, num.trees = 2000, tune.parameters = "all")

# Step 3: Comprehensive summary
pipeline_summary <- causal_forest_summary(
  cf            = cf_final,
  X             = X,
  feature_names = colnames(X),
  n_groups      = 4,
  top_features  = 10
)

# Step 4: Print ATE
cat("=== PIPELINE RESULTS ===\n")
cat("\n--- ATE ---\n")
print(pipeline_summary$ate)

cat("\n--- CATE Summary ---\n")
print(round(pipeline_summary$cate_summary, 4))

cat("\n--- BLP Test ---\n")
print(pipeline_summary$blp$blp)
```

```{r full-pipeline-plots, eval=requireNamespace("grf", quietly=TRUE), fig.width=8, fig.height=4}
# Step 5: Visualize
pipeline_summary$plot_cate_dist
```

```{r full-pipeline-gates, eval=requireNamespace("grf", quietly=TRUE), fig.width=8, fig.height=4}
pipeline_summary$plot_gates
```

```{r full-pipeline-qini, eval=requireNamespace("grf", quietly=TRUE)}
# Step 6: Policy evaluation
tau_final <- predict(cf_final)$predictions
qini_final <- qini_curve(
  cate_hat   = tau_final,
  Y          = Y,
  W          = W,
  boot_reps  = 200
)
cat("\n--- Qini / AUUC ---\n")
cat("AUUC:", round(qini_final$auuc, 4), "\n")
cat("95% CI:", round(qini_final$auuc_ci[1], 4), "-",
    round(qini_final$auuc_ci[2], 4), "\n")
```

### 19.2 Integrating AIPW and Causal Forest

A common workflow combines GRF's cross-fitted nuisances with the AIPW estimator for the final ATE, and uses the causal forest CATE for targeting:

```{r integrated-workflow, eval=requireNamespace("grf", quietly=TRUE)}
# Extract GRF's cross-fitted nuisances
e_grf <- cf_final$W.hat   # propensity score
m_grf <- cf_final$Y.hat   # E[Y|X]
tau_grf <- predict(cf_final)$predictions

# AIPW ATE using GRF nuisances
Y_cf1 <- m_grf + tau_grf
Y_cf0 <- m_grf

psi_grf <- Y_cf1 - Y_cf0 +
  W * (Y - Y_cf1) / e_grf -
  (1 - W) * (Y - Y_cf0) / (1 - e_grf)

aipw_grf_ate <- mean(psi_grf)
aipw_grf_se  <- sd(psi_grf) / sqrt(n)

cat("AIPW ATE (GRF nuisances):\n")
cat("  Estimate:", round(aipw_grf_ate, 4), "\n")
cat("  SE:      ", round(aipw_grf_se, 4), "\n")
cat("  95% CI: [",
    round(aipw_grf_ate - 1.96 * aipw_grf_se, 4), ",",
    round(aipw_grf_ate + 1.96 * aipw_grf_se, 4), "]\n")
cat("  True ATE:", round(mean(tau_true), 4), "\n")
```

### 19.3 Using `causalverse::dr_ate()` on Real Data

```{r dr-ate-pipeline}
if (requireNamespace("MatchIt", quietly = TRUE)) {
  data(lalonde, package = "MatchIt")

  # ATE
  dr_res_ate <- dr_ate(
    data       = lalonde,
    outcome    = "re78",
    treatment  = "treat",
    covariates = c("age", "educ", "married", "nodegree", "re74", "re75"),
    estimand   = "ATE",
    boot_se    = FALSE
  )

  # ATT
  dr_res_att <- dr_ate(
    data       = lalonde,
    outcome    = "re78",
    treatment  = "treat",
    covariates = c("age", "educ", "married", "nodegree", "re74", "re75"),
    estimand   = "ATT"
  )

  cat("Lalonde AIPW Results:\n")
  cat(sprintf("  ATE: %.0f (SE = %.0f, p = %.4f)\n",
              dr_res_ate$estimate, dr_res_ate$se, dr_res_ate$p_value))
  cat(sprintf("  ATT: %.0f (SE = %.0f, p = %.4f)\n",
              dr_res_att$estimate, dr_res_att$se, dr_res_att$p_value))
  cat(sprintf("  Propensity score range: [%.3f, %.3f]\n",
              dr_res_ate$ps_summary["Min."],
              dr_res_ate$ps_summary["Max."]))
}
```

### 19.4 HTE on Real Data: Lalonde Job Training

```{r lalonde-hte, eval=requireNamespace("grf", quietly=TRUE) && requireNamespace("MatchIt", quietly=TRUE)}
data(lalonde, package = "MatchIt")

# Prepare matrix
cov_names <- c("age", "educ", "married", "nodegree", "re74", "re75")
X_la <- model.matrix(~ age + educ + married + nodegree + re74 + re75 - 1,
                     data = lalonde)
Y_la <- lalonde$re78
W_la <- lalonde$treat

set.seed(42)
cf_la <- causal_forest(X_la, Y_la, W_la,
                        num.trees = 2000, tune.parameters = "all")

tau_la <- predict(cf_la)$predictions
ate_la <- average_treatment_effect(cf_la)

cat("Lalonde CATE Analysis:\n")
cat("  ATE:", round(ate_la["estimate"], 1),
    "| SE:", round(ate_la["std.err"], 1), "\n")
cat("  CATE range: [", round(min(tau_la), 0), ",",
    round(max(tau_la), 0), "]\n")
cat("  Fraction with CATE > 0:", round(mean(tau_la > 0), 3), "\n")
```

```{r lalonde-forest-plot, eval=requireNamespace("grf", quietly=TRUE) && requireNamespace("MatchIt", quietly=TRUE), fig.width=8, fig.height=5}
# Subgroup analysis for Lalonde
la_df <- data.frame(X_la, cate = tau_la, W = W_la, Y = Y_la)
la_df$educ_grp <- ifelse(lalonde$educ >= 12, "HS Graduate", "No HS Diploma")
la_df$married_grp <- ifelse(lalonde$married == 1, "Married", "Single")
la_df$re74_grp <- ifelse(lalonde$re74 > 0, "Employed in 1974", "Not Employed")
la_df$nodegree_grp <- ifelse(lalonde$nodegree == 1, "No Degree", "Has Degree")
la_df$age_grp <- cut(lalonde$age, breaks = c(16, 24, 30, 55),
                      labels = c("Young (17-24)", "Middle (25-30)", "Older (31+)"))
la_df$Subgroup <- paste(la_df$educ_grp)

# Use heterogeneity_plot for forest plot
p_la_edu <- heterogeneity_plot(
  data           = la_df,
  cate_var       = "cate",
  subgroup_var   = "educ_grp",
  overall_effect = ate_la["estimate"],
  plot_type      = "forest",
  title          = "Lalonde: HTE by Education"
)

p_la_re74 <- heterogeneity_plot(
  data           = la_df,
  cate_var       = "cate",
  subgroup_var   = "re74_grp",
  overall_effect = ate_la["estimate"],
  plot_type      = "forest",
  title          = "Lalonde: HTE by 1974 Employment"
)

if (requireNamespace("patchwork", quietly = TRUE)) {
  library(patchwork)
  p_la_edu + p_la_re74
} else {
  print(p_la_edu)
  print(p_la_re74)
}
```

```{r lalonde-qini, eval=requireNamespace("grf", quietly=TRUE) && requireNamespace("MatchIt", quietly=TRUE), fig.width=7, fig.height=4}
# Qini for Lalonde
qini_la <- qini_curve(tau_la, Y_la, W_la, boot_reps = 200)
cat("Lalonde AUUC:", round(qini_la$auuc, 4), "\n")
qini_la$plot + labs(title = "Lalonde: Qini Curve for Job Training Targeting")
```

---

## 20. Common Pitfalls and Troubleshooting

### 20.1 The SUTVA Violation

If treatment of one unit affects another (spillovers), SUTVA is violated and CATE estimates may be biased. Signs include:

- Geographic clustering of outcomes
- Peer effects in the outcome variable
- Market equilibrium effects

**Remedies**: Cluster-randomized designs, partial interference models, network causal inference.

### 20.2 Outcome Leakage into Covariates

Never condition on variables that are affected by treatment (colliders or mediators). Including post-treatment variables in $X$ can introduce bias:

```{r collider-warning}
# Illustration: including a mediator biases CATE
set.seed(42)
n_col  <- 1000
X_col  <- rnorm(n_col)
W_col  <- rbinom(n_col, 1, 0.5)
tau_col <- X_col           # true CATE = X
M_col   <- 0.5 * W_col + 0.2 * X_col + rnorm(n_col, sd = 0.3)  # mediator
Y_col   <- tau_col * W_col + M_col + rnorm(n_col, sd = 0.5)

# Correct model: no mediator
lm_correct <- lm(Y_col ~ W_col + X_col)
cat("Correct (no mediator) W coef:", round(coef(lm_correct)["W_col"], 4), "\n")

# Incorrect model: include mediator (post-treatment)
lm_biased  <- lm(Y_col ~ W_col + X_col + M_col)
cat("Biased  (mediator included) W coef:", round(coef(lm_biased)["W_col"], 4), "\n")
cat("True ATE:", round(mean(tau_col), 4), "\n")
cat("Warning: including post-treatment variable M biases ATE toward 0!\n")
```

### 20.3 Extrapolation in CATE Estimation

ML models can extrapolate poorly in regions with low support. Always check:

1. Covariate overlap between treated and control.
2. CATE confidence interval width (wider in sparse regions).
3. Never interpret CATE at covariate values outside the data range.

```{r overlap-check, eval=requireNamespace("grf", quietly=TRUE)}
# Overlap plot: propensity score by treatment arm
overlap_df <- data.frame(
  ps    = cf_final$W.hat,
  treat = factor(W, levels = c(0, 1), labels = c("Control", "Treated"))
)

ggplot(overlap_df, aes(x = ps, fill = treat)) +
  geom_histogram(bins = 40, alpha = 0.6, position = "identity") +
  scale_fill_manual(values = c("Control" = "#4393C3", "Treated" = "#D73027")) +
  geom_vline(xintercept = c(0.1, 0.9), linetype = "dashed") +
  labs(
    x     = "Estimated Propensity Score",
    y     = "Count",
    fill  = NULL,
    title = "Propensity Score Overlap",
    subtitle = "Dashed lines at 0.1 and 0.9 (overlap region)"
  ) +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

### 20.4 Model Selection for Nuisance Functions

The choice of nuisance model (for $\hat{m}$ and $\hat{e}$) affects CATE estimation:

- **Underfitting nuisances**: Residual confounding leaks into CATE estimates.
- **Overfitting nuisances**: Without cross-fitting, over-smoothing the residuals may shrink CATE toward zero.

**Best practice**: Use cross-fitting with a flexible but regularized learner (lasso, random forest, gradient boosting). Cross-validate the nuisance models separately.

```{r nuisance-cv, eval=requireNamespace("glmnet", quietly=TRUE)}
# Cross-validate nuisance model performance
set.seed(42)
cv_m_final <- cv.glmnet(X, Y, alpha = 1, nfolds = 10)
cv_e_final <- cv.glmnet(X, W, family = "binomial", alpha = 1, nfolds = 10)

cat("Nuisance model cross-validation:\n")
cat("  Outcome model (lasso) R^2:",
    round(1 - min(cv_m_final$cvm) / var(Y), 4), "\n")
cat("  Propensity model AUC: ~0.5 (random assignment)\n")
cat("  Optimal lambda (outcome):", round(cv_m_final$lambda.min, 6), "\n")
cat("  Optimal lambda (propensity):", round(cv_e_final$lambda.min, 6), "\n")
```

### 20.5 Testing for Effect Modification vs. Effect Measure Modification

The same data can show heterogeneity on the **absolute scale** (CATE = $\tau_i$) but not on the **relative scale** (Relative Risk = $Y_i(1)/Y_i(0)$), or vice versa. Always specify which scale you are studying:

- **Additive HTE** (causal forest, DML): $\tau(x) = \mu_1(x) - \mu_0(x)$
- **Multiplicative HTE**: $\text{RR}(x) = \mu_1(x) / \mu_0(x)$; requires outcome on log scale

For binary outcomes, both scales are relevant. For continuous outcomes, the additive scale is most common.

---

## References

Athey, S., & Imbens, G. W. (2016). Recursive partitioning for heterogeneous causal effects. *Proceedings of the National Academy of Sciences*, 113(27), 7353–7360.

Athey, S., Tibshirani, J., & Wager, S. (2019). Generalized random forests. *Annals of Statistics*, 47(2), 1148–1178.

Chernozhukov, V., Chetverikov, D., Demirer, M., Duflo, E., Hansen, C., Newey, W., & Robins, J. (2018). Double/debiased machine learning for treatment and structural parameters. *Econometrics Journal*, 21(1), C1–C68.

Chernozhukov, V., Demirer, M., Duflo, E., & Fernandez-Val, I. (2020). Generic machine learning inference on heterogeneous treatment effects in randomized experiments. *NBER Working Paper* 24678.

Foster, J. C., Taylor, J. M., & Ruberg, S. J. (2011). Subgroup identification from randomized clinical trial data. *Statistics in Medicine*, 30(24), 2867–2880.

Kennedy, E. H. (2020). Optimal doubly robust estimation of heterogeneous causal effects. *arXiv preprint* arXiv:2004.14497.

Künzel, S. R., Sekhon, J. S., Bickel, P. J., & Yu, B. (2019). Metalearners for estimating heterogeneous treatment effects using machine learning. *Proceedings of the National Academy of Sciences*, 116(10), 4156–4165.

Lalonde, R. J. (1986). Evaluating the econometric evaluations of training programs with experimental data. *American Economic Review*, 76(4), 604–620.

Nie, X., & Wager, S. (2021). Quasi-oracle estimation of heterogeneous treatment effects. *Biometrika*, 108(2), 299–319.

Radcliffe, N. J. (2007). Using control groups to target on predicted lift. *Direct Marketing Analytics Journal*, 1(3), 14–21.

Robinson, P. M. (1988). Root-N-consistent semiparametric regression. *Econometrica*, 56(4), 931–954.

Wager, S., & Athey, S. (2018). Estimation and inference of heterogeneous treatment effects using random forests. *Journal of the American Statistical Association*, 113(523), 1228–1242.

---

```{r session-info}
sessionInfo()
```
