---
title: "11. Causal Mediation Analysis"
author: "Mike Nguyen"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{11. Causal Mediation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = TRUE, warning = FALSE, message = FALSE)
```

```{r setup}
library(causalverse)
library(dplyr)
library(ggplot2)
```

## Introduction

Mediation analysis asks: **through what mechanism does a treatment affect an outcome?** Rather than simply estimating *whether* a treatment $T$ affects an outcome $Y$, mediation analysis decomposes the total effect into:

- The **indirect effect** (or mediated effect): the portion of the treatment effect that operates through a mediator $M$
- The **direct effect**: the portion of the treatment effect that does not pass through $M$

This decomposition is central to mechanism research in economics, psychology, epidemiology, marketing, and virtually every empirical social science discipline. Understanding mechanisms is essential for policy design (which pathway to target?), theory testing (does the data support the proposed mechanism?), and effect modification (does the pathway vary across groups?).

**The path diagram:**

```
T ────────────────────────────────────────► Y
│                                           ▲
│         (path a)          (path b)        │
└──────────► M ──────────────────────────────┘
                                    (path c')
```

- Path $a$: Effect of $T$ on $M$
- Path $b$: Effect of $M$ on $Y$ (controlling for $T$)
- Path $c'$: Direct effect of $T$ on $Y$ (controlling for $M$)
- Total effect $c = c' + a \times b$

### Key Estimands

Let $T \in \{0, 1\}$ be a binary treatment, $M$ a mediator, $Y$ an outcome, and $X$ a vector of baseline covariates. Define:

| Estimand | Symbol | Definition |
|---|---|---|
| Average Causal Mediation Effect | ACME / $\delta(t)$ | $E[Y_i(t, M_i(1)) - Y_i(t, M_i(0))]$ |
| Average Direct Effect | ADE / $\zeta(t)$ | $E[Y_i(1, M_i(t)) - Y_i(0, M_i(t))]$ |
| Total Effect | TE / $\tau$ | $E[Y_i(1, M_i(1)) - Y_i(0, M_i(0))]$ |
| Proportion Mediated | PM | $\delta(t) / \tau$ |

Note that $\tau = \delta(t) + \zeta(1-t)$ for $t \in \{0,1\}$. When treatment-mediator interaction is absent, $\delta(0) = \delta(1)$ and $\zeta(0) = \zeta(1)$, simplifying to the familiar Baron-Kenny decomposition.

### Why Mediation Is Hard

Mediation analysis faces two fundamental challenges:

1. **Post-treatment bias**: $M$ is a post-treatment variable. Conditioning on $M$ when estimating the direct effect $c'$ can introduce collider bias or amplify unmeasured confounding, even when $T$ is randomized.

2. **Mediator-outcome confounding**: Even if $T$ is randomized, there may be unmeasured variables that confound the $M \to Y$ relationship. Unlike treatment-outcome confounding (which an experiment solves), mediator-outcome confounding is never eliminated by randomizing $T$.

These challenges mean that the classical Baron-Kenny (1986) approach  -  while useful for description  -  does not carry a causal interpretation without strong untestable assumptions.

---

## 1. The Potential Outcomes Framework for Mediation

The modern, counterfactual approach to mediation (Rubin 2004; Robins and Greenland 1992; Pearl 2001; Imai et al. 2010) defines mediation effects in terms of **potential outcomes**.

### Notation

For unit $i$:
- $Y_i(t, m)$: potential outcome under treatment $T = t$ and mediator value $M = m$
- $M_i(t)$: potential mediator value under treatment $T = t$
- The observed outcome is $Y_i = Y_i(T_i, M_i(T_i))$

### Formal Definitions

The **Average Causal Mediation Effect** (ACME) at treatment level $t$:

$$\delta(t) = E[Y_i(t, M_i(1)) - Y_i(t, M_i(0))], \quad t \in \{0, 1\}$$

This is the expected change in $Y$ when we change $M$ from the value it would take under control ($M_i(0)$) to the value it would take under treatment ($M_i(1)$), while holding $T$ fixed at $t$. It captures the mechanism through $M$.

The **Average Direct Effect** (ADE) at treatment level $t$:

$$\zeta(t) = E[Y_i(1, M_i(t)) - Y_i(0, M_i(t))], \quad t \in \{0, 1\}$$

This is the expected change in $Y$ from changing $T$ from 0 to 1, while holding $M$ fixed at the value it would take under treatment level $t$.

The **Total Effect**:

$$\tau = E[Y_i(1, M_i(1)) - Y_i(0, M_i(0))] = \delta(t) + \zeta(1-t)$$

### Sequential Ignorability

Identification of ACME requires the **Sequential Ignorability** assumption (Imai et al. 2010):

**Assumption SI.1** (No unmeasured treatment-outcome confounding):
$$\{Y_i(t', m), M_i(t)\} \perp\!\!\!\perp T_i \mid X_i = x$$

**Assumption SI.2** (No unmeasured mediator-outcome confounding):
$$Y_i(t', m) \perp\!\!\!\perp M_i \mid T_i = t, X_i = x$$

SI.1 holds when $T$ is randomly assigned (or when conditioning on $X$ achieves ignorability). SI.2 is the strong assumption: it requires that, after conditioning on $T$ and $X$, there are no unmeasured variables that simultaneously affect $M$ and $Y$.

**SI.2 is never guaranteed by design**  -  not even by randomizing $T$. This is why sensitivity analysis (Section 5) is essential.

### Identification Formula

Under sequential ignorability, for continuous $M$ and $Y$:

$$\delta(t) = \int \int E[Y | M = m, T = t, X = x] \left[f_{M|T=1,X=x}(m) - f_{M|T=0,X=x}(m)\right] dF_{X}(x) \, dm$$

This integral representation forms the basis for the product-of-coefficients estimator (in linear models) and the simulation-based estimator in the `mediation` package.

### Simulated Example: Setup

```{r po-setup}
set.seed(2024)
n <- 1000

# Data-generating process (DGP):
#  T -> M (path a = 0.5)
#  M -> Y (path b = 0.4)
#  T -> Y directly (path c' = 0.2)
#  U: unmeasured common cause of M and Y (violates SI.2 in reality)

X <- rnorm(n)            # observed covariate
T <- rbinom(n, 1, plogis(0.3 * X))   # treatment (randomized conditional on X)
U <- rnorm(n, 0, 0.5)   # unobserved M-Y confounder

# Mediator model: M ~ T + X + U
M <- 0.5 * T + 0.3 * X + U + rnorm(n, 0, 0.7)

# Outcome model: Y ~ T + M + X + U
Y <- 0.2 * T + 0.4 * M + 0.2 * X + 0.5 * U + rnorm(n, 0, 0.5)

# True effects (analytically):
# ACME = a * b = 0.5 * 0.4 = 0.20
# ADE  = c'    = 0.20
# TE   = 0.20 + 0.20 = 0.40

med_data <- data.frame(T = T, M = M, Y = Y, X = X)

cat("True ACME:", 0.5 * 0.4, "\n")
cat("True ADE: ", 0.2, "\n")
cat("True TE:  ", 0.5 * 0.4 + 0.2, "\n")
```

---

## 2. Baron-Kenny Steps (Classical Approach)

Baron and Kenny (1986) proposed a four-step procedure to establish mediation. Despite its widespread use, it has well-known limitations  -  particularly the absence of a formal causal interpretation.

### Step-by-Step Implementation

```{r bk-steps}
# Step 1: Regress Y on T (establish total effect, path c)
step1 <- lm(Y ~ T + X, data = med_data)
c_total <- coef(step1)["T"]
cat("Step 1 - Total effect (path c):", round(c_total, 4), "\n")

# Step 2: Regress M on T (establish path a)
step2 <- lm(M ~ T + X, data = med_data)
a_coef <- coef(step2)["T"]
cat("Step 2 - Path a (T -> M):", round(a_coef, 4), "\n")

# Step 3: Regress Y on T and M (establish paths b and c')
step3 <- lm(Y ~ T + M + X, data = med_data)
b_coef  <- coef(step3)["M"]
cp_coef <- coef(step3)["T"]
cat("Step 3 - Path b (M -> Y):", round(b_coef, 4), "\n")
cat("Step 3 - Path c' (direct):", round(cp_coef, 4), "\n")

# Indirect effect: a * b
indirect <- a_coef * b_coef
cat("\nIndirect effect (a*b):", round(indirect, 4), "\n")
cat("Direct effect (c')   :", round(cp_coef, 4), "\n")
cat("Total (a*b + c')     :", round(indirect + cp_coef, 4), "\n")
cat("Proportion mediated  :", round(indirect / (indirect + cp_coef), 3), "\n")
```

### Sobel Test

The Sobel (1982) test for the significance of the indirect effect uses the delta method:

$$\text{SE}_{a \times b} = \sqrt{b^2 \, \text{Var}(a) + a^2 \, \text{Var}(b)}$$

$$z_{\text{Sobel}} = \frac{a \times b}{\text{SE}_{a \times b}}$$

```{r sobel-test}
# Standard errors from the regression models
se_a <- summary(step2)$coefficients["T", "Std. Error"]
se_b <- summary(step3)$coefficients["M", "Std. Error"]

# Sobel standard error (delta method)
se_sobel <- sqrt(b_coef^2 * se_a^2 + a_coef^2 * se_b^2)

# Sobel z-statistic
z_sobel <- indirect / se_sobel
p_sobel <- 2 * pnorm(-abs(z_sobel))

cat("Sobel test:\n")
cat("  Indirect effect:", round(indirect, 4), "\n")
cat("  SE (Sobel)     :", round(se_sobel, 4), "\n")
cat("  z-statistic    :", round(z_sobel, 4), "\n")
cat("  p-value        :", round(p_sobel, 4), "\n")
```

### Bootstrap Sobel Test

The Sobel test assumes normality of the indirect effect. Bootstrap methods are more appropriate when this assumption is questionable (as is common when path coefficients are small or sample sizes are moderate).

```{r boot-sobel}
set.seed(42)
n_boot <- 2000
boot_indirect <- numeric(n_boot)

for (i in seq_len(n_boot)) {
  idx   <- sample(nrow(med_data), replace = TRUE)
  bd    <- med_data[idx, ]

  m2    <- lm(M ~ T + X, data = bd)
  m3    <- lm(Y ~ T + M + X, data = bd)

  boot_indirect[i] <- coef(m2)["T"] * coef(m3)["M"]
}

# Percentile bootstrap CI
ci_boot <- quantile(boot_indirect, c(0.025, 0.975))

cat("Bootstrap 95% CI for indirect effect:\n")
cat("  Lower:", round(ci_boot[1], 4), "\n")
cat("  Upper:", round(ci_boot[2], 4), "\n")
cat("  Bootstrap SE:", round(sd(boot_indirect), 4), "\n")

# Visualize bootstrap distribution
ggplot(data.frame(indirect = boot_indirect), aes(x = indirect)) +
  geom_histogram(bins = 50, fill = "#4393C3", color = "white", alpha = 0.8) +
  geom_vline(xintercept = ci_boot, color = "#D73027", linetype = "dashed", linewidth = 1) +
  geom_vline(xintercept = indirect, color = "#2CA02C", linetype = "solid", linewidth = 1.2) +
  labs(
    title    = "Bootstrap Distribution of Indirect Effect",
    subtitle = paste0("95% CI: [", round(ci_boot[1], 3), ", ", round(ci_boot[2], 3), "]"),
    x        = "Indirect Effect (a × b)",
    y        = "Frequency",
    caption  = "Green line = point estimate; red dashed lines = 95% CI"
  ) +
  ama_theme()
```

### Limitations of Baron-Kenny

The Baron-Kenny approach has several critical limitations:

1. **No causal identification**: The product $a \times b$ estimates the ACME only if sequential ignorability holds  -  an assumption that BK does not require researchers to state or test.

2. **Sobel test is underpowered**: The distribution of $a \times b$ is generally non-normal, especially when $a$ or $b$ are small, making the Sobel z-test conservative.

3. **No sensitivity analysis**: BK provides no tools to ask "how sensitive is this to unmeasured confounding?"

4. **Step-4 criterion is misleading**: BK's requirement that $c' < c$ (partial mediation) or $c' \approx 0$ (full mediation) is not necessary for mediation  -  suppressor effects can produce $|c'| > |c|$.

5. **No nonparametric identification**: BK relies on linear models and does not extend to binary outcomes without modification.

---

## 3. Imai et al. (2010) Causal Mediation

The `mediation` package (Imai, Keele, and Tingley 2010) implements a general framework for causal mediation analysis based on the potential outcomes approach. Unlike Baron-Kenny, it:

- Explicitly invokes and tests sequential ignorability
- Allows nonlinear models (logistic, Poisson, etc.)
- Provides quasi-Bayesian simulation-based confidence intervals
- Includes formal sensitivity analysis for violations of SI.2

### Using `causalverse::mediation_analysis()`

The `causalverse` wrapper provides a convenient interface:

```{r causalverse-mediation}
set.seed(42)
result <- mediation_analysis(
  data      = med_data,
  outcome   = "Y",
  treatment = "T",
  mediator  = "M",
  covariates = "X",
  sims      = 1000,
  seed      = 42
)

# Print summary table
print(result$summary_df)

cat("\nProportion mediated:", round(result$prop_mediated * 100, 1), "%\n")
```

```{r causalverse-mediation-plot, fig.width=7, fig.height=5}
result$plot + ama_theme() +
  labs(title = "Causal Mediation Decomposition",
       subtitle = sprintf("Proportion mediated: %.1f%%", result$prop_mediated * 100))
```

### Direct Use of `mediation` Package

```{r mediation-pkg, eval=requireNamespace("mediation", quietly=TRUE)}
library(mediation)

# Fit the two-equation system
med_fit <- lm(M ~ T + X, data = med_data)
out_fit <- lm(Y ~ T + M + X, data = med_data)

# Run mediation analysis
set.seed(42)
med_out <- mediation::mediate(
  model.m  = med_fit,
  model.y  = out_fit,
  treat    = "T",
  mediator = "M",
  sims     = 1000
)

summary(med_out)
```

```{r mediation-plot, eval=requireNamespace("mediation", quietly=TRUE), fig.width=7, fig.height=5}
plot(med_out, main = "Causal Mediation Analysis (Imai et al.)")
```

### Quasi-Bayesian vs. Bootstrap Confidence Intervals

The `mediation` package offers two simulation approaches:

| Method | `boot` argument | Description |
|---|---|---|
| Quasi-Bayesian | `FALSE` (default) | Draws from the asymptotic normal distribution of model parameters; faster |
| Nonparametric Bootstrap | `TRUE` | Resamples data with replacement; slower but more robust |

```{r mediation-boot, eval=requireNamespace("mediation", quietly=TRUE)}
# Bootstrap-based CIs (slower but robust)
set.seed(42)
med_boot <- mediation::mediate(
  model.m  = med_fit,
  model.y  = out_fit,
  treat    = "T",
  mediator = "M",
  boot     = TRUE,
  sims     = 500   # use 1000+ in practice
)

# Compare ACME CIs
s_qb   <- summary(med_out)
s_boot <- summary(med_boot)

compare_df <- data.frame(
  Method     = c("Quasi-Bayesian", "Bootstrap"),
  ACME_est   = c(s_qb$d.avg, s_boot$d.avg),
  ACME_lower = c(s_qb$d.avg.ci[1], s_boot$d.avg.ci[1]),
  ACME_upper = c(s_qb$d.avg.ci[2], s_boot$d.avg.ci[2])
)

print(compare_df)
```

---

## 4. Sensitivity Analysis for Mediation

Even under treatment randomization, sequential ignorability can fail because of unmeasured mediator-outcome confounders. The sensitivity analysis in Imai et al. (2010) asks: **how large must the correlation between the mediator-residual and the outcome-residual ($\rho$) be to overturn the ACME estimate?**

### Sensitivity Parameter $\rho$

Define $\rho$ as the correlation between the error terms of the mediator model and the outcome model:

$$\text{Corr}(\varepsilon_M, \varepsilon_Y) = \rho$$

A non-zero $\rho$ implies a common unmeasured cause of $M$ and $Y$, violating SI.2. Under sequential ignorability $\rho = 0$.

The sensitivity analysis traces ACME as a function of $\rho \in [-1, 1]$ and finds the $\rho^*$ at which ACME $= 0$.

### Running Sensitivity Analysis

```{r sens-analysis, eval=requireNamespace("mediation", quietly=TRUE)}
# Sensitivity analysis
sens_out <- mediation::medsens(
  med_out,
  rho.by = 0.05,
  effect.type = "indirect",
  sims = 100
)

summary(sens_out)
```

```{r sens-plot, eval=requireNamespace("mediation", quietly=TRUE), fig.width=7, fig.height=5}
plot(sens_out,
     main    = "Sensitivity Analysis: ACME vs. Error Correlation (ρ)",
     xlab    = "Sensitivity Parameter ρ",
     ylab    = "Estimated ACME",
     ylim    = c(-0.3, 0.5))
abline(h = 0, lty = 2, col = "red")
```

### Interpreting Sensitivity Results

```{r sens-interpret, eval=requireNamespace("mediation", quietly=TRUE)}
# Find the critical rho (where ACME crosses zero)
# Extract from sensitivity object
rho_vals <- sens_out$rho
acme_vals <- sens_out$d0

# Find crossing
if (any(acme_vals > 0) && any(acme_vals < 0)) {
  cross_idx <- which(diff(sign(acme_vals)) != 0)[1]
  rho_star  <- rho_vals[cross_idx]
  cat("Critical rho (ACME = 0):", round(rho_star, 3), "\n")
  cat("Interpretation: The ACME remains positive as long as the error\n")
  cat("correlation between mediator and outcome models is below", round(rho_star, 3), "\n")
} else {
  cat("ACME does not cross zero in the range of rho examined.\n")
  cat("This suggests a robust mediation effect.\n")
}
```

### $R^2$ Sensitivity Plot

The $R^2$ sensitivity analysis presents results in terms of proportions of variance explained by the unmeasured confounder, which is often more interpretable than $\rho$.

```{r sens-r2-plot, eval=requireNamespace("mediation", quietly=TRUE), fig.width=7, fig.height=5}
plot(sens_out,
     sens.par = "R2",
     main     = "Sensitivity: Proportion of Variance Explained",
     r.type   = "total",
     sign.prod = "positive")
```

### E-Values for Mediation

The E-value (VanderWeele and Ding 2017) quantifies the minimum strength of unmeasured confounding needed to explain away the mediated effect. For a risk ratio scale:

$$E\text{-value} = \text{RR} + \sqrt{\text{RR} \times (\text{RR} - 1)}$$

For linear models, we can convert effect sizes to approximate relative risks:

```{r evalue-mediation}
# Manual E-value calculation for mediation
# Effect size: Cohen's d for indirect effect
acme_est <- result$acme
sd_Y     <- sd(med_data$Y)
d_acme   <- acme_est / sd_Y   # Standardized indirect effect

# Approximate RR (for small effects: RR ≈ exp(ln(OR)) or use z-to-RR approximation)
# Using approximation: d ≈ ln(OR)/1.81, then OR to RR
# For simplicity, we use: E-value for standardized mean difference
# E-value = exp(0.91 * sqrt(d^2 + 4)) + exp(0.91 * sqrt(d^2 + 4)) - 1  [Sjolander 2020]
# Simpler: for continuous outcomes, evaluate using the unstandardized scale

# E-value based on ratio metric (approximate)
# Point estimate E-value: RR = exp(0.91 * d) [common approximation]
RR_approx <- exp(0.91 * abs(d_acme))
Evalue    <- RR_approx + sqrt(RR_approx * (RR_approx - 1))

cat("Standardized ACME (Cohen's d):", round(d_acme, 3), "\n")
cat("Approximate RR:               ", round(RR_approx, 3), "\n")
cat("E-value for ACME:             ", round(Evalue, 3), "\n")
cat("\nInterpretation: An unmeasured confounder would need to be associated\n")
cat("with both the mediator and outcome by a risk ratio of at least",
    round(Evalue, 2), "to\nexplain away the observed mediated effect.\n")
```

---

## 5. `causalverse::med_ind()`  -  Monte Carlo Indirect Effects

The `med_ind()` function in `causalverse` implements the Monte Carlo method (Selig and Preacher 2008) for estimating indirect effects. Rather than relying on the Sobel normal approximation, it simulates the distribution of $a \times b$ directly from the joint sampling distribution of path coefficients.

### How It Works

Given:
- Path coefficient $a$ with variance $\text{Var}(a)$
- Path coefficient $b$ with variance $\text{Var}(b)$
- Covariance $\text{Cov}(a, b)$

The function draws $(a^*, b^*)$ pairs from a bivariate normal distribution and computes the distribution of $a^* \times b^*$.

### Basic Usage

```{r med-ind-basic, fig.width=7, fig.height=5}
# Extract path coefficients and their variances from our models
a_val   <- coef(step2)["T"]
b_val   <- coef(step3)["M"]
var_a   <- vcov(step2)["T", "T"]
var_b   <- vcov(step3)["M", "M"]

# Covariance between a and b is typically 0 (different regression models)
# unless we use a system estimator; here we set it to 0
cov_ab  <- 0

# Run Monte Carlo simulation
mc_result <- med_ind(
  a          = a_val,
  b          = b_val,
  var_a      = var_a,
  var_b      = var_b,
  cov_ab     = cov_ab,
  ci         = 95,
  iterations = 20000,
  seed       = 42
)

cat("Monte Carlo 95% CI for indirect effect:\n")
cat("  Lower:", round(mc_result$lower_quantile, 4), "\n")
cat("  Upper:", round(mc_result$upper_quantile, 4), "\n")
cat("  Point estimate (a*b):", round(a_val * b_val, 4), "\n")

# Display the distribution plot
mc_result$plot + ama_theme()
```

### Comparison of Methods

```{r method-comparison}
comparison <- data.frame(
  Method    = c("Sobel (Delta)", "Bootstrap", "Monte Carlo"),
  Estimate  = rep(round(a_val * b_val, 4), 3),
  Lower_95  = c(
    round(a_val * b_val - 1.96 * se_sobel, 4),
    round(ci_boot[1], 4),
    round(mc_result$lower_quantile, 4)
  ),
  Upper_95  = c(
    round(a_val * b_val + 1.96 * se_sobel, 4),
    round(ci_boot[2], 4),
    round(mc_result$upper_quantile, 4)
  )
)

print(comparison)
```

### When to Use Monte Carlo vs. Bootstrap

| Scenario | Recommendation |
|---|---|
| Small samples (n < 100) | Bootstrap (more accurate in tails) |
| Moderate samples (100–500) | Monte Carlo or bootstrap (similar) |
| Large samples (n > 500) | Monte Carlo (faster, asymptotically equivalent) |
| Non-normal path coefficients | Bootstrap |
| Complex models (SEM, MLM) | Monte Carlo (easier to implement) |
| Time constraints | Monte Carlo |

---

## 6. Multiple Mediators

Real-world mechanisms often involve multiple pathways. Researchers must distinguish between:

- **Parallel mediation**: Each mediator operates independently from $T$
- **Sequential mediation**: Mediators are in a causal chain ($T \to M_1 \to M_2 \to Y$)

### Parallel Mediation

```{r parallel-mediation}
set.seed(123)
n <- 500

# DGP: T -> M1 -> Y (indirect1 = 0.3 * 0.4 = 0.12)
#      T -> M2 -> Y (indirect2 = 0.5 * 0.3 = 0.15)
#      T -> Y  (direct = 0.1)

T2  <- rbinom(n, 1, 0.5)
M1  <- 0.3 * T2 + rnorm(n, 0, 0.8)
M2  <- 0.5 * T2 + rnorm(n, 0, 0.8)
Y2  <- 0.4 * M1 + 0.3 * M2 + 0.1 * T2 + rnorm(n, 0, 0.8)

df_par <- data.frame(T = T2, M1 = M1, M2 = M2, Y = Y2)

# Estimate each indirect effect separately
# M1 pathway
m_m1   <- lm(M1 ~ T, data = df_par)
m_y_m1 <- lm(Y ~ T + M1 + M2, data = df_par)

a1  <- coef(m_m1)["T"]
b1  <- coef(m_y_m1)["M1"]
ie1 <- a1 * b1

# M2 pathway
m_m2   <- lm(M2 ~ T, data = df_par)
a2  <- coef(m_m2)["T"]
b2  <- coef(m_y_m1)["M2"]
ie2 <- a2 * b2

# Total indirect and direct
cp  <- coef(m_y_m1)["T"]
tot <- ie1 + ie2 + cp

cat("Parallel Mediation Results:\n")
cat("  Indirect via M1:", round(ie1, 4), "(true: 0.12)\n")
cat("  Indirect via M2:", round(ie2, 4), "(true: 0.15)\n")
cat("  Direct effect:  ", round(cp, 4), "(true: 0.10)\n")
cat("  Total effect:   ", round(tot, 4), "(true: 0.37)\n")
cat("  Prop. via M1:   ", round(ie1/tot, 3), "\n")
cat("  Prop. via M2:   ", round(ie2/tot, 3), "\n")
```

### Sequential Mediation

```{r sequential-mediation}
set.seed(456)
n <- 500

# DGP: T -> M1 -> M2 -> Y
# T -> M1: a1 = 0.6
# M1 -> M2: d21 = 0.5
# M2 -> Y: b2 = 0.4
# T -> M2 direct: a2 = 0.1 (spurious)
# T -> Y direct: cp = 0.15

T3  <- rbinom(n, 1, 0.5)
M1s <- 0.6 * T3 + rnorm(n, 0, 0.8)
M2s <- 0.5 * M1s + 0.1 * T3 + rnorm(n, 0, 0.7)
Y3  <- 0.4 * M2s + 0.15 * T3 + rnorm(n, 0, 0.7)

df_seq <- data.frame(T = T3, M1 = M1s, M2 = M2s, Y = Y3)

# Sequential paths
sm1   <- lm(M1 ~ T, data = df_seq)
sm2   <- lm(M2 ~ T + M1, data = df_seq)
sy    <- lm(Y ~ T + M1 + M2, data = df_seq)

a1s   <- coef(sm1)["T"]          # T -> M1
d21s  <- coef(sm2)["M1"]         # M1 -> M2
b2s   <- coef(sy)["M2"]          # M2 -> Y

# Specific indirect effects:
ie_seq1  <- a1s * d21s * b2s     # T -> M1 -> M2 -> Y
ie_seq2  <- a1s * coef(sy)["M1"] # T -> M1 -> Y (if M1 also affects Y)
ie_seq3  <- coef(sm2)["T"] * b2s # T -> M2 -> Y (bypassing M1)
direct_s <- coef(sy)["T"]

cat("Sequential Mediation (T -> M1 -> M2 -> Y):\n")
cat("  T -> M1 -> M2 -> Y:", round(ie_seq1, 4), "(true:", 0.6*0.5*0.4, ")\n")
cat("  T -> M2 -> Y (direct T->M2):", round(ie_seq3, 4), "\n")
cat("  Direct T -> Y:      ", round(direct_s, 4), "(true: 0.15)\n")
```

### SEM for Multiple Mediators with `lavaan`

```{r lavaan-sem, eval=requireNamespace("lavaan", quietly=TRUE)}
library(lavaan)

# Parallel mediation SEM model
model_par <- "
  # Mediator equations
  M1 ~ a1*T
  M2 ~ a2*T

  # Outcome equation
  Y ~ cp*T + b1*M1 + b2*M2

  # Defined parameters
  indirect1 := a1*b1
  indirect2 := a2*b2
  total_indirect := a1*b1 + a2*b2
  total := cp + a1*b1 + a2*b2
  prop_m1 := a1*b1 / (cp + a1*b1 + a2*b2)
  prop_m2 := a2*b2 / (cp + a1*b1 + a2*b2)
"

fit_par <- lavaan::sem(model_par, data = df_par, se = "bootstrap",
                       bootstrap = 500, seed = 42)

# Extract parameter estimates with CIs
lavaan::parameterEstimates(
  fit_par,
  boot.ci.type = "perc",
  level = 0.95,
  output = "text"
)[c("indirect1", "indirect2", "total_indirect", "total", "prop_m1", "prop_m2"), ]
```

```{r lavaan-fit-indices, eval=requireNamespace("lavaan", quietly=TRUE)}
# Model fit indices
lavaan::fitmeasures(fit_par, c("cfi", "rmsea", "srmr", "tli", "chisq", "df", "pvalue"))
```

---

## 7. Moderated Mediation (Conditional Process Analysis)

Moderated mediation (also called conditional process analysis, Hayes 2013) asks whether the mediated effect varies as a function of a moderator $W$.

The **Index of Moderated Mediation (IMM)** tests whether the indirect effect is significantly different across levels of $W$.

### Model Specification

**First-stage moderated mediation** (PROCESS Model 7):
$$M = a_1 T + a_2 W + a_3 (T \times W) + e_M$$
$$Y = b_1 M + c_1 T + c_2 W + e_Y$$

Conditional indirect effect at $W = w$:
$$\text{IE}(w) = (a_1 + a_3 w) \times b_1$$

```{r modmed-setup}
set.seed(789)
n <- 600

# DGP: T -> M (moderated by W)
#      M -> Y (path b = 0.4)
#      T -> Y directly (c' = 0.1)
# a(W) = 0.3 + 0.4*W

W  <- rnorm(n)                             # Continuous moderator
T4 <- rbinom(n, 1, 0.5)
M4 <- (0.3 + 0.4 * W) * T4 + 0.2 * W + rnorm(n, 0, 0.7)
Y4 <- 0.4 * M4 + 0.1 * T4 + 0.1 * W + rnorm(n, 0, 0.7)

df_mm <- data.frame(T = T4, M = M4, W = W, Y = Y4)

# Fit PROCESS-style models
# First stage: M ~ T + W + T*W
m_mm_med <- lm(M ~ T * W, data = df_mm)
# Second stage: Y ~ T + M + W
m_mm_out <- lm(Y ~ T + M + W, data = df_mm)

summary(m_mm_med)
```

### Conditional Indirect Effects

```{r conditional-ie}
# Extract coefficients
a1_mm <- coef(m_mm_med)["T"]
a3_mm <- coef(m_mm_med)["T:W"]
b1_mm <- coef(m_mm_out)["M"]

# Evaluate at typical moderator values (mean - 1 SD, mean, mean + 1 SD)
w_vals <- c(mean(W) - sd(W), mean(W), mean(W) + sd(W))
w_labs <- c("-1 SD", "Mean", "+1 SD")

cie_df <- data.frame(
  W_level   = w_labs,
  W_value   = w_vals,
  path_a    = a1_mm + a3_mm * w_vals,
  indirect  = (a1_mm + a3_mm * w_vals) * b1_mm
)

print(cie_df)
```

### Bootstrap CIs for Conditional Indirect Effects

```{r modmed-boot}
set.seed(42)
n_boot2 <- 1000
boot_cie <- matrix(NA, nrow = n_boot2, ncol = 3)

for (i in seq_len(n_boot2)) {
  idx <- sample(nrow(df_mm), replace = TRUE)
  bd  <- df_mm[idx, ]

  bm  <- lm(M ~ T * W, data = bd)
  by  <- lm(Y ~ T + M + W, data = bd)

  ba1 <- coef(bm)["T"]
  ba3 <- coef(bm)["T:W"]
  bb1 <- coef(by)["M"]

  boot_cie[i, ] <- (ba1 + ba3 * w_vals) * bb1
}

# 95% percentile CIs
ci_cie <- apply(boot_cie, 2, quantile, c(0.025, 0.975))

cie_results <- data.frame(
  W_level  = w_labs,
  Estimate = cie_df$indirect,
  Lower_95 = ci_cie[1, ],
  Upper_95 = ci_cie[2, ]
)

print(cie_results)
```

### Visualization: Conditional Indirect Effects

```{r modmed-viz, fig.width=8, fig.height=5}
# Compute CIE across range of W
w_range   <- seq(min(W), max(W), length.out = 100)
n_boot3   <- 500
boot_cie2 <- matrix(NA, nrow = n_boot3, ncol = length(w_range))

set.seed(42)
for (i in seq_len(n_boot3)) {
  idx <- sample(nrow(df_mm), replace = TRUE)
  bd  <- df_mm[idx, ]
  bm  <- lm(M ~ T * W, data = bd)
  by  <- lm(Y ~ T + M + W, data = bd)
  ba1 <- coef(bm)["T"]
  ba3 <- coef(bm)["T:W"]
  bb1 <- coef(by)["M"]
  boot_cie2[i, ] <- (ba1 + ba3 * w_range) * bb1
}

cie_line <- (a1_mm + a3_mm * w_range) * b1_mm
cie_lo   <- apply(boot_cie2, 2, quantile, 0.025)
cie_hi   <- apply(boot_cie2, 2, quantile, 0.975)

plot_df <- data.frame(W = w_range, CIE = cie_line, Lower = cie_lo, Upper = cie_hi)

ggplot(plot_df, aes(x = W, y = CIE)) +
  geom_ribbon(aes(ymin = Lower, ymax = Upper), fill = "#4393C3", alpha = 0.25) +
  geom_line(linewidth = 1.2, color = "#2171B5") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#D73027") +
  geom_vline(xintercept = w_vals, linetype = "dotted", color = "gray40") +
  annotate("text", x = w_vals, y = max(cie_hi) * 1.05,
           label = w_labs, size = 3.5, hjust = 0.5, color = "gray30") +
  labs(
    title    = "Conditional Indirect Effect as a Function of Moderator W",
    subtitle = "Shaded region = 95% bootstrap CI; dashed = zero line",
    x        = "Moderator W",
    y        = "Indirect Effect",
    caption  = "PROCESS Model 7: First-stage moderated mediation"
  ) +
  ama_theme()
```

### Index of Moderated Mediation (IMM)

```{r imm-test}
# IMM = a3 * b1 (the interaction term's contribution to the indirect effect slope)
imm_est <- a3_mm * b1_mm

# Bootstrap CI for IMM
imm_boot <- numeric(n_boot2)
set.seed(42)
for (i in seq_len(n_boot2)) {
  idx <- sample(nrow(df_mm), replace = TRUE)
  bd  <- df_mm[idx, ]
  bm  <- lm(M ~ T * W, data = bd)
  by  <- lm(Y ~ T + M + W, data = bd)
  imm_boot[i] <- coef(bm)["T:W"] * coef(by)["M"]
}

imm_ci <- quantile(imm_boot, c(0.025, 0.975))

cat("Index of Moderated Mediation:\n")
cat("  IMM estimate:", round(imm_est, 4), "\n")
cat("  95% CI:      [", round(imm_ci[1], 4), ",", round(imm_ci[2], 4), "]\n")
cat("  Significant:", ifelse(imm_ci[1] > 0 | imm_ci[2] < 0, "Yes", "No"), "\n")
```

### Johnson-Neyman (Floodlight) Analysis

The Johnson-Neyman technique identifies the exact values of $W$ where the indirect effect transitions from non-significant to significant (or vice versa).

```{r jn-floodlight, fig.width=8, fig.height=5}
# Find JN points: where does the 95% CI of CIE cross zero?
cross_lo <- which(diff(sign(cie_lo)) != 0)
cross_hi <- which(diff(sign(cie_hi)) != 0)

jn_points <- numeric(0)
if (length(cross_lo) > 0) jn_points <- c(jn_points, w_range[cross_lo[1]])
if (length(cross_hi) > 0) jn_points <- c(jn_points, w_range[cross_hi[1]])

# Floodlight plot
# Regions of significance: where both CI bounds are on same side of zero
sig_region <- (cie_lo > 0 & cie_hi > 0) | (cie_lo < 0 & cie_hi < 0)

# Plot with significance regions highlighted
ggplot(plot_df, aes(x = W)) +
  # Non-significant background
  geom_ribbon(aes(ymin = Lower, ymax = Upper), fill = "gray80", alpha = 0.4) +
  # Significant regions
  geom_ribbon(data = plot_df[sig_region, ],
              aes(ymin = Lower, ymax = Upper), fill = "#4393C3", alpha = 0.4) +
  geom_line(aes(y = CIE), linewidth = 1.2, color = "#2171B5") +
  geom_line(aes(y = Lower), linetype = "dashed", color = "gray50", linewidth = 0.7) +
  geom_line(aes(y = Upper), linetype = "dashed", color = "gray50", linewidth = 0.7) +
  geom_hline(yintercept = 0, linetype = "solid", color = "#D73027", linewidth = 0.8) +
  {if (length(jn_points) > 0)
    geom_vline(xintercept = jn_points, linetype = "longdash", color = "#D73027", linewidth = 1)
  } +
  labs(
    title    = "Floodlight Analysis: Johnson-Neyman Technique",
    subtitle = "Blue shading = region where indirect effect is significant (95% CI excludes zero)",
    x        = "Moderator W",
    y        = "Conditional Indirect Effect",
    caption  = "Red vertical lines = Johnson-Neyman points of significance transition"
  ) +
  ama_theme()
```

---

## 8. Mediation with Binary Outcomes

When the outcome (or mediator) is binary, the linear mediation model is mis-specified. The `mediation` package handles this via generalized linear models.

### Logistic Mediation

```{r binary-outcome}
set.seed(101)
n <- 800

# DGP: binary outcome
T5  <- rbinom(n, 1, 0.5)
M5  <- 0.6 * T5 + rnorm(n, 0, 1)                       # Continuous mediator
p5  <- plogis(-0.5 + 0.3 * T5 + 0.5 * M5)              # Logistic outcome
Y5  <- rbinom(n, 1, p5)

df_bin <- data.frame(T = T5, M = M5, Y = Y5)

cat("Outcome prevalence:", round(mean(Y5), 3), "\n")
cat("Mediator mean (treated):  ", round(mean(M5[T5==1]), 3), "\n")
cat("Mediator mean (control):  ", round(mean(M5[T5==0]), 3), "\n")
```

```{r logistic-mediation, eval=requireNamespace("mediation", quietly=TRUE)}
# Mediation with binary outcome
med_m_bin <- lm(M ~ T, data = df_bin)         # Mediator model: OLS
med_y_bin <- glm(Y ~ T + M, data = df_bin,    # Outcome model: logistic
                 family = binomial())

set.seed(42)
med_bin_out <- mediation::mediate(
  model.m  = med_m_bin,
  model.y  = med_y_bin,
  treat    = "T",
  mediator = "M",
  sims     = 500
)

summary(med_bin_out)
```

### Interpretation with Binary Outcomes

With a logistic outcome model, the ACME is estimated on the **probability scale** (marginal effects), not the log-odds scale. This is because the potential outcomes framework works with the expected value of the outcome, which for binary $Y$ is a probability.

```{r logistic-mfx}
# Marginal effects for the logistic outcome
m_logit_full <- glm(Y ~ T + M, data = df_bin, family = binomial())

# Average Marginal Effect (AME) of M on Y
# AME_M = E[dP(Y=1)/dM] = E[b * p(1-p)]
p_hat  <- predict(m_logit_full, type = "response")
b_logit <- coef(m_logit_full)["M"]
ame_M  <- mean(b_logit * p_hat * (1 - p_hat))

cat("Coefficient on M (log-odds):", round(b_logit, 4), "\n")
cat("AME of M on P(Y=1):        ", round(ame_M, 4), "\n")
cat("(Approx. indirect effect on probability scale)\n")
```

### Binary Mediator and Binary Outcome

```{r binary-both, eval=requireNamespace("mediation", quietly=TRUE)}
set.seed(202)
n <- 800

T6  <- rbinom(n, 1, 0.5)
pM  <- plogis(-0.3 + 0.8 * T6)
M6  <- rbinom(n, 1, pM)                # Binary mediator
pY  <- plogis(-1.0 + 0.4 * T6 + 1.2 * M6)
Y6  <- rbinom(n, 1, pY)

df_bin2 <- data.frame(T = T6, M = M6, Y = Y6)

med_m_bb <- glm(M ~ T, data = df_bin2, family = binomial())
med_y_bb <- glm(Y ~ T + M, data = df_bin2, family = binomial())

set.seed(42)
med_bb <- mediation::mediate(
  model.m  = med_m_bb,
  model.y  = med_y_bb,
  treat    = "T",
  mediator = "M",
  sims     = 500
)

summary(med_bb)
```

---

## 9. Mediation with Treatment Heterogeneity

### Subgroup-Specific ACME

In many settings, the mediated effect varies across subgroups. We can estimate subgroup-specific ACMEs by running separate mediation analyses within each group.

```{r subgroup-acme}
set.seed(303)
n <- 1000

# DGP: ACME differs between high and low X subgroups
X_grp <- rbinom(n, 1, 0.5)   # Binary subgroup indicator
T7    <- rbinom(n, 1, 0.5)

# ACME(X=0) = 0.1 * 0.4 = 0.04; ACME(X=1) = 0.6 * 0.4 = 0.24
a_grp <- ifelse(X_grp == 1, 0.6, 0.1)
M7    <- a_grp * T7 + rnorm(n, 0, 0.8)
Y7    <- 0.4 * M7 + 0.2 * T7 + rnorm(n, 0, 0.8)

df_het <- data.frame(T = T7, M = M7, Y = Y7, X = X_grp)

# Subgroup analysis
results_het <- lapply(0:1, function(x) {
  sub_df <- df_het[df_het$X == x, ]
  m_med  <- lm(M ~ T, data = sub_df)
  m_out  <- lm(Y ~ T + M, data = sub_df)
  a_sub  <- coef(m_med)["T"]
  b_sub  <- coef(m_out)["M"]
  data.frame(
    Subgroup = paste0("X = ", x),
    n        = nrow(sub_df),
    Path_a   = round(a_sub, 4),
    Path_b   = round(b_sub, 4),
    ACME     = round(a_sub * b_sub, 4),
    True_ACME = round(ifelse(x == 1, 0.6, 0.1) * 0.4, 4)
  )
})

do.call(rbind, results_het)
```

### Treatment-Mediator Interaction

When treatment modifies the effect of the mediator on the outcome, the ACME and ADE differ across treatment levels ($\delta(0) \neq \delta(1)$).

```{r treat-med-interaction, eval=requireNamespace("mediation", quietly=TRUE)}
set.seed(404)
n <- 600

T8  <- rbinom(n, 1, 0.5)
M8  <- 0.5 * T8 + rnorm(n, 0, 0.8)
# T × M interaction in outcome model
Y8  <- 0.3 * T8 + 0.4 * M8 + 0.3 * T8 * M8 + rnorm(n, 0, 0.7)

df_int <- data.frame(T = T8, M = M8, Y = Y8)

med_m_int <- lm(M ~ T, data = df_int)
med_y_int <- lm(Y ~ T * M, data = df_int)   # Include interaction

set.seed(42)
med_int <- mediation::mediate(
  model.m  = med_m_int,
  model.y  = med_y_int,
  treat    = "T",
  mediator = "M",
  sims     = 500
)

# When T*M interaction is present, d0 ≠ d1
s_int <- summary(med_int)
cat("ACME (T=0): ", round(s_int$d0, 4), "\n")
cat("ACME (T=1): ", round(s_int$d1, 4), "\n")
cat("ADE  (T=0): ", round(s_int$z0, 4), "\n")
cat("ADE  (T=1): ", round(s_int$z1, 4), "\n")
cat("Total effect:", round(s_int$tau.coef, 4), "\n")
```

---

## 10. Causal Forest for Heterogeneous Mediation

Causal forests (Wager and Athey 2018) estimate heterogeneous treatment effects using random forests. For mediation, we can use a two-step approach: estimate the CATE and then decompose it into mediated and direct components.

### Estimating CATE with `grf`

```{r grf-setup, eval=requireNamespace("grf", quietly=TRUE)}
library(grf)

set.seed(505)
n <- 1000

X_grf <- matrix(rnorm(n * 5), n, 5)  # 5 covariates
T_grf <- rbinom(n, 1, 0.5)

# ACME varies with X1:
#   a(X1) = 0.3 + 0.5 * X1
# b = 0.4 (constant)
# c' = 0.1 (constant)

M_grf <- (0.3 + 0.5 * X_grf[, 1]) * T_grf + rnorm(n, 0, 0.7)
Y_grf <- 0.4 * M_grf + 0.1 * T_grf + 0.2 * X_grf[, 1] + rnorm(n, 0, 0.6)

df_grf <- data.frame(Y = Y_grf, T = T_grf, M = M_grf, X_grf)
```

```{r grf-cate, eval=requireNamespace("grf", quietly=TRUE)}
# Step 1: Estimate CATE using causal forest
cf <- grf::causal_forest(
  X = X_grf,
  Y = Y_grf,
  W = T_grf,
  seed = 42
)

# CATE estimates
cate_hat <- predict(cf)$predictions

cat("CATE summary:\n")
print(summary(cate_hat))

# Average treatment effect
ate_grf <- average_treatment_effect(cf, target.sample = "all")
cat("\nATE:", round(ate_grf["estimate"], 4),
    "SE:", round(ate_grf["std.err"], 4), "\n")
```

```{r grf-heterogeneity, eval=requireNamespace("grf", quietly=TRUE), fig.width=7, fig.height=5}
# Visualize CATE vs X1 (the driver of heterogeneity)
ggplot(data.frame(X1 = X_grf[, 1], CATE = cate_hat), aes(x = X1, y = CATE)) +
  geom_point(alpha = 0.3, size = 0.8, color = "#4393C3") +
  geom_smooth(method = "loess", se = TRUE, color = "#D73027", linewidth = 1.2) +
  geom_hline(yintercept = ate_grf["estimate"], linetype = "dashed", color = "gray40") +
  labs(
    title    = "Heterogeneous Treatment Effects by X1",
    subtitle = "CATE estimated via causal forest; red = local regression; dashed = ATE",
    x        = "Covariate X1",
    y        = "Estimated CATE"
  ) +
  ama_theme()
```

### Decomposing CATE into Mediated vs. Direct Components

```{r grf-decompose, eval=requireNamespace("grf", quietly=TRUE)}
# Step 2: Estimate the mediator effect of T on M
cf_m <- grf::causal_forest(
  X = X_grf,
  Y = M_grf,
  W = T_grf,
  seed = 42
)

cate_m_hat <- predict(cf_m)$predictions   # a(x): T -> M

# Step 3: Estimate effect of M on Y (assuming linear, controlling for T and X)
# We use a simple approach: regress Y on T, M, X, then extract b
lm_b <- lm(Y_grf ~ T_grf + M_grf + X_grf)
b_hat <- coef(lm_b)["M_grf"]

# Estimated ACME(x) = a(x) * b
acme_hat <- cate_m_hat * b_hat

# Estimated ADE(x) = CATE(x) - ACME(x)
ade_hat  <- cate_hat - acme_hat

cat("Heterogeneous ACME via causal forest:\n")
cat("  Mean ACME:", round(mean(acme_hat), 4), "\n")
cat("  SD ACME:  ", round(sd(acme_hat), 4), "\n")
cat("  Mean ADE: ", round(mean(ade_hat), 4), "\n")

# Correlation of ACME with X1
cat("  Cor(ACME, X1):", round(cor(acme_hat, X_grf[,1]), 4),
    "(true: positive, since a(X1) increases in X1)\n")
```

```{r grf-decomp-plot, eval=requireNamespace("grf", quietly=TRUE), fig.width=8, fig.height=5}
ggplot(
  data.frame(X1 = X_grf[, 1], ACME = acme_hat, ADE = ade_hat, CATE = cate_hat),
  aes(x = X1)
) +
  geom_point(aes(y = ACME, color = "ACME"), alpha = 0.3, size = 0.8) +
  geom_point(aes(y = ADE,  color = "ADE"),  alpha = 0.3, size = 0.8) +
  geom_smooth(aes(y = ACME, color = "ACME"), method = "loess", se = FALSE, linewidth = 1.2) +
  geom_smooth(aes(y = ADE,  color = "ADE"),  method = "loess", se = FALSE, linewidth = 1.2) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  scale_color_manual(values = c("ACME" = "#4393C3", "ADE" = "#D73027"),
                     name = "Effect") +
  labs(
    title    = "Decomposition of CATE into Mediated and Direct Effects",
    subtitle = "ACME increases with X1 (as designed); ADE is approximately constant",
    x        = "Covariate X1",
    y        = "Effect Estimate"
  ) +
  ama_theme()
```

---

## 11. Natural Direct and Indirect Effects

Pearl (2001) and Robins and Greenland (1992) defined the **Natural Direct Effect** (NDE) and **Natural Indirect Effect** (NIE) using nested counterfactuals.

### Definitions

**Pure Natural Direct Effect** (PNDE):
$$\text{PNDE} = E[Y(1, M(0)) - Y(0, M(0))]$$
The effect of $T$ on $Y$, with $M$ set to the value it would have under control.

**Total Natural Indirect Effect** (TNIE):
$$\text{TNIE} = E[Y(1, M(1)) - Y(1, M(0))]$$
The change in $Y$ when we shift $M$ from its control-level value to its treated-level value, while holding $T = 1$.

**Total Effect decomposition:**
$$\tau = \text{PNDE} + \text{TNIE}$$

Note: The TNIE under this notation equals the ACME under Imai et al.'s notation (with $t = 1$).

### VanderWeele's Four-Way Decomposition

VanderWeele (2014) proposed decomposing the total effect into four components:

$$\tau = \underbrace{\text{CDE}}_{\text{controlled direct}} + \underbrace{\text{intref}}_{\text{interaction ref.}} + \underbrace{\text{intmed}}_{\text{interaction med.}} + \underbrace{\text{pie}}_{\text{pure indirect}}$$

- **CDE** ($c_{TE}^{11} - c_{TE}^{00}$ with $M = m_0$): Controlled direct effect
- **intref**: Interaction referent (excess from synergy, attributable to baseline)
- **intmed**: Interaction mediated (excess from synergy, attributable to mediation)
- **pie**: Pure indirect effect (mediation with no interaction)

This decomposition is especially informative when treatment and mediator interact.

### Implementation with Linear Models

```{r fourway-decomp}
set.seed(606)
n <- 800

T9  <- rbinom(n, 1, 0.5)
M9  <- 0.5 * T9 + rnorm(n, 0, 0.9)
m0  <- 0   # Reference level for M (e.g., mean under control)

# DGP with T*M interaction
Y9  <- 0.2 * T9 + 0.3 * M9 + 0.4 * T9 * M9 + rnorm(n, 0, 0.8)

df_fw <- data.frame(T = T9, M = M9, Y = Y9)

# Fit models
m_fw_med <- lm(M ~ T, data = df_fw)
m_fw_out <- lm(Y ~ T * M, data = df_fw)

# Extract coefficients
beta_T  <- coef(m_fw_out)["T"]        # c1
beta_M  <- coef(m_fw_out)["M"]        # b
beta_TM <- coef(m_fw_out)["T:M"]      # theta (interaction)
theta_a <- coef(m_fw_med)["T"]        # a (T -> M)

# Four-way decomposition at m0 = mean(M[T==0])
m0_ref <- mean(M9[T9 == 0])

# CDE: direct effect when M = m0
cde  <- beta_T + beta_TM * m0_ref

# Reference interaction: theta * (E[M|T=1] - m0) - a * theta
# Actually using VanderWeele 2014 formulas for linear interaction:
# intref = theta_TM * theta_a * (1 - delta_t)...
# For binary T:
intref <- beta_TM * m0_ref * theta_a
intmed <- beta_TM * theta_a^2
pie    <- beta_M * theta_a

# Total indirect
total_indirect <- pie + intmed

cat("VanderWeele Four-Way Decomposition:\n")
cat("  CDE (controlled direct):  ", round(cde, 4), "\n")
cat("  intref (interaction ref): ", round(intref, 4), "\n")
cat("  intmed (interaction med): ", round(intmed, 4), "\n")
cat("  pie (pure indirect):      ", round(pie, 4), "\n")
cat("  Total (sum):              ", round(cde + intref + intmed + pie, 4), "\n")
cat("\n  For comparison, OLS total effect:\n")
cat("  lm(Y~T) coef:", round(coef(lm(Y ~ T, data = df_fw))["T"], 4), "\n")
```

### CMAverse Package (if available)

```{r cmaverse-demo, eval=requireNamespace("CMAverse", quietly=TRUE)}
library(CMAverse)

# CMAverse implements multiple mediation frameworks
# including VanderWeele's four-way decomposition
cmares <- CMAverse::cmest(
  data     = df_fw,
  model    = "rb",          # Regression-based approach
  outcome  = "Y",
  exposure = "T",
  mediator = list("M"),
  EMint    = TRUE,           # Allow exposure-mediator interaction
  mreg     = list("linear"),
  yreg     = "linear",
  astar    = 0,              # Reference level of exposure
  a        = 1,              # Comparison level of exposure
  mval     = list(m0_ref),   # Reference mediator value
  estimation = "paramfunc",
  inference  = "bootstrap",
  nboot      = 200,
  seed       = 42
)

summary(cmares)
```

---

## 12. Difference-in-Differences Mediation

DiD designs can be extended to mediation when the mediator is also observed in pre/post periods. The key challenge is that DiD-style arguments about parallel trends must apply to both the mediator and the outcome.

### Setup: Panel Data Mediation

```{r did-mediation}
set.seed(707)
n   <- 200   # units
T_n <- 2     # pre/post

# Panel structure
unit <- rep(1:n, each = T_n)
time <- rep(0:1, times = n)
treated <- rep(rbinom(n, 1, 0.5), each = T_n)

# Unit fixed effects
unit_fe <- rep(rnorm(n, 0, 1), each = T_n)

# Treatment indicator: treated & post
D <- as.integer(treated == 1 & time == 1)

# Mediator: M affected by D (with parallel trends)
M_pre_effect <- 0
M_post_treat <- 0.6
M_val <- M_pre_effect + M_post_treat * D + unit_fe + rnorm(n * T_n, 0, 0.5)

# Outcome: Y affected by D directly and through M
Y_pre_effect <- 0
Y_post_treat <- 0.2   # direct effect
b_M_Y        <- 0.4   # M -> Y
Y_val <- Y_pre_effect + Y_post_treat * D + b_M_Y * M_val + unit_fe + rnorm(n * T_n, 0, 0.5)

df_did <- data.frame(
  unit    = unit,
  time    = time,
  treated = treated,
  D       = D,
  M       = M_val,
  Y       = Y_val
)

head(df_did)
```

### First-Differencing Mediator and Outcome

```{r did-med-fd}
# Compute first differences (within-unit change)
df_wide <- df_did %>%
  tidyr::pivot_wider(
    id_cols   = c(unit, treated),
    names_from  = time,
    values_from = c(M, Y, D)
  ) %>%
  dplyr::mutate(
    dM = M_1 - M_0,   # Change in mediator
    dY = Y_1 - Y_0,   # Change in outcome
    dD = D_1 - D_0    # Change in treatment (= treated for balanced panel)
  )

# Step 1: Regress dY on dD (total DiD effect)
fd_total <- lm(dY ~ dD, data = df_wide)
cat("Total DiD effect:", round(coef(fd_total)["dD"], 4), "(true: 0.2 + 0.6*0.4 = 0.44)\n")

# Step 2: Regress dM on dD (mediator DiD)
fd_med <- lm(dM ~ dD, data = df_wide)
cat("DiD effect on M:", round(coef(fd_med)["dD"], 4), "(true: 0.6)\n")

# Step 3: Regress dY on dD and dM (direct DiD effect)
fd_direct <- lm(dY ~ dD + dM, data = df_wide)
cat("Direct DiD (c'):", round(coef(fd_direct)["dD"], 4), "(true: 0.2)\n")
cat("Path b (M->Y):   ", round(coef(fd_direct)["dM"], 4), "(true: 0.4)\n")

# Indirect effect: a * b
a_did <- coef(fd_med)["dD"]
b_did <- coef(fd_direct)["dM"]
cat("Indirect effect: ", round(a_did * b_did, 4), "(true: 0.24)\n")
```

### Sequential Exogeneity of Mediator

In DiD mediation, the key identification assumption extends beyond parallel trends:

1. **Parallel trends for $Y$**: $E[\Delta Y(0) | D = 1] = E[\Delta Y(0) | D = 0]$
2. **Parallel trends for $M$**: $E[\Delta M(0) | D = 1] = E[\Delta M(0) | D = 0]$
3. **No unobserved time-varying confounders of $M \to Y$**

Assumption 3 is the mediator-version of sequential ignorability  -  it is typically not testable but can be probed by including additional controls or running pre-trend tests on the mediator.

---

## 13. Front-Door Criterion and Instrument-Based Mediation

Pearl's **front-door criterion** (1995) provides a remarkable identification result: when $T \to M \to Y$ and there is no direct $T \to Y$ path (all effects are fully mediated), we can identify the causal effect of $T$ on $Y$ even with unmeasured $T$-$Y$ confounders, provided:

1. $M$ blocks all directed paths from $T$ to $Y$
2. There are no unblocked backdoor paths from $T$ to $M$
3. All backdoor paths from $M$ to $Y$ are blocked by $T$

### Front-Door Identification

```{r frontdoor-id}
set.seed(808)
n <- 1000

# DGP satisfying front-door:
#   U -> T, U -> Y (unmeasured confounder)
#   T -> M -> Y (all effect through M)

U   <- rnorm(n)                                  # Unmeasured confounder
T_fd <- rbinom(n, 1, plogis(0.8 * U))           # T affected by U
M_fd <- 0.7 * T_fd + rnorm(n, 0, 0.8)           # M purely from T
Y_fd <- 0.5 * M_fd + 1.0 * U + rnorm(n, 0, 0.7) # Y from M and U

df_fd <- data.frame(T = T_fd, M = M_fd, Y = Y_fd)
# Note: U is not in df_fd (unmeasured)

# Front-door estimator:
# Step 1: E[M | T=1] - E[M | T=0]  (T -> M: no confounding since T is "almost" exogenous to M)
step1_fd <- lm(M ~ T, data = df_fd)
a_fd     <- coef(step1_fd)["T"]

# Step 2: Effect of M on Y, controlling for T (blocks backdoor T <- U -> Y)
step2_fd <- lm(Y ~ M + T, data = df_fd)
b_fd     <- coef(step2_fd)["M"]

# Front-door estimate = a * b
frontdoor_est <- a_fd * b_fd

# Naive OLS (biased due to U)
naive_est <- coef(lm(Y ~ T, data = df_fd))["T"]

# True effect
true_eff <- 0.7 * 0.5  # a * b = 0.35

cat("Front-door estimate:", round(frontdoor_est, 4), "(true:", true_eff, ")\n")
cat("Naive OLS estimate: ", round(naive_est, 4), "(biased upward due to U)\n")
cat("Bias in OLS:        ", round(naive_est - true_eff, 4), "\n")
```

### LATE-Based Mediation Decomposition

When treatment is assigned via an instrument $Z$ and the mediator is also instrumented, we can estimate the LATE-based mediation decomposition. This is relevant in settings like:
- Job training programs (instrument = random assignment, mediator = employment)
- Education experiments (instrument = lottery, mediator = credential attainment)

```{r late-mediation}
set.seed(909)
n <- 800

# IV setup: Z -> T -> M -> Y
Z   <- rbinom(n, 1, 0.5)                         # Instrument
T_iv <- rbinom(n, 1, plogis(-0.5 + 1.5 * Z))    # T ~ IV (compliers)
U_iv <- rnorm(n)                                  # Unobs. T-Y confounder
M_iv <- 0.6 * T_iv + rnorm(n, 0, 0.8)           # M from T
Y_iv <- 0.4 * M_iv + 0.2 * T_iv + 0.8 * U_iv + rnorm(n, 0, 0.6)

df_iv <- data.frame(Z = Z, T = T_iv, M = M_iv, Y = Y_iv)

# First-stage: T on Z
fs <- lm(T ~ Z, data = df_iv)
T_hat <- fitted(fs)

# Reduced form Y on Z (ITT for outcome)
rf_Y <- lm(Y ~ Z, data = df_iv)
# Reduced form M on Z (ITT for mediator)
rf_M <- lm(M ~ Z, data = df_iv)

# LATE via Wald: ITT_Y / (E[T|Z=1] - E[T|Z=0])
lat_total <- coef(rf_Y)["Z"] / coef(fs)["Z"]
lat_m     <- coef(rf_M)["Z"] / coef(fs)["Z"]

cat("LATE for Total Effect:", round(lat_total, 4), "(true ~0.44)\n")
cat("LATE for M:           ", round(lat_m, 4), "(true ~0.6)\n")
```

---

## 14. Lee Bounds for Mediation with Selection

When the mediator induces sample selection  -  for example, employment mediates the wage effect of training, but wages are only observed for the employed  -  we face a truncation problem. The **Lee (2009) bounds** provide sharp bounds on the treatment effect for always-employed workers ("always-takers" in the mediator).

### Setup with Selection into Mediator

```{r lee-bounds-setup}
set.seed(1001)
n <- 1000

T_lb  <- rbinom(n, 1, 0.5)
# Employment (M = 1 means employed):
pEmp  <- plogis(-0.5 + 1.2 * T_lb)
M_lb  <- rbinom(n, 1, pEmp)   # Binary mediator (employed = 1)

# Wage (Y only observed if employed):
Y_lb_full <- 2.0 + 0.5 * T_lb + rnorm(n, 0, 1)
Y_lb      <- ifelse(M_lb == 1, Y_lb_full, NA)   # Censored for unemployed

df_lb <- data.frame(T = T_lb, M = M_lb, Y = Y_lb)

cat("Employment rate (control):", round(mean(M_lb[T_lb==0]), 3), "\n")
cat("Employment rate (treated):", round(mean(M_lb[T_lb==1]), 3), "\n")
cat("Wage observed: ", sum(!is.na(Y_lb)), "/", n, "\n")
```

### Manual Lee Bounds Computation

When `MedBounds` is unavailable, we can compute Lee bounds manually:

```{r lee-manual}
# Lee (2009) bounds for the effect of T on Y among always-employed
# Step 1: Compute employment rates
p0 <- mean(M_lb[T_lb == 0])   # P(employed | T=0)
p1 <- mean(M_lb[T_lb == 1])   # P(employed | T=1)
q  <- p0 / p1                  # Trimming proportion (assuming p1 > p0)

cat("Trimming fraction q = p0/p1:", round(q, 4), "\n")

# Step 2: Trim the treated group's outcomes
# Lower bound: keep the bottom q-fraction of treated employed wages
# Upper bound: keep the top q-fraction of treated employed wages
Y_treated_emp <- sort(Y_lb[T_lb == 1 & !is.na(Y_lb)])
n_trim        <- floor(q * length(Y_treated_emp))

Y_trim_lower  <- Y_treated_emp[seq_len(length(Y_treated_emp) - n_trim)]
Y_trim_upper  <- Y_treated_emp[(n_trim + 1):length(Y_treated_emp)]

# Control group wages (observed)
Y_control_emp <- Y_lb[T_lb == 0 & !is.na(Y_lb)]

# Bounds: E[Y|T=1, trimmed] - E[Y|T=0, employed]
lb_lower <- mean(Y_trim_lower) - mean(Y_control_emp)
lb_upper <- mean(Y_trim_upper) - mean(Y_control_emp)

cat("\nLee Bounds (manual):\n")
cat("  Lower bound:", round(lb_lower, 4), "\n")
cat("  Upper bound:", round(lb_upper, 4), "\n")
cat("  True effect on wages:", 0.5, "\n")
cat("  The true effect lies within the bounds:",
    lb_lower <= 0.5 & 0.5 <= lb_upper, "\n")
```

### `causalverse::lee_bounds()` (requires MedBounds package)

The `causalverse::lee_bounds()` function provides a formal implementation with bootstrap standard errors:

```{r lee-bounds-func, eval=FALSE}
# Note: requires MedBounds package (not on CRAN)
# Installation: devtools::install_github("jonathandroth/MedBounds")

bounds_result <- lee_bounds(
  df       = df_lb,
  d        = "T",
  m        = "M",
  y        = "Y",
  numdraws = 500
)

print(bounds_result)
```

The output provides lower and upper bounds with bootstrap standard errors, allowing researchers to construct confidence intervals that account for both identification uncertainty (the bounds themselves) and estimation uncertainty.

---

## 15. Visualization Suite for Mediation

Publication-quality visualization is essential for communicating mediation results. This section provides a complete set of plots.

### 1. Path Diagram

```{r path-diagram, fig.width=8, fig.height=4}
# Manual path diagram using base R graphics
par(mar = c(1, 1, 2, 1), bg = "white")
plot(0, 0, type = "n", xlim = c(0, 10), ylim = c(0, 5),
     axes = FALSE, xlab = "", ylab = "",
     main = "Mediation Path Diagram")

# Boxes
rect(0.5, 2, 2.5, 3.5, col = "#EFF3FF", border = "#2171B5", lwd = 2)
rect(4, 2, 6, 3.5, col = "#FEE6CE", border = "#D94801", lwd = 2)
rect(7.5, 2, 9.5, 3.5, col = "#E5F5E0", border = "#31A354", lwd = 2)

# Labels
text(1.5, 2.75, "T\n(Treatment)", cex = 1.1, font = 2)
text(5.0, 2.75, "M\n(Mediator)", cex = 1.1, font = 2)
text(8.5, 2.75, "Y\n(Outcome)", cex = 1.1, font = 2)

# Arrows: T -> M (path a)
arrows(2.5, 2.75, 4.0, 2.75, lwd = 2, col = "#D94801", length = 0.15)
text(3.25, 3.1, paste0("a = ", round(a_coef, 3)), cex = 1, col = "#D94801")

# Arrows: M -> Y (path b)
arrows(6.0, 2.75, 7.5, 2.75, lwd = 2, col = "#31A354", length = 0.15)
text(6.75, 3.1, paste0("b = ", round(b_coef, 3)), cex = 1, col = "#31A354")

# Arrow: T -> Y directly (path c')
arrows(2.5, 2.2, 7.5, 2.2, lwd = 2, col = "#2171B5", length = 0.15)
text(5.0, 1.7, paste0("c' = ", round(cp_coef, 3)), cex = 1, col = "#2171B5")

# Annotation
text(5.0, 0.8,
     paste0("Indirect (a×b) = ", round(a_coef * b_coef, 3),
            "  |  Total = ", round(a_coef * b_coef + cp_coef, 3)),
     cex = 0.95, col = "gray30")
```

### 2. Effect Decomposition Bar Chart

```{r decomp-barchart, fig.width=7, fig.height=5}
# Use causalverse mediation result
decomp_df <- result$summary_df
decomp_df$effect <- factor(decomp_df$effect,
                           levels = c("ACME (Indirect)", "ADE (Direct)", "Total Effect"))

ggplot(decomp_df, aes(x = effect, y = estimate, fill = effect)) +
  geom_col(alpha = 0.85, width = 0.55) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                width = 0.18, linewidth = 0.9) +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray40") +
  geom_text(aes(label = sprintf("%.3f", estimate),
                y = estimate + 0.015 * sign(estimate)),
            vjust = -0.5, fontface = "bold", size = 4) +
  scale_fill_manual(values = c(
    "ACME (Indirect)" = "#4393C3",
    "ADE (Direct)"    = "#D73027",
    "Total Effect"    = "#2CA02C"
  )) +
  labs(
    title    = "Causal Mediation Decomposition",
    subtitle = paste0("Proportion mediated: ",
                      round(result$prop_mediated * 100, 1), "%"),
    x        = NULL,
    y        = "Effect Estimate (95% CI)",
    caption  = "Error bars = 95% quasi-Bayesian CIs; ACME via Imai et al. (2010)"
  ) +
  ama_theme() +
  theme(legend.position = "none")
```

### 3. Sensitivity Plot: ACME vs. Rho

```{r sensitivity-viz, fig.width=7, fig.height=5, eval=requireNamespace("mediation", quietly=TRUE)}
# Extract sensitivity data
rho_seq  <- seq(-0.9, 0.9, by = 0.05)
acme_rho <- sens_out$d0   # ACME estimates at each rho
ci_lo_s  <- sens_out$d0.ci[, 1]
ci_hi_s  <- sens_out$d0.ci[, 2]

sens_df <- data.frame(
  rho    = sens_out$rho,
  ACME   = sens_out$d0,
  lo     = sens_out$d0.ci[, 1],
  hi     = sens_out$d0.ci[, 2]
)

ggplot(sens_df, aes(x = rho, y = ACME)) +
  geom_ribbon(aes(ymin = lo, ymax = hi), fill = "#4393C3", alpha = 0.25) +
  geom_line(linewidth = 1.2, color = "#2171B5") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "#D73027", linewidth = 1) +
  geom_vline(xintercept = 0, linetype = "dotted", color = "gray50") +
  annotate("text", x = 0.05, y = min(sens_df$ACME, na.rm = TRUE) * 0.9,
           label = "Sequential\nIgnorability", hjust = 0, size = 3.5, color = "gray40") +
  labs(
    title    = "Sensitivity Analysis: ACME as Function of Error Correlation",
    subtitle = "Shaded = 95% CI; dashed = zero line; dotted = ρ = 0 (sequential ignorability)",
    x        = "Sensitivity Parameter ρ (Cor. of Mediator-Outcome Residuals)",
    y        = "Estimated ACME",
    caption  = "Imai et al. (2010) sensitivity; ρ = 0 = sequential ignorability holds"
  ) +
  ama_theme()
```

### 4. Proportion Mediated with Uncertainty

```{r prop-mediated-viz, fig.width=7, fig.height=4}
set.seed(42)
n_boot4  <- 1000
boot_pm  <- numeric(n_boot4)
boot_acme_v <- numeric(n_boot4)
boot_ade_v  <- numeric(n_boot4)

for (i in seq_len(n_boot4)) {
  idx <- sample(nrow(med_data), replace = TRUE)
  bd  <- med_data[idx, ]
  m2  <- lm(M ~ T + X, data = bd)
  m3  <- lm(Y ~ T + M + X, data = bd)
  a_b  <- coef(m2)["T"]
  b_b  <- coef(m3)["M"]
  cp_b <- coef(m3)["T"]
  ie_b <- a_b * b_b
  te_b <- ie_b + cp_b
  boot_acme_v[i] <- ie_b
  boot_ade_v[i]  <- cp_b
  boot_pm[i]     <- if (abs(te_b) > 1e-6) ie_b / te_b else NA
}

# Remove extreme values
boot_pm <- boot_pm[!is.na(boot_pm) & abs(boot_pm) < 5]

ci_pm <- quantile(boot_pm, c(0.025, 0.975))
pm_point <- result$prop_mediated

ggplot(data.frame(pm = boot_pm), aes(x = pm)) +
  geom_histogram(bins = 50, fill = "#4393C3", color = "white", alpha = 0.8) +
  geom_vline(xintercept = pm_point, color = "#2CA02C", linewidth = 1.3) +
  geom_vline(xintercept = ci_pm, color = "#D73027", linetype = "dashed", linewidth = 1) +
  scale_x_continuous(labels = scales::percent) +
  labs(
    title    = "Bootstrap Distribution of Proportion Mediated",
    subtitle = paste0("Point estimate: ", round(pm_point * 100, 1), "% [",
                      round(ci_pm[1] * 100, 1), "%, ",
                      round(ci_pm[2] * 100, 1), "%]"),
    x        = "Proportion of Total Effect Mediated",
    y        = "Bootstrap Frequency",
    caption  = "Green = point estimate; red dashed = 95% percentile CI"
  ) +
  ama_theme()
```

### 5. Stacked Decomposition Across Subgroups

```{r stacked-decomp, fig.width=8, fig.height=5}
# Decomposition across multiple datasets or subgroups
subgroup_results <- lapply(1:3, function(s) {
  set.seed(s * 100)
  n_sub <- 400
  T_s   <- rbinom(n_sub, 1, 0.5)
  M_s   <- (0.2 + 0.1 * s) * T_s + rnorm(n_sub, 0, 0.8)
  Y_s   <- (0.5 - 0.05 * s) * M_s + 0.15 * T_s + rnorm(n_sub, 0, 0.7)

  m_med <- lm(M_s ~ T_s)
  m_out <- lm(Y_s ~ T_s + M_s)
  a_s   <- coef(m_med)["T_s"]
  b_s   <- coef(m_out)["M_s"]
  cp_s  <- coef(m_out)["T_s"]
  ie_s  <- a_s * b_s

  data.frame(
    Subgroup  = paste0("Group ", s),
    Indirect  = ie_s,
    Direct    = cp_s,
    Total     = ie_s + cp_s
  )
})

sg_df <- do.call(rbind, subgroup_results)
sg_long <- tidyr::pivot_longer(sg_df, cols = c("Indirect", "Direct"),
                               names_to = "Component", values_to = "Estimate")

ggplot(sg_long, aes(x = Subgroup, y = Estimate, fill = Component)) +
  geom_col(position = "stack", alpha = 0.85, width = 0.6) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  geom_text(data = sg_df,
            aes(x = Subgroup, y = Total + 0.005, label = round(Total, 3),
                fill = NULL),
            inherit.aes = FALSE, fontface = "bold", size = 4) +
  scale_fill_manual(values = c("Indirect" = "#4393C3", "Direct" = "#D73027"),
                    name = "Effect Component") +
  labs(
    title    = "Mediation Decomposition Across Subgroups",
    subtitle = "Stacked bars show direct and indirect components; label = total effect",
    x        = NULL,
    y        = "Effect Estimate"
  ) +
  ama_theme()
```

---

## 16. Structural Equation Modeling (SEM) for Mediation

SEM provides a flexible framework for mediation analysis, especially with multiple mediators, latent variables, or complex causal structures.

### Simple Mediation in `lavaan`

```{r lavaan-simple, eval=requireNamespace("lavaan", quietly=TRUE)}
library(lavaan)

# lavaan syntax for simple mediation
model_simple <- "
  # Mediator model (a path)
  M ~ a*T + X

  # Outcome model (b and c' paths)
  Y ~ cp*T + b*M + X

  # Defined indirect and total effects
  indirect := a*b
  direct   := cp
  total    := cp + a*b
  prop_med := a*b / (cp + a*b)
"

fit_simple <- lavaan::sem(model_simple, data = med_data,
                          se = "bootstrap", bootstrap = 500, seed = 42)

# Parameter estimates
params <- lavaan::parameterEstimates(fit_simple, boot.ci.type = "perc")
params[params$label %in% c("a", "b", "cp", "indirect", "direct", "total", "prop_med"), ]
```

### Model Fit Assessment

```{r sem-fit, eval=requireNamespace("lavaan", quietly=TRUE)}
# Model fit indices
fit_indices <- lavaan::fitmeasures(fit_simple,
                                   c("chisq", "df", "pvalue", "cfi", "tli",
                                     "rmsea", "rmsea.ci.lower", "rmsea.ci.upper",
                                     "srmr", "aic", "bic"))
print(round(fit_indices, 4))
```

Interpretation guidelines for mediation SEM fit:
- **CFI / TLI > 0.95**: Good fit
- **RMSEA < 0.06**: Good fit (< 0.08 acceptable)
- **SRMR < 0.08**: Good fit
- **p-value for chi-square > 0.05**: Model cannot be rejected (but often overpowered with large N)

### Modification Indices

```{r sem-modindices, eval=requireNamespace("lavaan", quietly=TRUE)}
# Modification indices (misspecification diagnostics)
mi <- lavaan::modindices(fit_simple, sort. = TRUE, maximum.number = 10)
print(mi[mi$op != "~~" | mi$lhs != mi$rhs, ])
```

### Multiple Outcomes SEM

```{r sem-multi-outcome, eval=requireNamespace("lavaan", quietly=TRUE)}
set.seed(111)
n_sem <- 500

T_sem <- rbinom(n_sem, 1, 0.5)
M_sem <- 0.5 * T_sem + rnorm(n_sem, 0, 0.8)
Y1_sem <- 0.3 * M_sem + 0.2 * T_sem + rnorm(n_sem, 0, 0.7)
Y2_sem <- 0.4 * M_sem + 0.1 * T_sem + rnorm(n_sem, 0, 0.8)

df_sem2 <- data.frame(T = T_sem, M = M_sem, Y1 = Y1_sem, Y2 = Y2_sem)

model_multi <- "
  M  ~ a*T
  Y1 ~ b1*M + cp1*T
  Y2 ~ b2*M + cp2*T

  indirect_Y1 := a*b1
  indirect_Y2 := a*b2
  total_Y1    := cp1 + a*b1
  total_Y2    := cp2 + a*b2
"

fit_multi <- lavaan::sem(model_multi, data = df_sem2,
                         se = "bootstrap", bootstrap = 300, seed = 42)

params_multi <- lavaan::parameterEstimates(fit_multi, boot.ci.type = "perc")
params_multi[params_multi$label %in%
               c("indirect_Y1", "indirect_Y2", "total_Y1", "total_Y2"), ]
```

### Bayesian SEM with `blavaan`

```{r bsem-demo, eval=requireNamespace("blavaan", quietly=TRUE)}
library(blavaan)

model_bayes <- "
  M ~ a*T
  Y ~ b*M + cp*T
  indirect := a*b
  total    := cp + a*b
"

# Bayesian SEM (uses Stan or JAGS under the hood)
bfit <- blavaan::bsem(
  model_bayes,
  data    = med_data[1:200, ],   # Subset for speed
  n.chains = 2,
  burnin   = 500,
  sample   = 1000,
  seed     = 42,
  bcontrol = list(cores = 1)
)

blavaan::summary(bfit, neff = TRUE, psrf = TRUE)
```

---

## 17. Complete Publication Pipeline

This section walks through a complete mediation analysis from data to publication-ready output.

### Step 1: Theory and Hypotheses

Before analyzing data, state your hypotheses explicitly:

- **H1**: $T$ affects $Y$ (total effect, $\tau \neq 0$)
- **H2**: $T$ affects $M$ (path $a \neq 0$)
- **H3**: $M$ affects $Y$ controlling for $T$ (path $b \neq 0$)
- **H4**: The indirect effect $ab$ is different from zero (ACME $\neq 0$)
- **H5 (optional)**: Mediation is partial, not complete ($c' \neq 0$)

### Step 2: Data and DGP

```{r pub-data}
set.seed(2025)
n_pub <- 500

# Simulate a realistic mediation scenario:
# Training program (T) -> Knowledge acquisition (M) -> Job performance (Y)
# Covariates: prior experience (X1), education level (X2)

X1  <- rnorm(n_pub, 5, 2)    # Prior experience (years)
X2  <- round(rnorm(n_pub, 14, 2))  # Education level
T_pub <- rbinom(n_pub, 1, plogis(-1 + 0.1 * X1 + 0.05 * X2))  # Training assignment

# Knowledge: T -> M (a = 0.45), also from X1 and X2
M_pub <- 0.45 * T_pub + 0.10 * X1 + 0.08 * X2 + rnorm(n_pub, 0, 0.9)

# Performance: M -> Y (b = 0.35), T -> Y (c' = 0.15), also from X1
Y_pub <- 0.35 * M_pub + 0.15 * T_pub + 0.12 * X1 + rnorm(n_pub, 0, 0.8)

df_pub <- data.frame(T = T_pub, M = M_pub, Y = Y_pub, X1 = X1, X2 = X2)

cat("Sample size:", nrow(df_pub), "\n")
cat("Treatment prevalence:", round(mean(T_pub), 3), "\n")
cat("Outcome mean:", round(mean(Y_pub), 3), "SD:", round(sd(Y_pub), 3), "\n")
```

### Step 3: Descriptive Statistics

```{r pub-desc}
desc_df <- df_pub %>%
  tidyr::pivot_longer(cols = everything(), names_to = "Variable", values_to = "Value") %>%
  dplyr::group_by(Variable) %>%
  dplyr::summarise(
    N    = dplyr::n(),
    Mean = round(mean(Value, na.rm = TRUE), 3),
    SD   = round(sd(Value, na.rm = TRUE), 3),
    Min  = round(min(Value, na.rm = TRUE), 3),
    Max  = round(max(Value, na.rm = TRUE), 3),
    .groups = "drop"
  )

knitr::kable(desc_df, caption = "Table 1. Descriptive Statistics")
```

### Step 4: Correlation Matrix

```{r pub-corr}
corr_mat <- round(cor(df_pub), 3)
knitr::kable(corr_mat, caption = "Table 2. Correlation Matrix")
```

### Step 5: Baron-Kenny Regressions

```{r pub-bk}
bk1 <- lm(Y ~ T + X1 + X2, data = df_pub)
bk2 <- lm(M ~ T + X1 + X2, data = df_pub)
bk3 <- lm(Y ~ T + M + X1 + X2, data = df_pub)

# Summary table
bk_table <- data.frame(
  Parameter  = c("T -> Y (total)", "T -> M (a)", "M -> Y (b)", "T -> Y (c', direct)"),
  Estimate   = c(
    coef(bk1)["T"], coef(bk2)["T"],
    coef(bk3)["M"], coef(bk3)["T"]
  ),
  SE = c(
    summary(bk1)$coef["T", 2], summary(bk2)$coef["T", 2],
    summary(bk3)$coef["M", 2], summary(bk3)$coef["T", 2]
  ),
  p_value = c(
    summary(bk1)$coef["T", 4], summary(bk2)$coef["T", 4],
    summary(bk3)$coef["M", 4], summary(bk3)$coef["T", 4]
  )
)

bk_table$Estimate <- round(bk_table$Estimate, 4)
bk_table$SE       <- round(bk_table$SE, 4)
bk_table$p_value  <- round(bk_table$p_value, 4)

knitr::kable(bk_table, caption = "Table 3. Baron-Kenny Path Estimates")
```

### Step 6: Causal Mediation (Imai et al.)

```{r pub-causal-med}
set.seed(42)
pub_result <- mediation_analysis(
  data       = df_pub,
  outcome    = "Y",
  treatment  = "T",
  mediator   = "M",
  covariates = c("X1", "X2"),
  sims       = 1000,
  seed       = 42
)

# Publication-ready table
med_table <- pub_result$summary_df
med_table$estimate  <- round(med_table$estimate, 4)
med_table$ci_lower  <- round(med_table$ci_lower, 4)
med_table$ci_upper  <- round(med_table$ci_upper, 4)
med_table$CI_95 <- paste0("[", med_table$ci_lower, ", ", med_table$ci_upper, "]")

knitr::kable(
  med_table[, c("effect", "estimate", "CI_95")],
  col.names = c("Effect", "Estimate", "95% CI"),
  caption   = "Table 4. Causal Mediation Analysis Results (Imai et al. 2010)"
)

cat("\nProportion mediated:", round(pub_result$prop_mediated * 100, 1), "%\n")
```

### Step 7: Sensitivity Analysis

```{r pub-sensitivity, eval=requireNamespace("mediation", quietly=TRUE)}
med_m_pub  <- lm(M ~ T + X1 + X2, data = df_pub)
med_y_pub  <- lm(Y ~ T + M + X1 + X2, data = df_pub)

set.seed(42)
med_pub_obj <- mediation::mediate(
  model.m  = med_m_pub,
  model.y  = med_y_pub,
  treat    = "T",
  mediator = "M",
  sims     = 500
)

sens_pub <- mediation::medsens(med_pub_obj, rho.by = 0.05, sims = 100)
s_pub    <- summary(sens_pub)

cat("Critical rho (ACME = 0):", round(s_pub$rho[1], 3), "\n")
cat("R² (M residuals):", round(s_pub$r.square.m[1], 3), "\n")
cat("R² (Y residuals):", round(s_pub$r.square.y[1], 3), "\n")
```

### Step 8: Publication Figure

```{r pub-figure, fig.width=9, fig.height=6}
pub_plot_df <- pub_result$summary_df
pub_plot_df$effect <- factor(pub_plot_df$effect,
                             levels = c("ACME (Indirect)", "ADE (Direct)", "Total Effect"))

ggplot(pub_plot_df, aes(x = effect, y = estimate, fill = effect)) +
  geom_col(alpha = 0.88, width = 0.5) +
  geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper),
                width = 0.15, linewidth = 0.9, color = "gray20") +
  geom_hline(yintercept = 0, linetype = "dashed", color = "gray50", linewidth = 0.7) +
  geom_text(aes(label = sprintf("β = %.3f", estimate),
                y = estimate + 0.008 * sign(estimate)),
            vjust = -0.3, size = 3.8, fontface = "bold") +
  scale_fill_manual(values = c(
    "ACME (Indirect)" = "#4393C3",
    "ADE (Direct)"    = "#D73027",
    "Total Effect"    = "#2CA02C"
  )) +
  scale_y_continuous(expand = expansion(mult = c(0.1, 0.2))) +
  labs(
    title    = "Figure 1. Causal Mediation Analysis: Training → Knowledge → Performance",
    subtitle = paste0("N = ", nrow(df_pub),
                      "; proportion mediated = ",
                      round(pub_result$prop_mediated * 100, 1), "%"),
    x        = NULL,
    y        = "Unstandardized Effect (95% CI)",
    caption  = paste0("Note: Effects estimated via Imai et al. (2010) using 1000 quasi-Bayesian simulations.\n",
                      "ACME = Average Causal Mediation Effect; ADE = Average Direct Effect.\n",
                      "Sequential ignorability assumed. See sensitivity analysis for robustness.")
  ) +
  ama_theme() +
  theme(legend.position = "none")
```

### Step 9: APA/AMA-Style Reporting Text

The following text demonstrates how to report mediation results following APA and top journal conventions:

---

*We conducted a causal mediation analysis following Imai et al. (2010) to examine whether knowledge acquisition (M) mediates the effect of the training program (T) on job performance (Y), controlling for prior experience (X1) and education level (X2). Results revealed a statistically significant total effect of training on performance (β = `r round(pub_result$total, 3)`, 95% CI [`r round(pub_result$summary_df$ci_lower[3], 3)`, `r round(pub_result$summary_df$ci_upper[3], 3)`]). The average causal mediation effect (ACME; indirect effect through knowledge acquisition) was β = `r round(pub_result$acme, 3)` (95% CI [`r round(pub_result$summary_df$ci_lower[1], 3)`, `r round(pub_result$summary_df$ci_upper[1], 3)`]), accounting for `r round(pub_result$prop_mediated * 100, 1)`% of the total effect. The average direct effect (ADE) was β = `r round(pub_result$ade, 3)` (95% CI [`r round(pub_result$summary_df$ci_lower[2], 3)`, `r round(pub_result$summary_df$ci_upper[2], 3)`]), indicating partial mediation. Sensitivity analysis confirmed that the ACME was robust to moderate violations of sequential ignorability.*

---

### Step 10: Reporting Checklist

The following checklist reflects best practices from VanderWeele (2016) and the APA mediation reporting standards:

| Item | Status |
|---|---|
| State hypothesized causal model | ✓ Stated in theory section |
| Report total effect | ✓ Reported with CI |
| Report ACME (indirect effect) | ✓ Reported with CI |
| Report ADE (direct effect) | ✓ Reported with CI |
| Report proportion mediated | ✓ Reported with CI |
| Specify identification assumptions | ✓ Sequential ignorability stated |
| Conduct sensitivity analysis | ✓ Imai-Keele medsens reported |
| Report simulation method and number | ✓ Quasi-Bayesian, 1000 sims |
| Address alternative explanations | Recommended in manuscript |
| Pre-register if possible | Recommended |

---

## 18. Summary and Reference Table

### Effect Estimand Reference

| Estimand | Symbol | When to Use |
|---|---|---|
| ACME (T=0) | $\delta(0)$ | Indirect effect in control population |
| ACME (T=1) | $\delta(1)$ | Indirect effect in treated population |
| Average ACME | $\bar{\delta}$ | Default in `mediation::mediate()` |
| ADE (T=0) | $\zeta(0)$ | Direct effect with M at control level |
| ADE (T=1) | $\zeta(1)$ | Direct effect with M at treated level |
| Total Effect | $\tau$ | Overall treatment effect |
| Proportion Mediated | PM | Fraction of TE through M |
| CDE | CDE($m_0$) | Direct effect with M set to $m_0$ |
| NDE / PNDE | NDE | Direct effect (with M at natural control level) |
| NIE / TNIE | NIE | Indirect effect (ACME equivalent) |

### Identification Assumptions Summary

| Method | Key Assumption | Test Available? |
|---|---|---|
| Baron-Kenny | Sequential ignorability (implicit) | No |
| Imai et al. (2010) | Sequential ignorability (explicit) | Via sensitivity (rho) |
| Natural effects | Cross-world independence | No |
| Front-door | No direct T → Y; no T → M confounding | Partially |
| SEM | Sequential ignorability + model form | Model fit tests |
| DiD mediation | Parallel trends for both M and Y | Pre-trend tests |
| Lee bounds | Monotonicity of selection | No (identifying bound) |

### Package Reference

| Package | Function | Purpose |
|---|---|---|
| `causalverse` | `mediation_analysis()` | Wrapper for Imai et al. mediation |
| `causalverse` | `med_ind()` | Monte Carlo indirect effects |
| `causalverse` | `lee_bounds()` | Lee bounds for selected mediators |
| `mediation` | `mediate()` | Core causal mediation |
| `mediation` | `medsens()` | Sensitivity analysis |
| `lavaan` | `sem()` | SEM-based mediation |
| `blavaan` | `bsem()` | Bayesian SEM mediation |
| `grf` | `causal_forest()` | Heterogeneous mediation |
| `CMAverse` | `cmest()` | VanderWeele decomposition |

---

## 19. Common Pitfalls and Practical Guidance

### Pitfall 1: Treating Baron-Kenny as Causal Without Justification

The Baron-Kenny "four steps" do not establish causality. They establish statistical associations. The causal interpretation requires:
1. A well-articulated theoretical mechanism
2. Sequential ignorability (or a strong substitute)
3. No treatment-mediator interaction that is ignored

**Guidance**: Always use the Imai et al. (2010) framework and explicitly acknowledge sequential ignorability as an assumption.

### Pitfall 2: Ignoring Measurement Error in the Mediator

When $M$ is measured with error ($M^* = M + \nu$), both the estimated path $b$ and ACME will be attenuated. This is a form of "regression dilution."

```{r measurement-error}
set.seed(1234)
n_me <- 500

T_me  <- rbinom(n_me, 1, 0.5)
M_me  <- 0.5 * T_me + rnorm(n_me, 0, 0.8)    # True M
M_obs <- M_me + rnorm(n_me, 0, 0.6)           # Observed M with error (noise SD = 0.6)
Y_me  <- 0.4 * M_me + 0.2 * T_me + rnorm(n_me, 0, 0.7)

# With true M
m2_true <- lm(M_me ~ T_me)
m3_true <- lm(Y_me ~ T_me + M_me)
ie_true  <- coef(m2_true)["T_me"] * coef(m3_true)["M_me"]

# With error-prone M
m2_obs <- lm(M_obs ~ T_me)
m3_obs <- lm(Y_me ~ T_me + M_obs)
ie_obs  <- coef(m2_obs)["T_me"] * coef(m3_obs)["M_obs"]

cat("Indirect effect with true M:           ", round(ie_true, 4), "(true: 0.20)\n")
cat("Indirect effect with mismeasured M:    ", round(ie_obs, 4), "(attenuated)\n")
cat("Attenuation factor (obs/true):         ", round(ie_obs/ie_true, 3), "\n")
cat("\nImplication: Measurement error in M biases ACME toward zero (attenuation bias).\n")
```

### Pitfall 3: Multicollinearity Between T and M in the Outcome Model

When $T$ and $M$ are highly correlated, estimating their separate effects in the outcome model becomes unstable. This inflates standard errors and can produce counterintuitive signs.

```{r multicollinearity-check}
set.seed(5678)
n_mc <- 400

T_col <- rbinom(n_mc, 1, 0.5)
M_col <- 0.9 * T_col + rnorm(n_mc, 0, 0.3)   # High T-M correlation
Y_col <- 0.3 * M_col + 0.1 * T_col + rnorm(n_mc, 0, 0.7)

cat("Correlation between T and M:", round(cor(T_col, M_col), 3), "\n")

m3_mc <- lm(Y_col ~ T_col + M_col)
cat("\nOutcome model with high T-M collinearity:\n")
print(summary(m3_mc)$coefficients)

# Variance Inflation Factor (VIF)
if (requireNamespace("car", quietly = TRUE)) {
  cat("\nVIF values:\n")
  print(car::vif(m3_mc))
} else {
  # Manual VIF
  r2_T <- summary(lm(T_col ~ M_col))$r.squared
  vif_T <- 1 / (1 - r2_T)
  cat("\nVIF for T:", round(vif_T, 2), "\n")
  cat("(VIF > 10 suggests problematic multicollinearity)\n")
}
```

### Pitfall 4: Improper Handling of Confounders

Including post-treatment confounders (variables that are causally between $T$ and $M$, or between $T$ and $Y$) in the models can introduce collider bias.

**Recommended covariate inclusion strategy:**

| Variable Type | Include in M model? | Include in Y model? |
|---|---|---|
| Pre-treatment baseline $X$ | Yes | Yes |
| Post-treatment variable $L$ (T → L → M) | No | No |
| $L$ as M-Y confounder (T → L, L → Y) | No | Yes (carefully) |
| Collider $C$ (T → C ← Y) | No | No |

### Pitfall 5: Reverse Causation (M → T)

If the mediator may also influence treatment (e.g., in longitudinal designs), then the standard mediation model is mis-specified. In such cases:

- Use cross-lagged panel models or bivariate LGCMs (if longitudinal data)
- Consider bidirectional SEM
- Apply g-estimation or marginal structural models (for time-varying treatments)

---

## 20. Power Analysis for Mediation

Mediation studies are frequently underpowered. Detecting an indirect effect of size $ab$ requires detecting two separate effects ($a$ and $b$), so power is typically lower than for a single-path analysis.

### Analytical Power for Sobel Test

```{r power-analysis}
# Power for Sobel test under various scenarios
# Following Fritz & MacKinnon (2007): power depends on a, b, and n

power_sobel <- function(n, a, b, se_a, se_b, alpha = 0.05) {
  # Sobel SE
  se_ab <- sqrt(b^2 * se_a^2 + a^2 * se_b^2)
  # Noncentrality parameter
  lambda <- abs(a * b) / se_ab
  # Approximate SE based on path lengths
  z_crit <- qnorm(1 - alpha/2)
  # Power
  pnorm(lambda - z_crit) + pnorm(-lambda - z_crit)
}

# Simulate power across sample sizes
power_grid <- expand.grid(
  n       = c(50, 100, 200, 400, 800),
  a       = c(0.2, 0.3, 0.5),
  b       = c(0.2, 0.3)
)

power_grid$power <- mapply(function(n, a, b) {
  # Approximate SEs for standardized paths
  se_a_approx <- sqrt((1 - a^2) / (n - 2))
  se_b_approx <- sqrt((1 - b^2 - a^2) / (n - 3))
  power_sobel(n, a, b, se_a_approx, se_b_approx)
}, power_grid$n, power_grid$a, power_grid$b)

power_grid$label <- paste0("a=", power_grid$a, ", b=", power_grid$b)

# Print summary
power_wide <- power_grid %>%
  dplyr::mutate(power = round(power, 3)) %>%
  dplyr::select(n, label, power) %>%
  tidyr::pivot_wider(names_from = label, values_from = power)

knitr::kable(power_wide,
             caption = "Table: Power for Sobel Test (alpha = 0.05, two-tailed)")
```

```{r power-plot, fig.width=8, fig.height=5}
ggplot(power_grid, aes(x = n, y = power, color = label, group = label)) +
  geom_line(linewidth = 1.1) +
  geom_point(size = 2.5) +
  geom_hline(yintercept = 0.80, linetype = "dashed", color = "gray40") +
  geom_hline(yintercept = 0.90, linetype = "dotted", color = "gray60") +
  annotate("text", x = 820, y = 0.81, label = "80% power", hjust = 1, size = 3.5) +
  annotate("text", x = 820, y = 0.91, label = "90% power", hjust = 1, size = 3.5) +
  scale_x_continuous(breaks = c(50, 100, 200, 400, 800)) +
  scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
  labs(
    title    = "Power Analysis for Mediation (Sobel Test)",
    subtitle = "Power to detect indirect effect ab at alpha = 0.05",
    x        = "Sample Size",
    y        = "Statistical Power",
    color    = "Effect sizes (a, b)",
    caption  = "Based on analytical Sobel power formula (Fritz & MacKinnon 2007)"
  ) +
  ama_theme()
```

### Monte Carlo Power for Bootstrap Tests

Bootstrap-based power can be estimated via simulation, which is more accurate for small samples:

```{r mc-power}
set.seed(2024)

# Power simulation: n = 200, a = 0.3, b = 0.3 (true ACME = 0.09)
n_sims  <- 300   # In practice use 1000+
n_pw    <- 200
a_true  <- 0.3
b_true  <- 0.3
n_boots <- 200   # Per simulation; use 1000+ in practice

power_sim <- sapply(seq_len(n_sims), function(s) {
  set.seed(s)
  T_pw <- rbinom(n_pw, 1, 0.5)
  M_pw <- a_true * T_pw + rnorm(n_pw, 0, sqrt(1 - a_true^2))
  Y_pw <- b_true * M_pw + rnorm(n_pw, 0, sqrt(1 - b_true^2))

  # Bootstrap indirect effect
  boot_ie <- replicate(n_boots, {
    idx <- sample(n_pw, replace = TRUE)
    m2  <- lm(M_pw[idx] ~ T_pw[idx])
    m3  <- lm(Y_pw[idx] ~ T_pw[idx] + M_pw[idx])
    coef(m2)[2] * coef(m3)[3]
  })

  ci <- quantile(boot_ie, c(0.025, 0.975))
  # Significant if CI excludes zero
  as.numeric(ci[1] > 0 | ci[2] < 0)
})

cat("Simulated power (n=200, a=0.3, b=0.3, 300 reps):\n")
cat("  Bootstrap power:", round(mean(power_sim), 3), "\n")
cat("  (True ACME = 0.09)\n")
```

### Sample Size Recommendations (Fritz & MacKinnon 2007)

| Effect magnitude | a (path T→M) | b (path M→Y) | N for 80% power |
|---|---|---|---|
| Small-small | 0.14 | 0.14 | ~21,000 |
| Small-medium | 0.14 | 0.39 | ~3,200 |
| Medium-medium | 0.39 | 0.39 | ~558 |
| Medium-large | 0.39 | 0.59 | ~116 |
| Large-large | 0.59 | 0.59 | ~71 |

*Note: These are for Sobel test at alpha = 0.05. Bootstrap tests require somewhat smaller samples for the same power.*

---

## 21. Mediation in Experimental vs. Observational Settings

### Randomized Experiment with Mediation

When $T$ is randomly assigned:
- SI.1 is automatically satisfied (no $T$-$Y$ confounding)
- SI.2 still needs to be assumed (no $M$-$Y$ confounding)
- The ACME is identified but not guaranteed by randomization alone

This is the "fundamental problem of mediation" (Imai et al. 2011): even a perfect experiment cannot fully identify causal mechanisms without additional assumptions.

### Designs That Strengthen Causal Claims

**Design 1: Concurrent manipulation of M**

If we can randomize both $T$ and $M$ independently (e.g., in a $2 \times 2$ factorial design), we can directly estimate:
- Main effect of $T$ (= ADE when M is at the randomized level)
- Effect of $M$ (= path $b$)
- Interaction $T \times M$

```{r factorial-design}
set.seed(1357)
n_fac <- 400

# 2x2 factorial: T and M both randomized
T_fac <- rbinom(n_fac, 1, 0.5)
M_fac <- rbinom(n_fac, 1, 0.5)   # Independently randomized M

# True effects: T main = 0.15, M main = 0.40, T*M interaction = 0.10
Y_fac <- 0.15 * T_fac + 0.40 * M_fac + 0.10 * T_fac * M_fac + rnorm(n_fac, 0, 0.8)

# Analysis
m_fac <- lm(Y_fac ~ T_fac * M_fac)
print(summary(m_fac)$coefficients)

cat("\nIn factorial design:\n")
cat("  ADE (when M = 0):", round(coef(m_fac)["T_fac"], 4), "(true: 0.15)\n")
cat("  ADE (when M = 1):", round(coef(m_fac)["T_fac"] + coef(m_fac)["T_fac:M_fac"], 4),
    "(true: 0.25)\n")
cat("  Effect of M:     ", round(coef(m_fac)["M_fac"], 4), "(true: 0.40)\n")
```

**Design 2: Encouragement design (LATE mediation)**

Randomize an encouragement $Z$ that affects $M$ but has no direct effect on $Y$ (exclusion restriction). This gives an IV for $M$.

**Design 3: Sequential assignment**

Randomize $T$ first, then re-randomize $M$ among the treated. This directly identifies $\delta(1)$ (the ACME at $T = 1$) without SI.2.

### Observational Mediation: Practical Recommendations

In observational settings, the following strategies strengthen causal claims:

1. **Control richly**: Include all plausible $M$-$Y$ confounders in both mediator and outcome models.
2. **Temporal ordering**: Ensure $T$ precedes $M$ precedes $Y$ temporally.
3. **Pre-treatment M**: If possible, include pre-treatment values of $M$ as covariates.
4. **Placebo mediator**: Test a placebo mediator (one that theoretically should not mediate) as a falsification test.
5. **Multiple sensitivity analyses**: Report results under a range of $\rho$ values, not just the significance-flipping threshold.

```{r obs-mediation-robust}
set.seed(7890)
n_obs <- 600

X_obs <- rnorm(n_obs)             # Pre-treatment covariate
T_obs <- rbinom(n_obs, 1, plogis(0.4 * X_obs))  # Observational T

# Unmeasured confounder U (affects M and Y)
U_obs <- rnorm(n_obs, 0, 0.5)

M_obs2 <- 0.5 * T_obs + 0.3 * X_obs + U_obs + rnorm(n_obs, 0, 0.7)
Y_obs2 <- 0.3 * M_obs2 + 0.2 * T_obs + 0.4 * X_obs + 0.6 * U_obs + rnorm(n_obs, 0, 0.6)

# Model A: Naive (ignoring U)
m_naiveM <- lm(M_obs2 ~ T_obs + X_obs)
m_naiveY <- lm(Y_obs2 ~ T_obs + M_obs2 + X_obs)

ie_naive <- coef(m_naiveM)["T_obs"] * coef(m_naiveY)["M_obs2"]
cat("Naive ACME (omitting U): ", round(ie_naive, 4), "\n")
cat("True ACME: 0.5 * 0.3 = ", 0.5 * 0.3, "\n")
cat("Bias (due to omitted U): ", round(ie_naive - 0.15, 4), "\n")
cat("\nThis illustrates that even controlling for X, omitting U biases ACME.\n")
cat("Sensitivity analysis is essential in observational mediation.\n")
```

---

## References

Baron, R. M., & Kenny, D. A. (1986). The moderator-mediator variable distinction in social psychological research: Conceptual, strategic, and statistical considerations. *Journal of Personality and Social Psychology*, 51(6), 1173–1182.

Hayes, A. F. (2013). *Introduction to mediation, moderation, and conditional process analysis*. Guilford Press.

Imai, K., Keele, L., & Tingley, D. (2010). A general approach to causal mediation analysis. *Psychological Methods*, 15(4), 309–334.

Imai, K., Keele, L., Tingley, D., & Yamamoto, T. (2011). Unpacking the black box of causality: Learning about causal mechanisms from experimental and observational studies. *American Political Science Review*, 105(4), 765–789.

Lee, D. S. (2009). Training, wages, and sample selection: Estimating sharp bounds on treatment effects. *Review of Economic Studies*, 76(3), 1071–1102.

Pearl, J. (1995). Causal diagrams for empirical research. *Biometrika*, 82(4), 669–688.

Pearl, J. (2001). Direct and indirect effects. *Proceedings of the Seventeenth Conference on Uncertainty in Artificial Intelligence*, 411–420.

Robins, J. M., & Greenland, S. (1992). Identifiability and exchangeability for direct and indirect effects. *Epidemiology*, 3(2), 143–155.

Selig, J. P., & Preacher, K. J. (2008). Monte Carlo method for assessing mediation: An interactive tool for creating confidence intervals for indirect effects. Available from http://quantpsy.org/.

Sobel, M. E. (1982). Asymptotic confidence intervals for indirect effects in structural equation models. *Sociological Methodology*, 13, 290–312.

VanderWeele, T. J. (2014). A unification of mediation and interaction: A four-way decomposition. *Epidemiology*, 25(5), 749–761.

VanderWeele, T. J. (2016). Mediation analysis: A practitioner's guide. *Annual Review of Public Health*, 37, 17–32.

VanderWeele, T. J., & Ding, P. (2017). Sensitivity analysis in observational research: Introducing the E-value. *Annals of Internal Medicine*, 167(4), 268–274.

Wager, S., & Athey, S. (2018). Estimation and inference of heterogeneous treatment effects using random forests. *Journal of the American Statistical Association*, 113(523), 1228–1242.

Fritz, M. S., & MacKinnon, D. P. (2007). Required sample size to detect the mediated effect. *Psychological Science*, 18(3), 233–239.

MacKinnon, D. P., Lockwood, C. M., Hoffman, J. M., West, S. G., & Sheets, V. (2002). A comparison of methods to test mediation and other intervening variable effects. *Psychological Methods*, 7(1), 83–104.

Preacher, K. J., & Hayes, A. F. (2008). Asymptotic and resampling strategies for assessing and comparing indirect effects in multiple mediator models. *Behavior Research Methods*, 40(3), 879–891.

Tingley, D., Yamamoto, T., Hirose, K., Keele, L., & Imai, K. (2014). mediation: R package for causal mediation analysis. *Journal of Statistical Software*, 59(5), 1–38.

Rubin, D. B. (2004). Direct and indirect causal effects via potential outcomes. *Scandinavian Journal of Statistics*, 31(2), 161–170.

---

## Session Information

```{r session-info}
sessionInfo()
```
