---
title: "3. Synthetic Difference-in-Differences"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{synthdid}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(causalverse)
```

# Synthetic DID

## Block Adoption (Assignment) 

This section contains the code from the `synthdid` package, which can handle block adoption (i.e., when all units are treated at the same time).


```{r}
library(synthdid)

# Estimate the effect of California Proposition 99 on cigarette consumption
data('california_prop99')

setup = panel.matrices(california_prop99)

tau.hat = synthdid_estimate(setup$Y, setup$N0, setup$T0)

print(summary(tau.hat))

# for only one treated unit, we can only use the placebo method
# se = sqrt(vcov(tau.hat, method='placebo'))
# sqrt(vcov(tau.hat, method='bootstrap'))
# sqrt(vcov(tau.hat, method='jackknife'))

# sprintf('point estimate: %1.2f', tau.hat)
# sprintf('95%% CI (%1.2f, %1.2f)', tau.hat - 1.96 * se, tau.hat + 1.96 * se)
plot(tau.hat) + ama_theme()
```

Using `fixest` dataset

```{r}
library(fixest)
library(tidyverse)
```

If you want to use the `base_stagg` dataset, you have to choose just 1 period, because function `synthdid_estimate` cannot handle staggered adoption. 

```{r, eval = FALSE}
df <- base_stagg |>
  
  filter(year_treated %in% c(10000, 8)) |>
  transmute(
    id = as.factor(id),
    year = as.integer(year),
    y = as.double(y),
    treated = as.integer(if_else(time_to_treatment >= 0, 1, 0))
  )
head(df)

setup = df |> 
  # specify column names
  panel.matrices(unit = "id", time = "year", outcome = "y", treatment = "treated")
```



```{r}
head(base_did)
setup = base_did |> 
  mutate(
    id = as.factor(id),
    period = as.integer(period),
    y = as.double(y),
    post = as.integer(post)
  ) |> 
  # correct treatment 
  mutate(treatment = as.integer(if_else(treat == 0, 0, post))) |> 
  
  panel.matrices(unit = "id", time = "period", outcome = "y", treatment = "treatment")


tau.hat = synthdid_estimate(setup$Y, setup$N0, setup$T0)
# sqrt(vcov(tau.hat, method='bootstrap'))
sqrt(vcov(tau.hat, method='jackknife'))

# to use this SE method, must have more contrls than treated units
# can try the base_stagg dataset from above 
# sqrt(vcov(tau.hat, method='placebo'))

# Comparing counterfactual estimate to real-world observation
synthdid::synthdid_plot(tau.hat
                        # , se.method = "placebo"
                        ) + ama_theme()

# Control unit contribution plot (black is the ATE, grey lines are CIs)
synthdid::synthdid_units_plot(tau.hat) + ama_theme(base_size = 8)

# pre-treatment parallel trends
# synthdid::synthdid_plot(tau.hat, overlay = 1)
# synthdid::synthdid_plot(tau.hat, overlay = .8)
# synthdid::synthdid_placebo_plot(tau.hat)
# synthdid::synthdid_rmse_plot(tau.hat)
```


```{r}
tau.sc   = sc_estimate(setup$Y, setup$N0, setup$T0)
tau.did  = did_estimate(setup$Y, setup$N0, setup$T0)
estimates = list(tau.did, tau.sc, tau.hat)
names(estimates) = c('Diff-in-Diff', 'Synthetic Control', 'Synthetic Diff-in-Diff')

print(unlist(estimates))

synthdid::synthdid_plot(estimates) + 
  ama_theme()
```

Spaghetti plots

```{r}
top.controls = synthdid_controls(tau.hat)[1:10, , drop=FALSE]
plot(tau.hat, spaghetti.units=rownames(top.controls)) + ama_theme()
```

```{r}
synthdid::synthdid_units_plot(estimates) +
  ama_theme()
```

```{r}
synthdid_units_plot(tau.hat, units = rownames(top.controls)) + ama_theme()
```

## Staggered Adoptions

In this implementation of the 

```{r}
library(tidyverse)
df <- fixest::base_stagg |>
   dplyr::mutate(treatvar = if_else(time_to_treatment >= 0, 1, 0)) |>
   dplyr::mutate(treatvar = as.integer(if_else(year_treated > (5 + 2), 0, treatvar)))


est <- causalverse::synthdid_est_ate(
  data               = df,
  adoption_cohorts   = 5:7,
  lags               = 2,
  leads              = 2,
  time_var           = "year",
  unit_id_var        = "id",
  treated_period_var = "year_treated",
  treat_stat_var     = "treatvar",
  outcome_var        = "y"
)

causalverse::synthdid_plot_ate(est)
```

